# Block 5: Tidy evaluation and data masking {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


`fit_model()` always fits `body_mass_g ~ flipper_length_mm`. To fit body mass on
bill length, or bill length on bill depth, we would have to write more functions.
The fix is to make the response and predictor arguments. The complication is that
dplyr verbs do not look up column names the way ordinary R does. This block
teaches the mental model and the four patterns you need.

## The problem

A first, reasonable attempt that does not work:

```{r}
#| eval: false
summarise_by_species <- function(data, var) {
  data |>
    dplyr::summarise(mean = mean(var, na.rm = TRUE), .by = species)
}

summarise_by_species(penguins, body_mass_g)
# Error in mean(var, na.rm = TRUE) : object 'var' not found
```

The function received `var = body_mass_g`. Inside the function, `var` is the name
of the argument. Yet the error says it cannot be found. Why?

## The mental model

Ordinary R looks up names in the calling environment: `mean(x)` searches the
function's environment, then the enclosing one, and so on.

Tidyverse verbs do something different. Inside `summarise()`, `filter()`,
`mutate()` and `arrange()`, names are looked up *inside the data frame first*. So
`summarise(mean = mean(body_mass_g))` finds `body_mass_g` as a column. This is
**data-masking**: the data frame masks the calling environment.

When you wrap a verb in your own function, the argument `var` lives in the
function's environment. dplyr looks for a column called `var`, does not find one,
then finds `var` in the function environment holding the symbol `body_mass_g`, and
does not know what to do with a bare symbol.

The fix is to tell dplyr "the user handed me a column name; substitute it into the
expression before evaluating". That is what the **embracing operator** `{{ }}`
does: it captures the unevaluated expression the user passed and re-injects it
where written.

:::{.callout-note}

In rlang's terms this is a "data-masked context", and `{{ }}` is sometimes called
"curly-curly" for its shape. The canonical reference is the rlang programming
vignette (`vignette("programming", package = "dplyr")`).

:::

## Embracing in practice

The fix:

```{r}
#| eval: false
summarise_by_species <- function(data, var) {
  data |>
    dplyr::summarise(mean = mean({{ var }}, na.rm = TRUE), .by = species)
}

summarise_by_species(penguins, body_mass_g)
# # A tibble: 3 x 2
#   species    mean
#   <fct>     <dbl>
# 1 Adelie    3701.
# 2 Gentoo    5076.
# 3 Chinstrap 3733.
```

Embrace the column every time it appears, and embrace the grouping column too if
it is an argument:

```{r}
#| eval: false
summary_grouped <- function(data, var, group) {
  data |>
    dplyr::summarise(
      mean = mean({{ var }}, na.rm = TRUE),
      sd   = sd({{ var }}, na.rm = TRUE),
      .by  = {{ group }}
    )
}

summary_grouped(penguins, body_mass_g, species)
summary_grouped(penguins, body_mass_g, island)
```

:::{.callout-note}

**Under the hood (optional).** `{{ var }}` is shorthand for `!!enquo(var)`.
`rlang::enquo()` captures the argument as a *quosure* (the expression plus the
environment it came from), and `!!` injects it back into the verb. You almost
never need to write `enquo()`/`!!` by hand, but seeing the expansion demystifies
the curly braces: they capture, then inject.

:::

## Data-masking versus tidy-selection

This distinction is rarely taught explicitly and is the single biggest later
source of confusion. Spend time here; the mental model pays back for years.

The same column name is evaluated in one of two ways depending on the verb and the
argument position:

```
                       USER WRITES
                            |
                 +----------+----------+
                 |                     |
                 v                     v
         DATA-MASKING            TIDY-SELECTION
                 |                     |
   "names are values you      "names label columns
    compute with"             you want to pick"
                 |                     |
                 v                     v
  filter(body_mass_g > 4000)   select(species, body_mass_g)
  mutate(z = x + y)            pivot_longer(cols = ends_with("_mm"))
  summarise(m = mean(x))       across(.cols = where(is.numeric), ...)
  arrange(desc(x))
```

The principle:

- **Data-masking** treats column names as **values you compute with**.
- **Tidy-selection** treats column names as **labels you use to pick columns**.

To filter penguins by mass you compute `body_mass_g > 4000` per row: the values
matter, so it is data-masking. To select the mass column for a table you compute
nothing; only the name matters, so it is tidy-selection.

### The reference table

Memorise the first three rows; the rest is reference.

| | Data-masking | Tidy-selection |
|---|---|---|
| **Names refer to** | values to compute with | columns to pick |
| **Forward a bare name** | `{{ var }}` | `{{ var }}` |
| **Forward a string** | `.data[[var]]` | `all_of(var)` |
| **Forward a character vector** | `pick(all_of(vars))` | `all_of(vars)` |
| **Example verbs** | `filter`, `mutate`, `summarise`, `arrange`, `aes` | `select`, `across` (cols), `pivot_longer` (cols) |
| **Helpers (`starts_with`, `where`)** | no | yes |

Three things to internalise: both contexts forward a bare name with `{{ }}`; they
need *different* syntax for strings; and selection helpers work only in
tidy-selection.

### The same task in both contexts

```{r}
#| eval: false
# Task A: compute with a column's values -> data-masking
mean_by_species <- function(data, var) {
  data |> dplyr::summarise(mean = mean({{ var }}, na.rm = TRUE), .by = species)
}

# Task B: pick a column by name -> tidy-selection
pick_with_species <- function(data, var) {
  data |> dplyr::select(species, {{ var }})
}
```

Both forward the argument with `{{ var }}`. The semantics differ: one applies a
function to the column's values, the other extracts the column.

### Activity 1: identify the context

:::{.task}
::::{.task-header}
Activity 1: identify the context
::::
::::{.task-container}

For each call, say whether the column reference is data-masking or tidy-selection.

1. `filter(penguins, body_mass_g > 4000)`
2. `select(penguins, species, body_mass_g)`
3. `mutate(penguins, mass_kg = body_mass_g / 1000)`
4. `pivot_longer(penguins, cols = ends_with("_mm"))`
5. `summarise(penguins, mean(body_mass_g), .by = species)`
6. `select(penguins, starts_with("bill"))`
7. `arrange(penguins, desc(body_mass_g))`
8. `filter_out(penguins, body_mass_g > 4000)`
9. `summarise(penguins, across(where(is.numeric), mean))`

::::
:::


`r hide("Solution")`

1. **Data-masking.** A comparison is computed per row.
2. **Tidy-selection.** Columns picked by name.
3. **Data-masking.** A new value computed from existing values.
4. **Tidy-selection** (the `cols` argument).
5. **Data-masking** (both the summary and `.by`).
6. **Tidy-selection.** A helper picks by name pattern.
7. **Data-masking.** Sorting by computed values.
8. **Data-masking.** A comparison is computed per row.
9. **Mixed.** `across()` takes a tidy-selection in `.cols` and applies a function
   in a data-masking context: the bridge case.

The trickiest are 4 (different contexts at different argument positions) and 8
(`across()` as the bridge).


`r unhide()`

### Activity 2: fix the broken function (data-masking)

:::{.task}
::::{.task-header}
Activity 2: fix the broken function (data-masking)
::::
::::{.task-container}

Identify the problem and fix it.

```{r, eval = FALSE}
my_filter <- function(data, var, threshold) {
  data |> dplyr::filter(var > threshold)
}
my_filter(penguins, body_mass_g, 4000)
# Error: object 'var' not found
```

::::
:::


`r hide("Solution")`

`filter()` is data-masking but `var` is not embraced, so dplyr looks for a literal
column `var`, fails, and then meets the bare symbol `body_mass_g`. Embrace `var`:

```{r, eval = FALSE}
my_filter <- function(data, var, threshold) {
  data |> dplyr::filter({{ var }} > threshold)
}
```

`threshold` does not need embracing: it is an ordinary number, not a column
reference. Only column-name arguments get the treatment.


`r unhide()`

### Activity 3: fix the broken function (tidy-selection)

:::{.task}
::::{.task-header}
Activity 3: fix the broken function (tidy-selection)
::::
::::{.task-container}

A different problem. Identify and fix it.

```{r, eval = FALSE}
keep_columns <- function(data, cols) {
  data |> dplyr::select(cols)
}
keep_columns(penguins, c("species", "body_mass_g"))
# Error: Can't select columns past the end.
```

::::
:::


`r hide("Solution")`

In tidy-selection, a character vector is not a set of column references by
default. Wrap it in `all_of()`:

```{r, eval = FALSE}
keep_columns <- function(data, cols) {
  data |> dplyr::select(dplyr::all_of(cols))
}
```

If instead the function takes a *bare* name, embracing works: `select({{ cols }})`.
The right tool depends on whether the input arrives as a bare name (`{{ }}`) or as
a string or character vector (`all_of()`).


`r unhide()`

### Activity 4: the bridge

:::{.task}
::::{.task-header}
Activity 4: the bridge
::::
::::{.task-container}

Write `mean_of_columns(data, cols)` that takes a character vector of column names
and returns the mean of each, grouped by species. You need both contexts: select
the columns (they arrive as strings), then compute a mean for each.

::::
:::


`r hide("Solution")`

```{r, eval = FALSE}
mean_of_columns <- function(data, cols) {
  data |>
    dplyr::summarise(
      dplyr::across(dplyr::all_of(cols), \(x) mean(x, na.rm = TRUE)),
      .by = species
    )
}

mean_of_columns(penguins, c("body_mass_g", "flipper_length_mm"))
```

`across()` is the bridge. Its `.cols` argument is tidy-selection (so `all_of()`
accepts the string vector); its `.fns` argument is data-masking (applying a
function to each selected column). The bridge is the whole reason `across()`
exists: it lets tidy-selection helpers reach into data-masking verbs.


`r unhide()`

## Strings and dynamic names

When a column name arrives as a string, embracing does not apply: strings are
values, not expressions to capture. Use the `.data` pronoun in data-masking
contexts:

```{r}
#| eval: false
summarise_by_string <- function(data, var_name) {
  data |>
    dplyr::summarise(mean = mean(.data[[var_name]], na.rm = TRUE), .by = species)
}

summarise_by_string(penguins, "body_mass_g")
```

For *output* column names built from an argument, R's grammar forbids a string on
the left of `=`. The **walrus operator** `:=` allows it, combined with glue-style
name injection:

```{r}
#| eval: false
summarise_dynamic <- function(data, var) {
  data |>
    dplyr::summarise(
      "{{ var }}_mean" := mean({{ var }}, na.rm = TRUE),
      "{{ var }}_sd"   := sd({{ var }}, na.rm = TRUE),
      .by = species
    )
}

summarise_dynamic(penguins, body_mass_g)
# # A tibble: 3 x 3
#   species   body_mass_g_mean body_mass_g_sd
#   <fct>                <dbl>          <dbl>
# 1 Adelie               3701.           459.
```

Two things happen on the left of `:=`. `"{{ var }}_mean"` is glue-style injection,
replacing `{{ var }}` with the argument's name; `:=` is the assignment that accepts
a string on its left. Both are needed.

## Choosing the pattern

Activities 1 to 4 each drilled one pattern in isolation. In practice the hard part
is deciding *which* pattern a situation calls for, because the answer depends on
two independent questions: how does the column name arrive (bare name, string, or
character vector), and what is the context (data-masking or tidy-selection)? This
activity makes that decision the task.

### Activity 5: choose the pattern

:::{.task}
::::{.task-header}
Activity 5: choose the pattern
::::
::::{.task-container}

For each scenario, name the correct forwarding pattern, then write the function.
All four group by `species`.

(a) `var` arrives as a **bare name**; compute its mean (data-masking).

(b) `var_name` arrives as a **string**; compute its mean (data-masking).

(c) `cols` arrives as a **character vector**; select those columns (tidy-selection).

(d) `var` arrives as a **bare name**; compute its mean into an output column named
    `<var>_mean` (data-masking with a dynamic output name).

::::
:::


`r hide("Solution")`

Each scenario maps onto exactly one of the four working patterns.

```{r, eval = FALSE}
# (a) bare name, data-masking -> {{ }}
summarise_mean <- function(data, var) {
  data |>
    dplyr::summarise(mean = mean({{ var }}, na.rm = TRUE), .by = species)
}

# (b) string, data-masking -> .data[[ ]]
summarise_mean_str <- function(data, var_name) {
  data |>
    dplyr::summarise(mean = mean(.data[[var_name]], na.rm = TRUE), .by = species)
}

# (c) character vector, tidy-selection -> all_of()
select_cols <- function(data, cols) {
  data |>
    dplyr::select(species, dplyr::all_of(cols))
}

# (d) bare name, dynamic output name -> {{ }} with :=
summarise_named <- function(data, var) {
  data |>
    dplyr::summarise("{{ var }}_mean" := mean({{ var }}, na.rm = TRUE), .by = species)
}
```

These are the four patterns from the closing callout, one per scenario. The
decision procedure is: first ask whether the name arrives bare, as a string, or as
a vector; then, if it is a string or vector, ask whether the verb is data-masking
(`.data[[ ]]`) or tidy-selection (`all_of()`). A bare name uses `{{ }}` in either
context.


`r unhide()`

### Activity 6: predict the error

:::{.task}
::::{.task-header}
Activity 6: predict the error
::::
::::{.task-container}

Before running anything, predict what each call does: does it succeed, return the
wrong answer, or error? Then explain why in terms of the masking model.

```{r, eval = FALSE}
# (i) embracing a string
summarise_mean <- function(data, var) {
  data |> dplyr::summarise(mean = mean({{ var }}, na.rm = TRUE), .by = species)
}
summarise_mean(penguins, "body_mass_g")

# (ii) a string into a data-masking verb without .data
summarise_mean_str <- function(data, var_name) {
  data |> dplyr::summarise(mean = mean(var_name, na.rm = TRUE), .by = species)
}
summarise_mean_str(penguins, "body_mass_g")
```

::::
:::


`r hide("Solution")`

(i) **Returns the wrong answer (or a warning), not an error.** `{{ }}` captures the
expression `"body_mass_g"`, which is a string literal, not a column reference.
`mean()` then receives a single character value, and `mean()` of a character
returns `NA` with a warning. Embracing is for bare names; a string needs
`.data[[ ]]`.

(ii) **Returns the wrong answer.** `var_name` is not embraced, so dplyr first looks
for a column called `var_name`, fails, then finds `var_name` in the function
environment holding the string `"body_mass_g"`. It computes the mean of that single
string, again yielding `NA`. The string never reaches the data frame as a column
reference because nothing told dplyr to treat it as one; that is the job of
`.data[[var_name]]`.

The general lesson: passing the wrong *kind* of object to a masked verb rarely
errors loudly. It often produces a silent `NA`, which is more dangerous than an
error. Match the pattern to how the name arrives.


`r unhide()`

## The payoff: generalising `fit_model`

```{r}
#| eval: false
fit_model <- function(data) {
  lm(body_mass_g ~ flipper_length_mm, data = data)
}
```

### Activity

:::{.task}
::::{.task-header}
Generalise fit_model
::::
::::{.task-container}

Make `fit_model()` accept a response and predictor as bare column names,
defaulting to body mass and flipper length so existing calls still work. There is
a wrinkle: `lm()`'s first argument is a formula. Try both the simple `{{ }}`
approach and the explicit `rlang::new_formula()` approach.

::::
:::


`r hide("Solution")`

The simple approach works because `lm()`'s formula is itself a data-masked
context, so embracing inside it is honoured:

```{r, eval = FALSE}
fit_model <- function(data,
                      response  = body_mass_g,
                      predictor = flipper_length_mm) {
  lm({{ response }} ~ {{ predictor }}, data = data)
}

fit_model(penguins_clean)                                  # defaults
fit_model(penguins_clean, bill_length_mm)                  # new response
fit_model(penguins_clean, bill_length_mm, bill_depth_mm)   # both
```

The robust approach builds the formula explicitly, needed if you want to
manipulate it first:

```{r, eval = FALSE}
fit_model <- function(data,
                      response  = body_mass_g,
                      predictor = flipper_length_mm) {
  formula <- rlang::new_formula(
    lhs = rlang::enexpr(response),
    rhs = rlang::enexpr(predictor)
  )
  lm(formula, data = data)
}
```

`enexpr()` captures the unevaluated argument; `new_formula()` assembles it. Use the
simple version unless you need to manipulate the formula.


`r unhide()`

When you change a function's interface, pin the old behaviour down first with a
**characterisation test**, then confirm the change preserves it:

```{r}
#| eval: false
test_that("fit_model defaults reproduce previous behaviour", {
  m_old <- lm(body_mass_g ~ flipper_length_mm,
              data = test_data |> dplyr::filter(species == "Adelie"))
  m_new <- fit_model(test_data |> dplyr::filter(species == "Adelie"))
  expect_equal(coef(m_old), coef(m_new))
})
```

:::{.callout-note}

**You have now seen the four patterns that cover most practical cases:**

- `{{ }}` to forward bare names,
- `:=` for dynamic output names,
- `.data[[ ]]` for string column names,
- `all_of()` for character vectors of column names.

rlang has more machinery (`enquo()`, `!!`, `enquos()`, `!!!`) but most user code
never needs it. These four patterns are the working set.

:::
