# Block 3: Themes and scales {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```



```{r}
#| include: false


library(palmerpenguins)
library(tidyverse)
library(ggdist)
library(ggrepel)
library(ggtext)
library(patchwork)
library(colorspace)
library(colorBlindness)
library(ggview)
library(ragg)
library(scales)
library(here)


```



Two foundational pieces of `ggplot2` sit between the figure construction and the figure looking the way you want: **themes** (how the non-data elements appear) and **scales** (how the data axes are formatted). Mastery of both transforms what your figures look like without changing what they show. This block establishes the patterns we will then use throughout Block 4.

## Themes

A complete theme controls every non-data visual element: panel backgrounds, gridlines, text styling, axis ticks, margins, legend appearance. `ggplot2` ships several complete themes; we have already used `theme_minimal()` in earlier examples. The ones worth knowing:

| Theme | When to use |
|---|---|
| `theme_minimal()` | Default for most figures. Clean panel, retained gridlines, no panel border. |
| `theme_classic()` | Traditional scientific look. Axis lines, no gridlines, no panel border. |
| `theme_bw()` | Black-and-white panel with thin border and gridlines. |
| `theme_void()` | Strip everything. Useful for maps and bespoke composed layouts. |
| `theme_grey()` | The `ggplot2` default. *Do not use this.* The grey panel background fails the data-ink test. |

The `base_size` argument sets the text size for every text element in the figure proportionally. Setting `theme_minimal(base_size = 12)` once is preferable to setting `axis.text = element_text(size = 12)` and twelve similar lines: every text element will scale together when you change `base_size` for a different audience.

### Local modifications with theme()

The `theme()` function makes local modifications on top of any complete theme. Inside `theme()`, every named argument is a visual element of the figure and takes an `element_*()` function:

- `element_text()` for text elements (titles, axis labels, legend text)
- `element_rect()` for rectangular backgrounds (panel, plot, legend, strip)
- `element_line()` for lines (gridlines, axis lines, tick marks)
- `element_blank()` to remove an element entirely
- `ggtext::element_markdown()` for rich text in titles and labels (introduced in Block 4)

A representative set of modifications, applied to a complete theme:

```{r}
#| eval: true
#| message: false
#| warning: false
#| fig-height: 4
penguins_complete |> 
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm, colour = species)) +
  geom_point(alpha = 0.7) +
  colorspace::scale_colour_discrete_qualitative(palette = "Dark2") +
  labs(title = "Penguin bill and flipper dimensions",
       x = "Bill length (mm)",
       y = "Flipper length (mm)",
       colour = "Species") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),          # remove minor gridlines
    axis.title = element_text(face = "bold"),    # bold axis labels
    plot.title = element_text(face = "bold",     # bold title at 1.2 × base_size
                              size = rel(1.2)),
    legend.position = "bottom",                  # legend below
    plot.background = element_rect(fill = "white", colour = NA)
  )
```

The `rel()` function sets a size relative to `base_size`. `rel(1.2)` means "1.2 times the base text size", which adjusts automatically when you change `base_size`. Use `rel()` for any text size you set inside `theme()`; use absolute sizes only when you need to override the relationship.

### A project theme function

If you make the same theme modifications in every script, wrap them in a function. Here we make a simple custom `theme_project()`:

```{r}
#| eval: true
theme_project <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = rel(1.1)),
      plot.subtitle = element_text(colour = "grey40"),
      legend.position = "bottom",
      plot.background = element_rect(fill = "white", colour = NA),
      strip.text = element_text(face = "bold")
    )
}
```

Now every figure in the project that uses `theme_project()` changes to your house style. Altering this function propagates everywhere without editing individual scripts. Save the function in `R/functions/plot_functions.R` and source it at the top of any script that produces figures.

### Theme demonstration: same data, four themes

The case for moving away from `theme_grey()` is more persuasive shown than argued. The figure below renders the same scatter plot four times: once with the `ggplot2` default, once with `theme_minimal()`, once with `theme_classic()`, and once with our custom `theme_project()`. Look at each in turn and ask which one most clearly serves the data.

```{r}
#| eval: true
#| message: false
#| warning: false
#| fig-height: 8
#| fig-width: 9
base_fig <- penguins_complete |> 
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm, colour = species)) +
  geom_point(alpha = 0.7) +
  colorspace::scale_colour_discrete_qualitative(palette = "Dark2") +
  labs(x = "Bill length (mm)", y = "Flipper length (mm)", colour = "Species")

p_grey    <- base_fig + theme_grey(base_size = 11)    + labs(title = "theme_grey() — DEFAULT, AVOID")
p_minimal <- base_fig + theme_minimal(base_size = 11) + labs(title = "theme_minimal()")
p_classic <- base_fig + theme_classic(base_size = 11) + labs(title = "theme_classic()")
p_project <- base_fig + theme_project(base_size = 11) + labs(title = "theme_project()")

(p_grey + p_minimal) / (p_classic + p_project) 
```

:::{.callout-note}
**Reflection — Tufte's data-ink ratio.** Edward Tufte (1983) proposed that the ratio of data-ink to total-ink in a figure should be high: most of the visible pixels should encode data, not decoration. Look at the four panels above and ask, for each one, which visual elements carry information about the data, and which are just visual noise. The grey background panel of `theme_grey()` does not encode anything; the panel border in some themes does not encode anything; the minor gridlines may or may not, depending on whether the reader is reading off precise values. *Polish is a discipline.* The discipline is to remove anything that does not encode data, until removing more would damage readability.
:::

The `theme_project()` panel is the most restrained of the four; it removes the grey background, the minor gridlines, and the legend at the top-right; bolds the axis titles for slight emphasis; and places the legend below the panel where it does not compete with the data for the reader's primary axis of attention. Each of these choices is defensible against the data-ink principle.

### Audience-appropriate base_size

The audience matrix in Block 2 specified base font sizes for different reading contexts. Translated into `theme_project()` calls:

| Audience | `base_size` argument |
|---|---|
| Journal / lab report | `theme_project(base_size = 12)` (floor) |
| Conference poster | `theme_project(base_size = 18)` |
| Seminar talk | `theme_project(base_size = 16)` |
| Public engagement | `theme_project(base_size = 16)` |

:::{.callout-warning}
Setting individual text element sizes (`axis.text = element_text(size = 14)`) without adjusting `base_size` leaves other text elements at the default. The result is a figure with mismatched text sizes that looks unbalanced. Always change `base_size` first; tweak individual elements only as deliberate exceptions.
:::

## Scales

Where themes control the non-data parts of the figure, scales control how the data axes are rendered. Axes are position encodings, the most accurate visual channel in the encoding hierarchy from Block 1 — yet the breaks and labels chosen by `ggplot2` by default often undermine that accuracy by placing ticks at awkward intervals or formatting numbers in ways that obscure them. The `scales` package gives you the tools to fix this.

### Pretty breaks

The default break algorithm sometimes places ticks at awkward intervals. `scales::breaks_pretty()` and `scales::breaks_extended()` produce better-looking breaks at round numbers.

```{r}
#| eval: true
#| message: false
#| warning: false
#| fig-height: 4
penguins_complete |> 
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(bins = 30, fill = "#0072B2", colour = "white") +
  scale_x_continuous(
    breaks = scales::breaks_pretty(n = 6)
  ) +
  scale_y_continuous(
    breaks = scales::breaks_pretty(n = 5),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = "Body mass (g)", y = "Count") +
  theme_project()
```

The `expansion(mult = c(0, 0.05))` argument adds no padding at the bottom (so bars sit on the axis) and 5% padding at the top (so the tallest bar does not touch the upper edge). This is the convention for bar charts and histograms.

### Label functions

Number formatting matters. Thousands separators, percentage signs, properly formatted p-values, and currency prefixes are all small touches that distinguish a polished figure from a draft. The `scales` package provides label functions for each:

| Function | Use |
|---|---|
| `scales::label_number()` | Generic number formatting; arguments include `big.mark`, `accuracy`, `prefix`, `suffix` |
| `scales::label_comma()` | Thousands separator (shorthand for `label_number(big.mark = ",")`) |
| `scales::label_percent()` | Multiplies by 100 and adds `%` |
| `scales::label_dollar()` | Currency; the `prefix` argument supports any symbol (e.g. `"£"`) |
| `scales::label_scientific()` | Scientific notation for very large or very small numbers |
| `scales::label_log()` | Log-scale labels formatted as `10^n` |
| `scales::label_pvalue()` | P-values with `<` for very small values |
| `scales::label_wrap()` | Wrap long category labels across multiple lines |

```{r}
#| eval: true
#| message: false
#| warning: false
#| fig-height: 4
penguins_complete |> 
  count(island) |> 
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = fct_reorder(island, prop), y = prop)) +
  geom_col(fill = "#0072B2") +
  scale_y_continuous(
    breaks = scales::breaks_pretty(n = 5),
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = NULL, y = "Proportion of sampled penguins") +
  theme_project() +
  coord_flip()
```

Two improvements over the default: y-axis labels render as `30%` rather than `0.3`, and the breaks sit at sensible round percentages rather than at arbitrary values like `0.275`. Both changes happen entirely in the `scale_*` call; the data are not modified.

### Demonstration: before and after

The cumulative effect of theme and scale choices is easier to see than to describe. The figure below shows the same data twice — first with `ggplot2` defaults, then with `theme_project()` and `scales::label_*()` applied. The data are identical; only the styling has changed.

```{r}
#| eval: true
#| message: false
#| warning: false
#| fig-height: 4
#| fig-width: 9
base_data <- penguins_complete |> 
  count(island) |> 
  mutate(prop = n / sum(n))

p_default <- base_data |> 
  ggplot(aes(x = island, y = prop)) +
  geom_col() +
  labs(x = "Island", y = "prop", title = "Default theme and scales")

p_improved <- base_data |> 
  ggplot(aes(x = fct_reorder(island, prop), y = prop)) +
  geom_col(fill = "#0072B2") +
  scale_y_continuous(
    breaks = scales::breaks_pretty(n = 5),
    labels = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(x = NULL, y = "Proportion of sampled penguins",
       title = "theme_project() and scales::label_percent()") +
  theme_project()

p_default + p_improved
```

Three concrete changes between the panels: the y-axis labels render as percentages, the bars are ordered (by `fct_reorder`) rather than alphabetical, and the grey panel background is gone. The data-ink ratio has improved without altering a single number.

### Log scales

When a variable spans more than two orders of magnitude, transform to log. The `scales` package handles both breaks and labels:

```{r}
#| eval: false
scale_y_continuous(
  transform = "log10",
  breaks = scales::breaks_log(n = 5),
  labels = scales::label_log()
)
```

`label_log()` formats the labels as `10^1`, `10^2`, and so on — the proper notation for log-scaled data. The figure should also state the log transformation in either the axis title or the legend so the reader does not misread the scale as linear.

:::{.callout-note}
**Why this matters.** A figure built with `theme_project()` and properly formatted `scales::label_*()` functions reads as professionally produced even if the underlying analysis is straightforward. A figure with `theme_grey()` and default axis labels reads as a draft even if the analysis is sophisticated. Polish is a discipline, not an embellishment.
:::


:::{.callout-note}
**Break (15 minutes).** Energy is high after the participatory exercises in Blocks 1 and 2 and the coding-focused Block 3. The remaining blocks are coding-heavy and benefit from fresh attention.
:::


