#  Block 2: Github Automations {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false

library(usethis)
library(gitcreds)
library(renv)

```



## Bridging Block 1 and Block 2 

In Block 1 you built a repository (the `my-first-repo` clone) containing a Quarto
document that loads `tidyverse` and `palmerpenguins` and produces a figure of body
mass against flipper length. That figure was developed on the `new-plot` branch and
merged into `main`. Block 2 now adds two automated checks to that **same** repository,
so the work here extends what you already have rather than starting afresh.

Before continuing, confirm three things:

- You are on the `main` branch.
- You have pulled the merged work, so the figure code is present in your local `main`.
- Your Quarto file has a known name. Rename it to `analysis.qmd` if it does not already
  match, since every task below refers to that name.

### Set up renv in your penguins repository

Block 2 assumes your project tracks its dependencies with `renv`, which Block 1 did
not set up. Initialise it now, once, at the console:

```{r}
#| eval: false
renv::init()
renv::snapshot()
```

`renv::init()` creates the project library and the `renv.lock` file; `renv::snapshot()`
records the currently loaded packages (`tidyverse`, `palmerpenguins`) into the lockfile.
Commit `renv.lock` and the `renv/` infrastructure through the Git pane before proceeding.

::: {.callout-note}
With `renv.lock` committed, the "Before you start" checks below will pass rather than
report a missing lockfile. Everything in this block now operates on your penguins
repository: the hook guards its lockfile, and the Action renders its document.
:::

---

### What you will learn

By the end of this tutorial you will have two automated checks running in your
penguins project:

1. A **git hook** that runs on your machine every time you commit, checking that your
   `renv.lock` file is up to date.

2. A **GitHub Actions workflow** that runs on GitHub's servers every time you push,
   automatically rendering your penguins Quarto document and storing the output for
   download.

Neither requires manual triggering. Both connect directly to the project workflow you
established in Block 1.

---

### The mental model: two gates, two purposes

Before writing any code, it is worth understanding what these tools are and why you would
use both.

**A git hook** is a script that fires automatically when a specific git event occurs — in
this case, when you run `git commit`. It runs on your own machine, before the commit
completes. If the script fails, the commit is blocked and you see an error message
explaining why. This makes hooks ideal for catching problems *before* they leave your
machine.

**GitHub Actions** is a set of instructions stored in your repository that GitHub's
servers execute automatically when a specified event occurs — in this case, when you push.
To understand what "GitHub's servers" means in practice: imagine a brand-new computer with
nothing installed — no R, no packages, no files. GitHub provides that computer, downloads
your repository onto it, and follows your instructions step by step. This is what is meant
by a "clean environment": there are no local settings, no previously installed packages,
and no assumptions about your machine. If your code works there, it will work anywhere.

The two tools are complementary rather than interchangeable. The hook catches problems
locally before they are committed. Actions verifies that your code works in a clean
environment after they are pushed. Together they form two checkpoints:

```markdown
You write code
↓
git commit 
↓
[HOOK: renv check]
↓
git push 
↓
[ACTIONS: render Quarto]
```



### A note on YAML

The GitHub Actions workflow file is written in YAML (rhymes with "camel"). YAML is a
format for writing structured configuration — think of it as a way of expressing a nested
list using indentation rather than brackets. You do not need to learn YAML deeply for this
tutorial; the file is fully annotated and you only need to make small edits to adapt it to
your project. The one rule that matters: **indentation is significant**. Use spaces, not
tabs, and keep the indentation consistent with the examples shown.



### A note on r-lib/actions

The R community maintains a library of pre-built GitHub Actions specifically for R users,
available at [https://github.com/r-lib/actions](https://github.com/r-lib/actions). These
cover common tasks such as installing R, restoring renv packages, and running checks.
Rather than writing these steps from scratch, you can use them in your workflow file with
a single line. This tutorial uses two of them: `setup-r` and `setup-renv`. When you are
ready to go further, the r-lib/actions repository contains a full set of examples for
other tasks.



## Before you start

You should be working in the penguins repository from Block 1, on the `main` branch, with
the merged figure code present and your Quarto file named `analysis.qmd`. You should also
have run `renv::init()` and `renv::snapshot()` as set out in the bridge above, and committed
`renv.lock`.

**Check that renv.lock is committed.** In RStudio or Positron, open the Git pane (usually
in the top-right panel). If `renv.lock` appears in the list of changed files, it has not
been committed yet. If it does not appear, it is already committed and you are ready to
proceed.

If `renv.lock` is not yet committed, run the following in R and then commit the result
using the Git pane:

```r
renv::snapshot()
```

**Check your branch name.** In the Git pane, your current branch name is shown at the
top. If it says `main`, you are ready to proceed. If it says `master` or something else,
note the name — you will need to substitute it in the workflow file in Part 2.

**Confirm your penguins Quarto document.** Your project should contain `analysis.qmd`,
the document carrying the body-mass-against-flipper-length figure you built in Block 1.
If you have not renamed it yet, do so now so that the tasks below match.



## Part 1: Your first hook — checking renv synchronisation

### Why this matters

`renv` tracks your project's package dependencies in `renv.lock`. When you install,
update, or remove a package, `renv.lock` must be updated with `renv::snapshot()` to
reflect the change. If you forget, anyone who restores from your lockfile — including
yourself on a new machine, or GitHub Actions — will get a different set of packages. This
is one of the most common reproducibility failures in R projects and one of the easiest
to prevent automatically.

### How the hook works

Every time you run `git commit`, the hook calls `renv::status()`. If your installed
packages and `renv.lock` are out of sync, the commit is blocked and you are told to run
`renv::snapshot()`. If they are in sync, the commit proceeds normally. You do not interact
with the hook directly — it fires on its own.

### Setting up the hook with usethis

The `usethis` package provides a function that creates the hook file and sets the correct
file permissions in a single R call, with no terminal commands required.

First, install `usethis` if you do not have it:

```{r}
#| eval: false
install.packages("usethis")
```

Make the following file

scripts/hooks/pre-commit.sh:

```markdown
#!/usr/bin/env bash

echo "--- Checking renv synchronisation..."
"Rscript" --no-save --no-restore-data -e '
  if (!isTRUE(renv::status()$synchronized)) {
    message("\nrenv is out of sync with renv.lock.\nRun renv::snapshot() then commit again.\n")
    quit(status = 1)
  }
  message("renv check passed.")
'

```

:::{.callout-warning}
This script assumes `Rscript` is available on your system PATH, which is the
default on macOS and Linux. On Windows, R is not always added to the PATH. If
the hook fails with a "command not found" error, you 
replace `Rscript` in the script above with the full path to your installation:

```{r,eval = FALSE}
file.path(R.home("bin"), "R") # find your R.exe
```

for example `"/c/Program Files/R/R-4.5.3/bin/x64/Rscript"`. Adjust the version
number to match your installed version of R.

```markdown
#!/usr/bin/env bash

echo "--- Checking renv synchronisation..."
"/c/Program Files/R/R-4.5.3/bin/x64/Rscript" --no-save --no-restore-data -e '
  if (!isTRUE(renv::status()$synchronized)) {
    message("\nrenv is out of sync with renv.lock.\nRun renv::snapshot() then commit again.\n")
    quit(status = 1)
  }
  message("renv check passed.")
'

```

:::


```{r}
#| eval: false

usethis::use_git_hook(
  hook   = "pre-commit",
  script = readLines("scripts/hooks/pre-commit.sh") |> paste(collapse = "\n")
)

```

`usethis::use_git_hook()` writes the script into `.git/hooks/pre-commit` and makes it
executable. You do not need to use the terminal for any of this.

> **What is `.git/`?** Every git repository contains a hidden folder called `.git/` that
> stores git's internal data — your commit history, configuration, and hooks. You will not
> normally interact with it directly. It does not appear in RStudio's Files pane because
> it is hidden; you can see it in your operating system's file explorer if you enable
> hidden files.

### Watch the hook block a commit

We will now deliberately trigger the hook on your penguins repository so you can see it
work. Your existing figure colours the points by species; suppose you decide to format the
body-mass axis with thousands separators, which the `scales` package provides. Installing
`scales` without snapshotting is exactly the kind of desync the hook is designed to catch.

:::{.task}
::::{.task-header}
**Task: watch the hook block a commit — this commit is meant to fail**
::::

::::{.task-container}

This task deliberately triggers the hook so you can see it work. The first commit attempt
**is supposed to be blocked**. That is the demonstration, not an error on your part.

1. Install `scales` at the console (a package you have not yet snapshotted):

   ```r
   install.packages("scales")   # or renv::install("scales")
   ```

2. Edit the figure chunk in `analysis.qmd` to use it. For example, load the package and
   format the y-axis:

   ````md
   ```{{r}}
   #| label: fig-penguins
   #| fig-cap: "Body mass against flipper length, by species."
   library(tidyverse)
   library(palmerpenguins)
   library(scales)

   pal <- c(
     "Adelie"    = "#FF8C00",
     "Chinstrap" = "#A034F0",
     "Gentoo"    = "#159090"
   )

   penguins |>
     ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
     geom_point(aes(colour = species)) +
     scale_colour_manual(values = pal) +
     scale_y_continuous(labels = label_comma()) +
     theme_minimal()
   ```
   ````

   Save the file.

3. The act of installing `scales` **without snapshotting** is what has put your library
   out of sync with `renv.lock`. Stage both files in the Git pane and attempt to commit.

4. The commit is **blocked**. The hook reports that `renv` is out of sync and tells you to
   run `renv::snapshot()`.

5. Run `renv::snapshot()` at the console to record `scales` in the lockfile. Stage the
   updated `renv.lock` alongside your other changes and commit again. This time it
   succeeds.

::::
:::

The point to take from step 3 is that the block was caused by the un-snapshotted install,
not by editing the document. Saving a file never trips the hook; an installed package
missing from `renv.lock` does.

::: {.callout-warning}
## Bypassing the hook
Occasionally you will need to commit despite the hook, for instance when it raises a false
positive or you must record an urgent change. The `--no-verify` flag skips the hook for a
single commit, leaving it in place for the next one:

```bash
git commit --no-verify -m "your message"
```

Use this as an escape hatch, not a routine. A hook you bypass by habit checks nothing.
RStudio's Commit button has no equivalent tick box, so a bypass must be done at the
terminal — which is a useful piece of friction.
:::

::: {.callout-note}
### Removing the hook permanently
If you want to disable the check altogether, delete or rename the hook file at
`.git/hooks/pre-commit` in your operating system's file explorer (it is hidden by
default). This is reversible only by recreating the hook with
`usethis::use_git_hook()`, so prefer `--no-verify` for one-off cases and reserve removal
for when you no longer want the check at all.
:::

## Part 2: Your first GitHub Action — rendering the penguins document

### Why this matters

Rendering your penguins document on GitHub's clean servers achieves two things. First, it
confirms that the document renders correctly using only the packages in `renv.lock`,
catching the common situation where something works on your machine but fails elsewhere.
Second, it produces a rendered HTML file that anyone with access to the repository can
download, without needing R or Quarto installed.

### usethis::use_github_action()

The `usethis` package provides `use_github_action()`, which sets up common GitHub Actions
workflows with a single R function call. Running it without arguments shows you the
available options:

```
usethis::use_github_action()
```

This opens an interactive menu. The built-in options are primarily aimed at R package
developers (running checks, computing test coverage, building documentation sites).
For an analysis project with a Quarto document, none of the built-in templates fits
directly, so we will write the workflow file manually. However, `use_github_action()` is
worth knowing about: as your use of GitHub grows, it provides a fast path to common
workflows without writing YAML from scratch. The r-lib/actions repository
([https://github.com/r-lib/actions](https://github.com/r-lib/actions)) lists all
available options and their corresponding `use_github_action()` calls.

### Step 1: Enable write permissions in your repository

GitHub Actions needs permission to write to your repository. Granting this now avoids a
common failure mode later.

1. Go to your repository on GitHub.com
2. Click **Settings** in the top navigation bar
3. Click **Actions** in the left sidebar, then **General**
4. Scroll to *Workflow permissions*
5. Select **Read and write permissions**
6. Click **Save**

### Step 2: Pin your R version, then create the workflow file

Because this workflow restores your exact packages from `renv.lock`, it must run on the R
version those packages were recorded under. Find yours by running `R.version.string` at
the console and note the number, for example `4.5.2`.

Create the following folder structure in your project. You can do this in RStudio's Files
pane: New Folder → `.github`, then inside that New Folder → `workflows`.

```
.github/
  workflows/
    render-quarto.yml
```

Open `render-quarto.yml` and paste the following. It restores the exact package versions
recorded in the `renv.lock` you built in Part 1, so the document renders against the same
packages you used locally. Every section is explained in the comments.

```yaml
# The name shown in the Actions tab on GitHub.
name: Render penguins analysis

# When to run this workflow.
# 'push' means: every time you push commits to GitHub.
# Adjust the branch name below if yours is not 'main'.
on:
  push:
    branches: ["main"]

jobs:
  render:
    runs-on: ubuntu-latest

    steps:
      # Step 1: Download your repository files onto GitHub's machine.
      - name: Check out repository
        uses: actions/checkout@v4

      # Step 2: Install R. Set the version to match your renv.lock,
      # i.e. the number from R.version.string on your machine.
      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: "4.5.2"   # set to YOUR version from R.version.string
          use-public-rspm: true

      # Step 3: Install renv and restore the exact packages from renv.lock.
      # Because this reads renv.lock, there is no package list to maintain
      # here: it is the same lockfile your hook guards in Part 1.
      - name: Set up renv
        uses: r-lib/actions/setup-renv@v2
        env:
          RENV_CONFIG_PAK_ENABLED: "true"

      # Step 4: Install Quarto so we can render the document.
      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      # Step 5: Render the penguins document.
      - name: Render Quarto document
        run: quarto render analysis.qmd

      # Step 6: Save the rendered output so you can download it.
      - name: Upload rendered output
        uses: actions/upload-artifact@v4
        with:
          name: rendered-analysis
          path: analysis.html
          retention-days: 7
```

Because `setup-renv` restores from `renv.lock`, and your lockfile now contains
`tidyverse`, `palmerpenguins`, and `scales` from Part 1, there is no `packages:` block to
edit. This is the payoff of the renv approach: the workflow's package list and your
project's package list cannot drift apart, because they are the same file.

:::{.callout-warning}
The workflow file refers to `analysis.qmd` and `analysis.html`. If your Quarto file has a
different name, change both the `.qmd` and `.html` references everywhere they appear,
including the upload path. A mismatch produces a failure whose cause is not obvious from
the log.
:::

:::{.callout-note}
The `@v2` and `@v4` suffixes pin each action to a major version, which protects
you from unexpected breaking changes. These major versions are occasionally
superseded; if a workflow begins to fail after working previously, check the
action's repository for a newer version.
:::

:::{.callout-warning collapse= "true"}

### Match the R version
If the `r-version` you pin differs from the version recorded in `renv.lock`,
`setup-renv` may rebuild packages from source, which makes the first run noticeably
slower and can occasionally fail on a package that needs system libraries. Setting
`r-version` from your own `R.version.string`, as Step 2 instructs, is the mitigation.

If we set up a workflow that was `renv` and `R` version agnostic:

````markdown
# The name shown in the Actions tab on GitHub.
name: Render penguins analysis

# When to run this workflow.
# 'push' means: every time you push commits to GitHub.
# Adjust the branch name below if yours is not 'main'.
on:
  push:
    branches: ["main"]

jobs:
  render:
    runs-on: ubuntu-latest

    steps:
      # Step 1: Download your repository files onto GitHub's machine.
      - name: Check out repository
        uses: actions/checkout@v4

      # Step 2: Install R. 'release' tracks the current release version,
      # so there is no version number to maintain here.
      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: "release"
          use-public-rspm: true

      # Step 3: Install the packages your document needs.
      # List every package loaded in analysis.qmd here.
      - name: Install R packages
        uses: r-lib/actions/setup-r-dependencies@v2
        with:
          packages: |
            any::palmerpenguins
            any::tidyverse
            any::scales

      # Step 4: Install Quarto so we can render the document.
      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      # Step 5: Render the penguins document.
      - name: Render Quarto document
        run: quarto render analysis.qmd

      # Step 6: Save the rendered output so you can download it.
      - name: Upload rendered output
        uses: actions/upload-artifact@v4
        with:
          name: rendered-analysis
          path: analysis.html
          retention-days: 7


````

:::

### Step 3: Commit and push

Save the file, then commit and push using the Git pane or terminal:

```
git add .github/workflows/render-quarto.yml
git commit -m "chore: add Quarto rendering workflow"
git push
```

Note that the pre-commit hook will run before this commit. It will pass because you have
not changed any packages since your last snapshot in Part 1.

### Step 4: Watch the workflow run

1. Go to your repository on GitHub.com
2. Click the **Actions** tab in the top navigation bar
3. You will see a workflow run listed, named *Render penguins analysis*
4. Click on it, then click the **render** job
5. Each step expands to show its output as it runs

A green tick next to a step means it passed. A red cross means it failed and the job has
stopped. The workflow typically takes three to five minutes on first run because packages
are being restored; on subsequent runs, caching makes the renv step much faster.

### Step 5: Download the rendered output

When the run completes with a green tick:

1. Scroll to the bottom of the run page
2. Under **Artefacts**, click **rendered-analysis**
3. A zip file downloads containing `analysis.html`

This is your penguins document, rendered on GitHub's servers from the exact package
versions in your lockfile, without any action on your part beyond pushing a commit.

---

### Reading a failure

Failures are normal, particularly on first run. Click on the red cross next to the
failing step to see the full log. The four most common causes are:

**`renv.lock` is not committed**

```
Error in renv::restore() : renv.lock does not exist
```

Fix: run `renv::snapshot()` locally, commit `renv.lock`, and push again.

**A package used in the document is not in `renv.lock`**

```
Error in library(somepackage) : there is no package called 'somepackage'
```

Fix: install the package locally, run `renv::snapshot()`, commit the updated `renv.lock`,
and push. This is precisely the situation Part 1's hook is designed to prevent before the
commit ever leaves your machine.

:::{.callout-note}

Renv looks for explicit library() in your scripts and Quarto documents. If you load a package with library() but forget to snapshot, the workflow will fail with this error. If you load a package without library() — for example, using `dplyr::filter()` without loading dplyr — the workflow will not know to install it and will fail with a different error: "could not find function 'filter'". In either case, the fix is the same: install the package locally, snapshot, commit, and push.

:::

**The Quarto document contains a code error**

The log will show the R error message from inside the document, with a line number. Fix:
correct the code locally, render the document in RStudio or Positron to confirm it works,
commit, and push.

**The workflow does not appear in the Actions tab**

The trigger branch name in the workflow file does not match your actual branch. Fix: open
the workflow file and change `"main"` to your branch name, commit, and push.

---

## How the two tools work together

With both in place, your penguins workflow now looks like this:

| Event | What fires | What it catches |
|---|---|---|
| `git commit` | Pre-commit hook | renv out of sync with `renv.lock` |
| `git push` | GitHub Actions | Render failures; missing packages; clean-environment problems |

The hook catches the most common problem before it is committed. Actions catches problems
that are invisible in your local environment and produces output anyone can download.
Neither replaces the other.

---

## Going further: running code on a schedule

So far, the workflow has been triggered by a push — useful for testing, but not always
the most natural fit. For many analytical tasks, a scheduled run is more appropriate: a
report that refreshes every morning, or a data file that is updated weekly, without you
having to do anything at all.

GitHub Actions supports scheduled runs using what are called **cron jobs**. A cron
expression is a compact way of specifying a time schedule. Replace the `on: push:`
trigger in your workflow file with the following:

```yaml
# Run every day at 9:00am UTC
on:
  schedule:
    - cron: '0 9 * * *'
```

The five fields in a cron expression represent, in order: minute, hour, day of month,
month, and day of week. An asterisk means "every". So `0 9 * * *` means: at minute 0 of
hour 9, every day of every month. The time is always in UTC, which is one hour behind
British Summer Time and five hours ahead of Eastern Time — adjust accordingly.

The syntax is not intuitive to read or write from memory. The website
[crontab.guru](https://crontab.guru) translates cron expressions into plain English and
lets you build expressions interactively. It is worth bookmarking.

Some useful examples:

| Cron expression | When it runs |
|---|---|
| `0 9 * * 1` | Every Monday at 9:00am UTC |
| `0 8 1 * *` | The first day of every month at 8:00am UTC |
| `*/30 * * * *` | Every 30 minutes |
| `0 9 * * 1-5` | Weekdays at 9:00am UTC |

You can also combine a schedule with a push trigger, so the workflow runs both on demand
and on a timer:

```yaml
on:
  push:
    branches: ["main"]
  schedule:
    - cron: '0 9 * * 1'
```

---

## Going further: committing output back to the repository

The workflow in Part 2 uploads the rendered document as a temporary artefact, which
GitHub deletes after seven days and which requires a manual download to access. An
alternative is to have the workflow commit the rendered file back into the repository
directly, making it a permanent part of your commit history and accessible at a fixed
URL.

Add the following step after the render step, replacing the upload-artifact step:

```yaml
      - name: Commit rendered output to repository
        run: |
          # Set the git identity for this commit.
          # These details appear in the commit history.
          git config --local user.email "actions@github.com"
          git config --local user.name "GitHub Actions"

          # Stage the rendered file.
          git add analysis.html

          # Commit it. The || handles the case where nothing has changed:
          # if there is no new output, git commit would normally fail with
          # an error. The || tells bash: if the left side fails, run the
          # right side instead — in this case, just print a message and
          # continue without failing the workflow.
          git commit -m "chore: update rendered analysis" || echo "No changes to commit"

          # Push the commit back to GitHub.
          git push origin || echo "No changes to commit"
```

This pattern is particularly useful for scheduled workflows: the workflow runs on a timer,
re-renders the document with current data, and pushes the updated HTML back to the
repository automatically. The `||` operator is a bash pattern worth understanding — it
means "or", and separates a command from a fallback to run if the first command fails.

Note that this approach writes to your repository's main branch directly, which adds
commits generated by GitHub Actions to your history. Whether this is desirable depends on
your project; for a solo analyst it is usually fine.

---

## Going further: projects without renv

The workflow in Part 2 uses `r-lib/actions/setup-renv@v2`, which requires a committed
`renv.lock` file. If you encounter a project or tutorial that does not use renv, you will
see a different approach: `r-lib/actions/setup-r-dependencies@v2`, which installs packages
listed explicitly in the workflow file:

```yaml
- name: Install packages
  uses: r-lib/actions/setup-r-dependencies@v2
  with:
    packages: |
      any::ggplot2
      any::dplyr
      any::quarto
```

This is less reproducible than `renv` because it installs the latest available version of
each package rather than the exact version you used locally. For your own projects — such
as the penguins repository you have built here — renv and `setup-renv` is the better
choice. For adapting workflows from others who do not use renv, `setup-r-dependencies` is
what you will encounter.

---

## Where to go next

The r-lib/actions repository ([https://github.com/r-lib/actions](https://github.com/r-lib/actions))
contains example workflows for a wide range of R tasks. Each example includes its
`usethis::use_github_action()` call where one exists. The patterns you have learned here
transfer directly: adjust the trigger, swap the render step, change what is uploaded.

**Render only on pull requests** rather than every push, to keep the per-push feedback
loop fast:

```yaml
on:
  pull_request:
    branches: ["main"]
```
