# Block 1: AI-Assisted R Programming {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```


## Prerequisites

You need a GitHub account, a working IDE (RStudio, Positron, or VS Code), (ideally)  an active GitHub Copilot subscription (the Education plan is sufficient and provides cloud-agent access), and basic Git and GitHub fluency at the level of cloning a repository, making a branch, opening a pull request, and merging. Part 6 also assumes a working knowledge of `renv` for project-local package management, and a first acquaintance with GitHub Actions as the mechanism that runs automated workflows on a repository. All of these are assumed from prior sessions.


## Learning outcomes

By the end of the workshop, you will be able to:

1. Articulate what a large language model is, and explain what choosing between underlying models materially affects.

2. Use GitHub Copilot inline completion in their chosen IDE, and identify what context the tool can and cannot see.

3. Use AI assistance to debug code they have written, and recognise the cases in which AI is most and least useful for debugging.

4. Use GitHub Copilot Chat in Ask mode for planning, error explanation, and approach selection, and distinguish it from inline completion.

5. Distinguish workspace-level custom instructions, scoped instructions, Agent Skills, and agent-facing files (`AGENTS.md` and custom agent profiles) as separate configuration mechanisms, and produce a working `copilot-instructions.md` for their own research.

6. Distinguish in-IDE Agent mode from the asynchronous GitHub cloud coding agent, and use the cloud agent to functionalise a script step under a characterisation test, reading continuous-integration checks to confirm the result is reproducible while keeping in their own hands the reference that makes the verification non-circular.


## Materials

Participants will work with a synthetic biological dataset (`voles_metabolism.csv`) and a workshop template repository that they fork at the start of the workshop. The template uses `renv` for project-local packages, so after forking and opening the project, run `renv::restore()` once to install the locked versions. This matters in Part 6, where the same lockfile is restored by the continuous-integration checks and by the cloud agent.


## A note on platform variation

Where workflow behaviour depends on which IDE participants are using, the relevant differences are noted in the text. RStudio is the most constrained environment: inline completion works, browser-based Copilot Chat works, but in-IDE Chat panels and in-IDE Agent mode do not. Positron and VS Code support all features. Where the workshop refers to in-IDE Agent mode, RStudio users should follow the cloud-agent material instead, which is platform-agnostic and is the focus of Part 6.

:::{.callout-note}

The Posit company is actively working on AI integration in RStudio and has started the "Posit Assistant" in beta testing

:::

**Verification is the workshop's only non-negotiable.** AI tools generate plausible code quickly. They do not guarantee correctness, and they do not relieve you of responsibility for what your code does. Every exercise in this workshop ends with a verification step. Treat that step as the most important part of the exercise, not an afterthought.


## Framing and mental models

This block establishes the conceptual scaffold that the rest of the workshop builds on. There are no participant coding exercises; the block is presentation, short discussion, and a comprehension check.

*⏱ ~25minutes* 


## What is a large language model?

A large language model (LLM) is a statistical model trained to predict the next token (roughly, the next fragment of text) given the tokens before it. Modern LLMs are trained on a substantial portion of the public internet, including code repositories, technical documentation, and prose. GitHub Copilot uses LLMs under the hood, with several providers currently available: Anthropic's Claude family, OpenAI's GPT family, and Google's Gemini family. Copilot exposes a model selector in the chat interface; the choice matters because models differ measurably in code quality, instruction-following, and propensity to hallucinate library calls or function signatures that do not exist.

For the present workshop, you do not need to develop strong opinions about which model is best. You do need to know that the choice exists, that it changes the output, and that if a model produces consistently poor suggestions for your work, switching is a free experiment.


## Two modes of execution

Throughout the workshop you will encounter two fundamentally different modes in which Copilot can act on your behalf. The distinction is the most important conceptual point in Part 1 and recurs in Parts 5 and 6.

The first mode is synchronous and in-IDE. Inline completion, Copilot Chat panels, and in-IDE Agent mode all run inside your editor. They respond in seconds, you watch their output as it appears, and you can interrupt at any point. The scope is limited to whatever the IDE chooses to send as context: the current file, perhaps neighbouring tabs, perhaps the workspace.

The second mode is asynchronous and runs in the cloud. The GitHub cloud coding agent takes an issue you have written, creates its own branch, edits files, runs tests, opens a pull request, and waits for your review. You do not watch it work. You verify by reviewing the pull request after the agent has finished, which typically takes a few minutes. The scope is the entire repository the agent is working in.

Both modes are agents in some sense, but they sit at opposite ends of a synchrony spectrum, and they invite different verification strategies. The synchronous mode invites you to watch and interrupt; the asynchronous mode invites you to write a precise specification, walk away, and review what comes back. That specification need not be rewritten for every task. As Part 5 shows, much of it can be captured once in reusable instructions and skills, so that by Part 6 the issue you hand the agent can be short while the standards behind it stay exacting.

| Property | In-IDE (synchronous) | Cloud agent (asynchronous) |
|---|---|---|
| Where it runs | Your editor | GitHub's servers |
| What you see | Real-time output | A completed pull request |
| Speed of feedback | Seconds | Minutes |
| Interruption | Yes, at any point | No, only by closing the PR |
| Scope of action | Files in editor context | Whole repository |
| IDE availability | Positron, VS Code; RStudio chat only | Any (browser-based) |
| Verification model | Watch and steer | Review and approve |


## Why the context window matters

LLMs do not have memory between calls. Each interaction starts fresh, with whatever you supply as context. The "context window" is the upper limit on how much text the model can consider in a single call. This is the technical fact that motivates everything else in the workshop: instructions exist to inject persistent context into a system that has none; skills exist to make procedural context available on demand; well-written issues exist because the cloud agent has nothing else to go on.

If you remember nothing else from Part 1, remember that the AI sees only what you give it. The rest of the workshop is mostly about giving it the right things.


## Discussion

Allocate five minutes to a think-pair-share. Two questions:

1. What kinds of task would you trust to an unattended cloud agent that runs while you are not watching?

2. What kinds of task would you not delegate to AI at all?

3. What AI providers do you use (if any)?

4. How do you use AI tools at the moment? Chat-in browser, IDE chats, terminal, agents?

The point of the discussion is not to reach consensus. It is to surface the criteria participants are already applying intuitively, which will inform their use of these tools after the workshop.


## Check your understanding

**Which statement best describes the difference between in-IDE Agent mode and the GitHub cloud coding agent?**

`r longmcq(c(
  "They are the same tool with different names",
  answer = "In-IDE Agent mode runs synchronously in your editor while you watch; the cloud agent runs asynchronously on GitHub's servers and produces a pull request",
  "In-IDE Agent mode is paid and the cloud agent is free",
  "The cloud agent runs in your editor and the in-IDE agent runs on GitHub's servers"
))`

**True or false: a large language model has no memory between calls, so each interaction starts with only the context you supply.** `r torf(TRUE)`


