# Block 2: AI - Inline completions {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```


Inline completion is the most familiar AI-coding feature: ghost text appears as you type, suggesting a continuation. You accept with Tab, reject with Esc, or ignore by continuing to type. This block teaches what context inline completion uses, and the practical consequence: comments matter more than anything else.

*⏱ ~35 minutes*


## Setup verification

Verify that inline completion is enabled and working in your chosen IDE.

**RStudio:** Tools → Global Options → Copilot → Enable GitHub Copilot, then sign in to GitHub.

**Positron:** Copilot is provided through the bundled Positron Assistant extension. Open the Copilot panel in the sidebar and sign in.

**VS Code:** Install the GitHub Copilot extension from the marketplace, then sign in via the prompt that appears.

In all three, verify by opening a **blank R script** and start typing `# Calculate the mean of 1 to 100`. After a second or two you should see grey ghost text suggesting `mean(1:100)` or similar (If nothing appears try hitting enter to move the blinking cursor to the next line).


## The context principle

Inline completion sees, at minimum, the current file. In Positron and VS Code (and likely in newer RStudio versions, though the behaviour has changed historically and is worth checking against current documentation rather than asserting), it may also draw on other open tabs. It does not search your project for unrelated files, and it does not remember previous sessions.

The practical consequence: the comment immediately before your cursor, and the code above it in the same file, dominate what you get. Everything else has at best a modest effect.


## Activity: The context demonstration

:::{.task}
::::{.task-header}
The context demonstration
::::
::::{.task-container}

For this exercise you will use `voles_metabolism.csv`, a small synthetic biological dataset deliberately chosen because its column names are unlikely to feature in Copilot's training data. The columns are `subject_id`, `sex`, `treatment_arm`, `mass_pre_g`, `mass_post_g`, and `metabolic_rate_post`. The data record a hypothetical metabolism trial in voles with three treatment arms and two sexes, sample size 72.

::::
:::

**1. Start with no context.** Open a blank R script. Type only this single comment **into your script**, then wait:

```{r}
#| eval: false
# calculate the mean change in mass by treatment
```

Record what Copilot suggests. With no `library()` call, no data loaded, and no column names visible, Copilot has to guess. It will produce something syntactically plausible but most likely wrong for your dataset, because it does not know what your columns are called.

**2. Restart with full context.** Start a new blank script. Add the following at the top:

```{r}
#| eval: false
# Voles metabolism trial: effect of treatment on body mass
# 72 voles, three treatment arms (Control, Low_dose, High_dose), two sexes
# Native pipe |>, tidyverse

library(tidyverse)

voles <- read_csv(here::here("data", "voles_metabolism.csv"))
glimpse(voles)
```

Then add the same comment as before, on a new line:

```{r}
#| eval: false
# calculate the mean change in mass by treatment
```

Record what Copilot suggests this time. The expected output resembles:


`r hide("Solution")`

```{r, eval = FALSE}

voles |>
  mutate(mass_change_g = mass_post_g - mass_pre_g) |>
  summarise(
    mean_change = mean(mass_change_g, na.rm = TRUE),
    .by = treatment_arm
  )

```


`r unhide()`

The exact form will vary between participants and between sessions; that variation is itself part of the lesson.

How different were the two outputs? Did the second version use column names that exist in the data? Did Copilot construct `mass_change_g` itself or did you have to prompt for it?


## Activity: Build a small analysis

:::{.task}
::::{.task-header}
Build a small analysis
::::
::::{.task-container}

Working on the same script you set up in Exercise 1, drive inline completion through a four-step analysis. Write the comment for each step first, then accept, reject, or modify the suggestion.

**1.** Filter to subjects with non-missing `metabolic_rate_post` and create a derived variable `mass_change_g`.

**2.** Produce a summary table of mean and standard deviation of `mass_change_g` by `treatment_arm` and `sex`, with sample sizes per cell.

**3.** Fit a linear model of `metabolic_rate_post` on `treatment_arm` and `sex`. Inspect the coefficients.

**4.** Produce one diagnostic plot relating `mass_change_g` to `metabolic_rate_post`, coloured by `treatment_arm`.

For each step, your deliverable is the comment you wrote, the code Copilot suggested, and a one-line note on whether you accepted it, modified it, or rejected it and wrote something else. The point of recording this is not bureaucratic; it is to make you notice when you accepted something without inspecting it.

Work in pairs. Each partner attempts steps 2 and 4 independently before comparing. The pair-comparison stage typically surfaces interesting variation: different suggestions, different acceptance decisions, and (often) different bugs.

::::
:::

## Check your understanding

**The single most reliable way to improve the quality of inline completion suggestions in a new script is to:**

`r longmcq(c(
  "Switch to a different underlying model",
  "Open more files in the editor as additional context",
  answer = "Write a detailed comment immediately before the code you want, including data frame name and column names",
  "Restart the IDE so Copilot has a fresh context"
))`


