# Part 3: Debugging with AI  {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```

AI is at its most useful for debugging code you wrote. You know what you intended; the AI can spot syntactic mistakes you missed, suggest fixes for runtime errors, and help you interpret unfamiliar error messages. The verification cost is low, because you are comparing the AI's diagnosis against your own intention rather than against an unknown ground truth.

This block uses four error types, ordered by how much the debugger needs to know. The first error is a typo Copilot can catch with no data; the second requires knowledge of the data values; the third requires knowledge of ggplot2's grammar; the fourth runs cleanly, returns a plausible answer, and is wrong, so that only knowledge of the expected result reveals it. The differences matter, because they show where AI can substitute for documentation lookup, where it cannot substitute for inspecting the data, and where it cannot substitute for knowing the biology.

*⏱ ~40 minutes* 


## Setup

Have the voles dataset loaded:

```{r}
#| eval: false
library(tidyverse)
voles <- read_csv(here::here("data", "voles_metabolism.csv"))
```


## Exercise 1: A syntax error

You wrote the following to count voles per treatment:

```{r}
#| eval: false
ggplot(voles, aes(x = treatment_arm)) +
  geom_barchart()
```

R returns:

```
Error in geom_barchart(): could not find function "geom_barchart"
```

**1. Paste both the code and the error message into your script as a block comment, then add `# Fix this code:` and let Copilot suggest a continuation:**

```{r}
#| eval: false
# This code gives an error:
# ggplot(voles, aes(x = treatment_arm)) +
#   geom_barchart()
#
# Error: could not find function "geom_barchart"
#
# Fix this code:
```

Before accepting any suggestion, write a single sentence explaining why the original failed. Your sentence is the verification step. If you can write the sentence, you understood the fix; if you cannot, the AI has solved the problem but you have not learned anything.


`r hide("Solution")`

```{r, eval = FALSE}

ggplot(voles, aes(x = treatment_arm)) +
  geom_bar()

```

The function is `geom_bar()`, not `geom_barchart()`. ggplot2's geom functions use short names; there is no `_chart` suffix anywhere in the package. The error message "could not find function" is the standard R response when a function does not exist in any loaded package, and it is the easiest class of error to recognise.


`r unhide()`


## Exercise 2: A logic error

You wrote the following to filter to high-dose voles and compute mean post-treatment mass:

```{r}
#| eval: false
high_dose <- voles |>
  filter(treatment_arm == "high_dose")

high_dose |>
  summarise(mean_mass = mean(mass_post_g))
```

The code runs without error and returns:

```
# A tibble: 1 × 1
  mean_mass
      <dbl>
1       NaN
```

This is harder than the syntax error, because there is no error message to paste. The code is doing exactly what you asked it to; the bug is in what you asked for. AI cannot diagnose this without first knowing what your data look like.

**1. Before asking AI for help, inspect the data.** Three diagnostic calls take seconds and tell you everything you need to know:

```{r}
#| eval: false
voles |> distinct(treatment_arm)
voles |> count(treatment_arm)
nrow(high_dose)
```

The first two reveal that the factor levels are `Control`, `Low_dose`, and `High_dose` (initial capital, underscore between words). The third returns zero. At this point you have already diagnosed the bug without involving an AI.

**2. With the diagnostic information in hand, paste both the code and the diagnostic output into Chat (or as a block comment in your script) and ask for confirmation of the diagnosis:**

```
# I am trying to compute mean post-treatment mass for high-dose voles. My code:

#  high_dose <- voles |>
#    filter(treatment_arm == "high_dose")

#  high_dose |>
#    summarise(mean_mass = mean(mass_post_g))

# returns NaN. But voles |> distinct(treatment_arm) returns
# "Control", "Low_dose", "High_dose". I think the filter is failing
# because of the case. Is that the only problem?
```

The exercise is to ask the question with the diagnostic context already in place, not to ask AI to diagnose blind. AI is faster than you at typo-spotting and syntax recall, but slower than `distinct()` at telling you what is actually in your data.


`r hide("Solution")`

```{r, eval = FALSE}

high_dose <- voles |>
  filter(treatment_arm == "High_dose")

high_dose |>
  summarise(mean_mass = mean(mass_post_g, na.rm = TRUE))

```

The capitalisation now matches the actual factor level. Adding `na.rm = TRUE` is precautionary; for this dataset there are no missing values in `mass_post_g`, but in real biological data there usually are.


`r unhide()`


## Exercise 3: An aesthetic-mapping error

You wanted a scatter plot of post-treatment metabolic rate against post-treatment mass, coloured by treatment arm:

```{r}
#| eval: false
ggplot(voles, aes(x = mass_post_g, y = metabolic_rate_post)) +
  geom_point(color = treatment_arm) +
  labs(title = "Metabolic rate by mass")
```

R returns:

```
Error in layer(): object 'treatment_arm' not found
```

**1. Paste code and error as before, and ask Copilot for a fix. Before accepting, write a sentence on why ggplot2 refused to find `treatment_arm`.**


`r hide("Solution")`

```{r, eval = FALSE}

ggplot(voles, aes(x = mass_post_g, y = metabolic_rate_post)) +
  geom_point(aes(colour = treatment_arm)) +
  labs(title = "Metabolic rate by mass")

```

`treatment_arm` is a column in the data, so the mapping must go inside `aes()`. The `color =` argument outside `aes()` expects a fixed value such as `"red"` or `"#0072B2"`, not a variable name. When ggplot2 looked for an object called `treatment_arm` in the global environment, it did not find one, and produced the error. British spelling `colour` works as an alias for `color` in ggplot2.


`r unhide()`


## Exercise 4: A correct-looking but wrong result

You want the mean change in body mass by treatment arm. You write:

```{r}
#| eval: false
voles |>
  mutate(mass_change_g = mass_pre_g - mass_post_g) |>
  summarise(mean_change = mean(mass_change_g), .by = treatment_arm)
```

The code runs. There is no error and no `NaN`. It returns something like:

```
# A tibble: 3 × 2
#  treatment_arm mean_change
#  <chr>               <dbl>
# 1 Control            -0.21 
# 2 Low_dose           -1.49 
# 3 High_dose          -3.19 
```

Every value is a small number of grams, which is entirely plausible for a vole, so nothing here looks wrong on its face. But the result is the reverse of the truth. The treatment was administered to increase body mass, and the dataset was built so that mass rises with dose. The table reports that mass falls, and falls most at the highest dose.

This is the error class that neither your diagnostics nor an AI can catch for you. `glimpse()`, `distinct()`, and `count()` all report a clean dataset, because the data are clean; the fault is in the arithmetic, and the arithmetic produces a plausible number. The subtraction is the wrong way round. `mass_pre_g - mass_post_g` measures loss, when you wanted gain.

:::{.callout-note}

A chat window is more likely to catch this error because it uses patterns and conventions from training data to identify likely mistakes in logic or interpretation. Inline comments and code execution only check for syntax or runtime errors, not for whether the calculation matches common scientific practice.

:::

**1. Before you trust any result that runs cleanly, write down what a correct result should look like.** For this table: the sign should be positive, because the treatment adds mass, and the magnitude should increase from Control to High_dose. Only then compare. The mismatch between "should be positive and increasing" and "is negative and decreasing" is the diagnosis.

**2. You may give the code to AI for a second opinion, but it can only help if you supply the expected direction.** A prompt that says "this should show increasing mass gain with dose, but it shows decreasing values; what is wrong?" will get the sign error named. A prompt that gives only the code and the output will get a description of what the code does, which is not the same as catching that it does the wrong thing.


`r hide("Solution")`

```{r, eval = FALSE}

voles |>
  mutate(mass_change_g = mass_post_g - mass_pre_g) |>
  summarise(mean_change = mean(mass_change_g), .by = treatment_arm)

```

`mass_change_g` is post minus pre, not pre minus post. With the subtraction the right way round the means are positive and increase with dose, matching the design of the study. The original ran without complaint because subtraction is defined in both directions; the only signal that it was wrong was knowing what the answer should be.


`r unhide()`


## The debugging checklist

When code fails, the order matters. Inspect the data first, ask the AI second.

1. Read the error message in full. The function name and the phrase ("could not find function", "object not found", "non-numeric argument") usually localise the bug.

2. Inspect the relevant data. `glimpse()`, `distinct()` on factor columns, `summary()` on numeric columns, and `nrow()` after a filter are the four commands that catch most logic errors in seconds.

3. Verify variable names exactly, including capitalisation and underscores. Most "object not found" errors are typos.

4. If steps 1 to 3 do not localise the bug, paste the code, the error message, and the diagnostic output into Chat together. The combination is much more useful than the error alone.

5. After AI suggests a fix, write a one-sentence explanation of why the original failed. If you cannot write the sentence, you have not understood the fix.

6. Before trusting a result that runs without error, state in advance what it should look like: the sign, the rough magnitude, and the expected ordering. Compare the output against that statement. This is the one check AI cannot perform for you, because it requires knowing the biology.


## Check your understanding

**When R returns `Error in layer(): object 'X' not found` inside a ggplot2 call, the most common cause is:**

`r longmcq(c(
  "The data argument was supplied incorrectly",
  answer = "A column name was used outside aes() where ggplot2 expected a fixed value",
  "The ggplot2 package is not loaded",
  "The variable contains missing values"
))`


