# Part 4: Github Copilot Chat in Ask mode  {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```

Github Copilot Chat uses the same underlying AI model as inline completion, but for a different kind of help. Inline completion is best when you are actively writing code and want quick suggestions or continuations, which are usually included in standard Copilot usage. Chat is better for higher-level tasks such as planning an approach, understanding an error, or discussing design decisions, and these features are more likely to be usage-limited or *metered* because they require more computation. The two tools complement each other rather than replace one another.

*⏱ ~40 minutes* 


## What Chat sees, and what it does not

Here Githhub Copilot Chat differs fundamentally from a general chatbot such as ChatGPT, and the difference is easy to get wrong. ChatGPT sees *only the text you paste*; it has no connection to your files. Github Copilot is integrated with your editor and repository and can draw on context you did not paste. How much it draws on, and whether it can act on your machine can vary depending on set-up. Three matter for this workshop, and they are not equivalent.

- **Browser Chat on github.com**: open your repository on github.com and select the Copilot icon. It works in the context of that repository and can reach its files, but it has no connection to your local machine, runs no code, and cannot see your R session. This is the one that behaves the same for everyone.

- **Positron**: Positron also uses GitHub Copilot, signed in with your GitHub account, unless you configure a different provider such as an Anthropic API key. It is delivered through a Positron-modified build of the Copilot Chat extension, and the difference that matters is the context it adds. Positron hands Copilot context about your interactive data science work, such as your loaded data, plots, **and console history**, and in agent mode it can execute code in the console and view the output. So although the model is Copilot, in Positron it is fed your session, and the "cannot see your data" assumption does not hold the way it does in browser Chat.

- Your `copilot-instructions.md`, if present, is loaded automatically into every Copilot Chat request in that workspace. Exactly what is pulled in automatically has changed between versions and differs across surfaces, so the reliable habit is to set the context yourself with the references above rather than to assume (*see Part 5*).


For these exercises we should run them either in browser Chat, *or* in a session where the relevant file and environment are not already in context. In Positron, Copilot may already have your data frame in context, or be able to inspect it by running code, which removes the contrast entirely. Either is worth noticing. RStudio users should use browser Chat throughout. 

:::{.callout-note}
An assistant can describe your data confidently for two different reasons: 1) it has been given real information about your session, 2) or it is recalling something from training data. Only the first is reliable.

It has real information when one of these holds: you pasted the output (`glimpse()`, `summary()`, the printed result), or the data sits in a file it can read.

In Positron the IDE injects your session context (look for the Console session tab); or, in agent mode, it can run code and read the output.

It is guessing from training data when none of those apply, which is always the case in browser Chat on github.com. Tells: it substitutes a well-known dataset (mtcars, iris, penguins) you never mentioned, or asserts column names, factor levels, or values you never supplied.

Even when context is supplied, the model's claim about what it can see is unreliable, and the context may be structure rather than actual values. Verify any data-specific claim against your real object.

:::


**Browser Chat** (all IDEs, including RStudio): navigate to your repository on [github.com](github.com) and click the Copilot icon in the top right. This is the one that works for everyone.

**In-IDE Chat panel** (Positron and VS Code only): open the Copilot panel in the sidebar, and use the `#` references above to control what it sees. Use  quick actions as slash commands. Type `/` in the chat pane or in inline chat to see the list, pick one, and optionally add instructions before sending. The commands shipped with the Assistant include `/fix` to propose a correction, `/explain` to describe what code does, `/doc` to add documentation, and `/exportQuarto` to turn a chat into a Quarto document. They are designed especially for inline chat, where they act on the code you have selected. The Assistant can also trigger them from plain language: asking "can you convert this to Quarto?" invokes the export without your typing the command. 

RStudio users should use the browser Chat throughout this block. Positron and VS Code users may use either; the exercises work in both, subject to the context caveat above.


## Exercise 1: Reading a stack trace

This exercise demonstrates the dependency of Chat's answer quality on the context you supply.

**For Positron disable current session context to see the different responses you get**

**1. Open a fresh Chat session. Paste only the following error message and ask "What does this error mean?":**

```
Error in `mutate()`:
ℹ In argument: `mass_change = mass_post - mass_pre_g`.
Caused by error:
! object 'mass_post' not found
```

Read the answer. It will be generic and slightly tentative, because Chat is reasoning from the error alone.

**2. Now open a new Chat session and paste the same error along with the code that produced it and the output of `glimpse(voles)`:**

```
I am running this code:

  voles |>
    mutate(mass_change = mass_post - mass_pre_g)

and getting:

  Error in `mutate()`:
  ℹ In argument: `mass_change = mass_post - mass_pre_g`.
  Caused by error:
  ! object 'mass_post' not found

The data structure is:

  Rows: 72
  Columns: 6
  $ subject_id          <chr> "V001", "V002", "V003", ...
  $ sex                 <chr> "M", "F", "M", ...
  $ treatment_arm       <chr> "Control", "Control", ...
  $ mass_pre_g          <dbl> 28.3, 27.9, 30.1, ...
  $ mass_post_g         <dbl> 28.5, 28.0, 30.5, ...
  $ metabolic_rate_post <dbl> 13.21, 12.84, 13.55, ...

What is wrong?
```

Read this answer. It will name the specific bug (a missing `_g` suffix on `mass_post`) and propose the fix. Chat could not name it from the error alone because the column names are a property of the data loaded in your R session, which Chat cannot see; supplying the `glimpse()` output put them in front of it. A generic question gets a generic answer; a specific question gets a specific answer.


## How much of your data to show

The contents of your data can reach the model by three routes, and each is a disclosure:

- You paste it into the chat: an error, an output, a printed result.

- It reads a data file that sits in a repository or workspace it can see. Browser Chat on github.com works in the context of the repository you open it from, so a committed voles.csv is within reach; the VS Code extension can surface project files through workspace indexing.

- The IDE injects your session context. In Positron, the assistant is given context about your interactive data science work, such as your loaded data, plots, and console history, so a data frame already in memory can reach it without a paste. Positron

Describing your data to Chat is therefore a recurring task, but it is not the only way your data is exposed. Before you decide what to show, know what the surface can already reach: a data file committed to the repo, or a tibble loaded in a Positron session, may have been disclosed already.

When you do describe a data frame deliberately, three commands reveal increasing amounts:

- `names(voles)`, or equivalently `colnames(voles)`, lists the column names only, in their original order, and discloses nothing about the values. Prefer this to `ls(voles)`, which returns the same names but sorts them alphabetically and so loses the column order.

- `summary(voles)` adds the range, quartiles, and missing-value counts for numeric columns, and is less informative for character or factor columns. It reveals the shape of the data, not individual records.

- `glimpse(voles)` and `str(voles)` add the first few actual values of every column.

Use the least-disclosing call that still lets the AI help, and scale up only if the answer needs it. For the synthetic voles data this is moot, because no real subject is described. For your own data it is not. Human-participant data, and sensitive ecological data such as the locations of a protected species, may not be shared with an external tool at all under your data-management plan or ethics approval. names() is almost always safe; summary() and str() expose real values and may not be.

## Exercise 2: Refining an approach question

Run this in browser Chat, or with no analysis file referenced. If your script is open in the editor, Copilot may infer the design from it, and the first answer will be less generic than described below. The refinement in step 2 still helps in that case, because your scientific question and your design intent are not written in any file; they exist only in your head until you state them.

**1. Open a Chat session and ask, exactly as written:**

```
I have a dataset with pre and post measurements. How should I analyse it?
```

Read the answer. With no design and no scientific question supplied, it will be a tour of paired t-tests, mixed models, change scores, ANCOVA, and so on, because the question is too vague to permit a recommendation.

**2. Refine the question, with the dataset structure and the scientific question explicit:**

```
I have 72 voles, balanced across three treatment arms (Control, Low_dose,
High_dose) and two sexes (M, F). Each animal has body mass measured before
and after treatment, and a single post-treatment metabolic-rate reading.

My biological question is whether the high-dose treatment changes mass
relative to control, and whether the effect differs between sexes.

The dataset is small (n = 12 per cell). I will report effect sizes
alongside any inferential test. Given this design, what is a defensible
primary analysis, and what are the main threats to its validity?
```

Read the second answer. It will be substantially more useful, because the question now has enough information to permit a recommendation. The exercise is about question-craft: AI tools answer the question you actually asked, not the question you meant.


## Exercise 3: Verify the AI did not change behaviour

Asking AI to comment or to refactor code carries a specific risk: that it changes what the code does while claiming only to have tidied it. Reading the diff catches changed code. It does not always catch changed output, because a small alteration is easy to miss by eye. There is a cheap check for the output.

**1. Compute and keep a reference result:**

```{r}
#| eval: false
original <- voles |>
  mutate(mass_change_g = mass_post_g - mass_pre_g) |>
  summarise(mean_change = mean(mass_change_g), .by = treatment_arm)
```

**2. Ask Chat to refactor or comment the pipeline.** Use a prompt such as: "Add explanatory comments to this pipeline and make it more readable, but do not change what it returns." Run the version it gives you, assigning the result to `refactored`.

**3. Compare the two results:**

```{r}
#| eval: false
all.equal(original, refactored)
```

`all.equal()` returns `TRUE` when the two objects are identical to within numerical tolerance. Anything else is a report of the differences, and a signal that the refactor changed the output. For tibbles, `waldo::compare(original, refactored)` gives a more readable account of where they diverge.

The point is not that AI always changes your code; often it does not. The point is that you cannot tell by trust, and the check costs one line. Whenever you ask AI to transform code that already works, verify that the output survived the transformation.


## When to use Chat versus inline

A short decision rule covers most cases:

| Task | Inline | Chat |
|---|---|---|
| Writing the next line of code | Yes | No |
| One-line typo or syntax fix | Yes | No |
| Reading an unfamiliar error | Marginal | Yes |
| Choosing between two approaches | No | Yes |
| Planning the structure of a script | No | Yes |
| Refactoring more than a few lines | No | Yes |

The rule is not strict; both tools have considerable overlap. The principle is: inline for typing, Chat for thinking.


## Check your understanding

**Which statement best describes the context available to in-IDE Copilot Chat, compared with a general chatbot such as ChatGPT?**

`r longmcq(c(
  "Like ChatGPT, it sees only the text you paste into the prompt",
  answer = "It is integrated with your editor and project, so it can draw on open files, a workspace index, and files you reference, but it cannot see the objects loaded in your R session",
  "It has full automatic access to every file in the repository and to your running R session",
  "It can execute your R code to read the data directly"
))`


# Part 5: Custom instructions and skills

The biggest single improvement you can make to your AI workflow is to give Copilot persistent context about how you write code, so you do not have to re-state your conventions in every comment and every Chat session. This block introduces the configuration mechanisms that allow you to do this, and gets you to write the most important one.

*⏱ ~55 minutes*


## Before you start: which assistant are you actually using?

:::{.callout-warning collapse = "true"}

If you are working in Positron, read this orientation first, because Positron involves two AI tools whose names are almost identical and which handle the files in this part differently. Confusing the two is the single most common reason a configuration file appears to be ignored.

**Positron Assistant** came first. It is the AI feature built into Positron, a bring-your-own-key client that connects to a provider you choose (Anthropic, OpenAI, GitHub Copilot, and others). For the purposes of this part, its useful property is that it works well with GitHub Copilot's instruction files: the .github/ conventions described here are read through the Copilot path. 

Its limitation is skills: it does not reliably load them. Whether a given request even reaches the Copilot path that would honour a skill is not fully determined by the provider dropdown, and the routing internals are undocumented, so a skill that fires for one user can be silently ignored for another with an apparently identical setup. Treat instruction files as dependable in Positron Assistant and skills as not.

**Posit Assistant** is the newer tool, currently in preview and under active development as the successor that unifies Positron Assistant and Databot into one agent running across both Positron and RStudio. It inverts the trade-off. It does have a native skills hook, but it searches its own location, under .positai/skills/ or .agents/skills/, configurable via skills.paths in ~/.positai/settings.json. 

It does not read Copilot's .github/ files, so the copilot-instructions.md, scoped instructions, and .github/skills in this part will not be picked up by Posit Assistant unless you place or point them at one of its own directories.

The blunt rule of thumb: author for Copilot in .github/, and rely on instruction files rather than skills if you are in Positron Assistant. If you also use Posit Assistant, treat its skill discovery as a separate, parallel setup rather than assuming one folder serves both. Where this part depends on behaviour that differs between the two tools it says so, and where the behaviour is undocumented it says that too, because both are moving targets and guessing wastes your time.

:::

## Providing context to models

Github copilot (or any LLM) provide several ways to suppl context, distinguished by how they load

**Workspace-level custom instructions** live in `.github/copilot-instructions.md`. Github Copilot loads this file automatically for **every request made in the repository**, whether in Chat, code review, or the cloud agent. Use it for things that are always true of your project: which packages you use, your naming conventions, your file structure, your statistical conventions. The file should be short and broad in scope; treat two pages as a ceiling, not a target.

**Scoped instructions** live in `.github/instructions/*.instructions.md`, with a YAML front-matter `applyTo` glob that restricts the file to particular paths (for example `applyTo: "**/*.qmd"` to apply only to Quarto documents). Use them for rules that apply to one kind of file but not the rest. They load automatically when Copilot works on a matching file, with one important limitation: on GitHub.com they are currently honoured only by the cloud agent and by code review, not by Chat or inline completion. If your work is mainly in the editor, confirm your surface before relying on this mechanism.

**Agent Skills** live in `.github/skills/<skill-name>/SKILL.md`, with a YAML front-matter `name` and `description`. Unlike the first two mechanisms, skills are not always loaded; they are invoked. Either you type `/skill-name` in Chat or agent mode, or Copilot infers from the description that the skill is relevant to the task. Skills suit procedures: how to make a publication figure, how to write a function, how to review code. Because they load only when needed, they can carry more material than an always-on file, though it is still worth putting the most important rules first.

A fourth, related mechanism is custom prompt files (`.github/prompts/*.prompt.md`), invoked explicitly as slash commands. These are outside the scope of this workshop.

The mechanisms answer different questions. Custom instructions answer "how should code in this project be written?" Scoped instructions answer "how should this kind of file be written?" Skills answer "how should this specific task be executed?" The visualisation skill demonstrated shortly is an example of the third.


## How to write effective instructions

A `copilot-instructions.md` file is loaded into every Chat interaction in the repository and is read by the cloud agent before it starts work, so its length is a recurring cost and brevity is a virtue. A handful of accurate lines outperforms a long file.

A few principles carry most of the value:

- Write each instruction as a single, simple statement. If you have several points, use several lines rather than one dense paragraph; the model parses discrete statements more reliably than prose.

- Describe the conventions you actually follow, not the ones you aspire to. An instruction your own code routinely breaks teaches the model a rule the surrounding code contradicts, which makes its suggestions less consistent.

- Be concrete and positive. "Use the native pipe `|>`" and "snake_case for all object names" are actionable; "write clean, idiomatic code" is not.

- Avoid contradiction and overlap. When a workspace instruction and a scoped instruction conflict, Copilot's choice between them is not deterministic, so do not rely on one quietly overriding the other.

Some other set-up and principles can be found at the
[Github documentation on onboarding an instruction file](https://docs.github.com/en/copilot/how-tos/copilot-on-github/customize-copilot/add-custom-instructions/add-repository-instructions)

`r hide("Example")`

```
# Copilot Instructions: R Analytical Repository

This repository contains analytical R code for computationally reproducible
research. Most items below are written for you to follow as you generate code.
A short Verification section lists the few things that are also checked in CI,
because a silent failure there is costly.

## Canary
Start any response that writes or edits R code with this as the first line of
the first code block:
`# copilot-instructions: active v4`

## Priorities
When two instructions conflict and no test settles it, prefer in this order:
correctness, reproducibility, transparency, interpretability, maintainability.
Note the trade-off in the pull request description.

## Agent behaviour
Never wait for human input in committed code. Do not use `readline()`,
`menu()`, `file.choose()`, `browser()`, or `View()`. When an input is missing,
add a `# DEFAULT:` comment naming the assumption and the value you chose, then
carry on.

## Naming and style
- Use the native pipe `|>`, not `%>%`.
- Use snake_case for all objects and functions.
- Assign with `<-`, not `=`.
- Write `TRUE` and `FALSE`, never `T` and `F`.
- Suffix dataframe and tibble objects with `_df`; raw import objects use `_raw`
  instead and are never reassigned.
- Suffix fitted models with `_model`.
- Suffix ggplot objects with `_plot`.
- Namespace-qualify calls (`pkg::fun()`) inside `R/functions/`. In analysis
  scripts, load packages with `library()` at the top and use bare names below.
- Avoid hidden global state and partial argument matching.
- Abstract logic into a function once it appears in three or more places.

## Reproducibility
- Never use `setwd()` or `rm(list = ls())`.
- Build every path with `here::here()`; relative paths only.
- Assume each script runs from a fresh R session.
- Manage dependencies with `renv`.
- Set a seed before any stochastic operation.
- Save `sessioninfo::session_info()` to the output directory at the end of each
  top-level script.

## Dependencies (renv)
- Manage all dependencies with `renv`; the project library is the only source
  of truth.
- Never call `install.packages()`. Add a package with `renv::install()`, then
  `renv::snapshot(prompt = FALSE)`.
- When `renv.lock` already lists the packages you need, restore rather than
  install: `renv::restore(prompt = FALSE)`.
- Never edit `renv.lock` by hand.
- Set `options(renv.config.auto.snapshot = FALSE)` so the lockfile is never
  written as a side effect; snapshot only deliberately.
- Use `renv::snapshot(type = "all")` so the lockfile is reproducible across
  runs rather than depending on usage detection.


## Reading and writing data
- Always use functions from readr`readr::read_*(col_types = ...)` not `read.`.
- Keep raw data files immutable; never overwrite anything in the raw data
  directory.
- Separate raw data, derived data, scripts, outputs, and documentation.
- Write outputs only to dedicated output directories, never outside the
  repository.
- Assign cleaned and derived results to new objects; do not overwrite the raw
  import object.

## Data integrity
- Never drop missing data silently. Filter with a stated reason, or use
  `tidyr::drop_na()` on named columns.
- Prefer `purrr::map_*()` and `vapply()` over `sapply()`.

## Statistical practice
- Distinguish exploratory from confirmatory analyses, in prose or chunk labels.
- Validate model assumptions with `performance::check_model()` for regression
  models and record the inspection.

## Figures
For any figure, quick or polished:
- Use tidyverse style, the native pipe, and British spelling.
- Build paths with `here::here()`.
- No pie charts, dual y-axes, 3D effects, or dynamite plots (bar plus error
  bar) when raw data exist.
- Never use `theme_grey()`; default to `theme_minimal()` or `theme_classic()`.
- For categorical groups, add `shape` or `linetype` alongside colour.
- Use `linewidth`, not `size`, for line and density layers.
- For figures meant for an audience, follow the pub-figures skill in full.

## Code chunks
- Keep chunks short: one logical step each (import, clean, transform, model,
  diagnose, or visualise), and no more than about 40 lines.
- Comment the rationale for the step, not the mechanics of each line.
- Include at least one comment per chunk.

## Quarto
For `.qmd` files: render from a clean session, give chunks informative labels,
and avoid hidden dependencies between chunks.

## Verification (also checked in CI)
These are enforced because a silent miss is costly:
- No `setwd(`, `%>%`, `install.packages(`, or absolute paths in committed code.
- `renv::status()` reports a synchronised library.
- Each script and `.qmd` runs or renders in a fresh `--vanilla` session without
  error.
- The raw data directory is unchanged by any run.
- Outputs resolve inside the repository.

## Specialist instructions and skills
Specialised instruction files hold domain standards and repository policy;
skills hold procedural workflows. When one applies, follow it alongside these
instructions. Skill selection follows each skill's own description.

```

`r unhide()`


## Activity Build your a copilot-instructions.md file

:::{.task}
::::{.task-header}
Build your a copilot-instructions.md file
::::
::::{.task-container}

In the workshop template repository (which you forked at the start of the day), you will create `.github/copilot-instructions.md`. Rather than start from a blank file, generate a skeleton and then refine it. The refining is the real exercise.

**Step 1: Before and After.** 

This is a controlled before-and-after: the same prompt, run twice, with the instructions file the only thing that changes between the runs.

Run the prompt:

```
Read in data/voles_metabolism.csv and make a quick exploratory plot of mas_post_g
by treatment arm.
```
> You can do this in the Positron IDE or by heading to the repository on Github

**Step 2: Draft a skeleton.** 

*If you are in Positron*: open the assistant chat and prompt, "Create .github/copilot-instructions.md by reading the current workspace." The assistant reads the repository and drafts a file covering the structure, languages, and packages it detects.

*If you are in RStudio*: there is no Copilot Chat to generate the file for you, so create .github/copilot-instructions.md by hand. Either take the example from above, or, if it is easier, skip ahead, push an empty file, and use GitHub.com's web Chat to generate a draft once it is on the remote. Hand-authoring is entirely fine here; the file is short.

**Step 3: Trim it hard.** Generated instructions run long and generic. Cut the draft down to what is always true of the project and broadly useful, treating two pages as a ceiling, not a target. Delete anything the model padded or guessed at.

**Step 4: Make it yours.** The generator reads the workshop template, not your own work, and it cannot read your judgement. Edit the file so that it specifies, at minimum:

**Step 5: Commit and push.** Commit .github/copilot-instructions.md and push to your fork on GitHub. This step is what makes the file available where it is read most reliably, and it is the point at which Positron and RStudio users converge: from here on, everyone works the same way.

**Step 6: Verify on GitHub.com**. On your fork's web page, open Copilot Chat (the web surface honours copilot-instructions.md for both Chat and the cloud agent, regardless of which IDE you used) and ask:

The deliverable is the committed `copilot-instructions.md` in your forked repository. After committing, open a fresh chat session and ask:

```
Read in data/voles_metabolism.csv and make a quick exploratory plot of mas_post_g
by treatment arm.
```

::::
:::


The file loads automatically in a workspace chat, so the explicit `#` reference is belt-and-braces here: it confirms the file is being found. If the sketch follows your conventions, in particular the one you wrote for item 4, the instructions are working. If it ignores them, check that the file is at `.github/copilot-instructions.md` and that instruction files are enabled in your settings.

Compare the answer with what someone else gets from their own instructions file. The differences are the personality of your standards.


## Activity: Write a scoped instruction for Quarto documents

A scoped instruction is a file of rules that applies only to certain files in your project. It lives in .github/instructions/ and carries an applyTo glob in its front matter that names the paths it covers, for example applyTo: "**/*.qmd" to apply only to Quarto documents. When Copilot works on a file that matches the glob, the rules load automatically; when it works on any other file, they do not. 

:::{callout-note}

## When to use a scoped file

The test for whether something belongs in a scoped file rather than the
always-on copilot-instructions.md is simple: would the rule be wrong, not
merely irrelevant, in the files the glob excludes? If a rule is true
everywhere (use the native pipe, use snake_case), it belongs in the always-on
file; putting it in a scoped file gains nothing and adds the scoped mechanism's
reliability problems. A rule earns a scoped file only when it is a property of
one file type and would mislead the model elsewhere.

:::

Quarto `.qmd` files are a clear case for an R user, because they carry concerns a plain `.R` script does not have at all: chunk options, how code is echoed into the rendered document, figure captions and dimensions, and a working-directory trap specific to rendering. An instruction to set echo: false is correct inside a `.qmd` and meaningless inside `R/02_analysis.R`. 


:::{.task}
::::{.task-header}

Create a scoped instruction file
::::
::::{.task-container}

Create `.github/instructions/quarto.instructions.md` with the following
content:

```
---
applyTo: "**/*.qmd"
---

# Quarto document conventions

These rules apply to Quarto source files only. They concern how code is
presented and rendered in the document, not the analysis logic itself.

## Chunk options
- Set chunk options with the hashpipe (`#|`) inside the chunk, not in the
  `{r}` header. For example `#| echo: false`, not `{r, echo=FALSE}`.
- Give every code chunk a label (`#| label: fig-mass-by-arm`) so cross-
  references and cached results are stable.
- Every chunk that produces a figure must set `#| fig-cap:` with a caption;
  do not emit an uncaptioned figure into a rendered document.

## Echo and output for a written report
- Default to `#| echo: false` so the rendered document shows results, not
  source, unless the document is explicitly a code walkthrough.
- Use `#| warning: false` and `#| message: false` on chunks that load
  packages or fit models, so loading noise does not appear in the output.

## Figures
- Label figure chunks with a `fig-` prefix and reference them in prose with
  `@fig-...`, not by a hardcoded number.
- Set figure dimensions explicitly with `#| fig-width:` and `#| fig-height:`
  rather than relying on defaults.

## Reproducibility
- Resolve paths with `here::here()` inside chunks; never assume the working
  directory is the document's folder, because Quarto and the R session may
  differ on this.

```

**Test it.**

With the scoped file committed, open a .qmd file and ask Copilot to add a code chunk that reads the voles data and plots mass_post_g by treatment_arm. A correct response uses hashpipe chunk options, labels the chunk with a fig- prefix, sets a fig-cap:, and resolves the path with here::here(). Then open a plain .R script and ask for the same plot: the hashpipe and caption rules should not appear, because the scoped file does not apply there. That contrast, the same request producing different output in the two file types, is the demonstration.

::::
:::

## Agent skills

A skill is a packaged procedure the assistant loads when a task calls for it, rather than a set of rules attached to a file type or applied to every request. For Copilot it lives in `.github/skills/<skill-name>/SKILL.md` and carries a name and a description in its front matter; the description is what the assistant reads to decide the skill is relevant, or you invoke it directly by name. Because a skill loads only when needed, it can be longer and carry supporting material, templates, tables, reference files, that an always-on instruction file could not afford to include on every request.

A caution before the demonstration, following from the orientation at the start of this part. Skills are the one mechanism here that does not transfer cleanly across the two Positron tools. Positron Assistant **does not reliably load skills**, so the `pub-figures` skill below should not be expected to fire their.

Posit Assistant **does support skills, but it does not read Copilot's .github/skills location; its skills must live under .positai/skills/** (or another path registered in skills.paths). So a skill authored once for Copilot is not automatically available to either Positron tool: Positron Assistant will not load it, and Posit Assistant will not find it unless you also place it under `.positai/skills/`. 

The skill being demonstrated is `pub-figures`, which governs the production of every figure in the project that a human will read. The skill encodes a chart-selection table, a list of unconditional prohibitions (no pie charts, no dual y-axes, no dynamite plots when raw data are available), an accessible colour scheme, audience-specific styling, a mandatory colour-vision-deficiency check, and a checklist.

:::{.callout-warning}

Run this exercise where the skill will actually load. With Github Copilot, that is in a `.github/skills/pub-figures` directory and executed on GitHub.com (the cloud agent), not the editor. If you want to reproduce it inside your IDE instead, that path is Posit Assistant with the skill placed under `.positai/skills/pub-figures/`; Positron Assistant will not load it, and the Copilot editor will not either. The steps below are written for the GitHub.com surface.

:::

**1. Default output.** In a fresh Chat session (with no skill loaded), ask:

```
Produce a ggplot2 boxplot of mass_post_g by treatment_arm for the voles
dataset.
```

The result is a default-styled boxplot: grey background, ggplot2 default colours, no consideration of accessibility, no annotation of sample sizes. It runs and answers the literal question, and that is all.

**2. With the skill.** Open a new Chat session, invoke the skill (`/pub-figures`), and ask the same question. The skill requires the agent to elicit the audience before producing code. The agent will ask whether the figure is for a journal/lab report, a poster, a talk, or public engagement. Answer "journal/lab report" and confirm the plan it produces.

```
your-repo/
└── .github/
    └── skills/
        └── pub-figures/
            ├── SKILL.md
            ├── raincloud-template.R
            ├── emmeans-bridge.R
            ├── save-figure.R
            ├── theme-project.R


```

The output is materially different. A raincloud (`ggdist::stat_halfeye` plus a boxplot plus jittered points) should replace the bare boxplot, because the skill specifies raincloud as the default for discrete-continuous comparisons where raw data are available. The palette is colour bline friendly. Axis labels carry units. The pre-flight checklist is reported in the chat.

**3. Iteration.** Ask the agent to add pairwise statistical annotation. The skill prohibits hallucinating p-values, so the agent should refuse unless a fitted model object is available. Provide a fitted `lm()` and the `emmeans` contrasts; the agent will use `ggpubr::stat_pvalue_manual()` to draw brackets and labels from the model-derived data. The four-element figure legend appears at the bottom of the output.


## How to write skills

You will receive these skills rather than author them in this workshop, but it is worth seeing how a skill is built, because the same structure applies when you write your own, because it explains why the demonstration above behaved as it did, and because it explains why the `characterisation-test` skill in Part 6 is able to constrain the cloud agent so tightly.

A skill is a folder, `.github/skills/<skill-name>/SKILL.md`, with a YAML front-matter block and a Markdown body. Two parts decide whether it works:

- The `description` in the front matter is the trigger. It is the text the agent reads to decide whether the skill is relevant to a request, so it must name the condition under which the skill applies, specifically. "Used when producing any ggplot2 figure a human will read" loads when it should; "for plots" loads erratically. A skill invoked by name with `/skill-name` does not depend on the description, but a skill you want loaded automatically does.

- The body is a procedure: when to use the skill, the ordered steps, any hard prohibitions, and a final checklist. Because a skill is read only when invoked, it can be long without slowing ordinary interactions, which is why the `pub-figures` skill can carry a chart-selection table, a prohibition list, and an accessibility check at no cost to a routine question. Supporting files such as templates or reference tables can sit in the skill folder alongside `SKILL.md`.


## Check your understanding

**An Agent Skill differs from a workspace-level custom instructions file in that:**

`r longmcq(c(
  "Skills only work in Positron and VS Code, custom instructions work in all IDEs",
  answer = "Custom instructions are loaded automatically for every Chat interaction, whereas skills are invoked by name or loaded only when judged relevant",
  "Skills are paid features and custom instructions are free",
  "Custom instructions can include code but skills cannot"
))`


# Part 6: Functionalise a script step and prove the result is unchanged

Suppose you have a computation that already works and you want the cloud agent to
tidy it into a reusable function. The agent can do that in minutes. The difficulty
is verification: a refactor that "improves" code can quietly change its output, so
how do you confidently check that the function still does what the script did? The
answer this part teaches is to freeze the original result before the agent touches
anything. You run the analysis once, save its output to a reference file (an
`.rds`), and then require the agent's function to reproduce that reference exactly.
You hold the reference; the agent does the refactor and proves that it matches. The
reference is the check, and because you captured it, the check cannot be quietly
redefined by the agent.

*⏱ ~30 minutes*


## Recap: in-IDE Agent versus cloud agent

The distinction from Part 1 matters here. In-IDE Agent mode, available in Positron and VS Code, edits files in your editor in real time while you watch. The cloud agent runs on GitHub's servers, takes an issue as input, creates a branch and a pull request without you watching, and notifies you when it has finished. This block uses the cloud agent throughout, because it is platform-agnostic and exposes the full software-engineering workflow the workshop is building towards. RStudio users are therefore on equal footing in this part.

## The repository layout
Before any code, here is the full layout for this part. Each file is marked by who produces it: [provided] ships in the template, [you] is the one file you generate in Step 0, and [agent] is what the cloud agent creates or rewires. The three-way status is the conceptual spine of the whole exercise, so refer back to it as you go.

```
repo/
├── .github/
│   ├── workflows/
│   │   ├── copilot-setup-steps.yml   [provided]  builds the agent's R environment
│   │   └── r-checks.yml              [provided]  independent CI check on push and PR
│   └── skills/
│       └── characterisation-test/
│           └── SKILL.md              [provided]  tells the agent how to behave
├── data/
│   └── voles_metabolism.csv          [provided]  the dataset
├── R/
│   ├── 01_load_data.R                [provided]  loads the CSV, runs glimpse()
│   ├── 02_analysis.R                 [provided → rewired by agent]
│   └── functions/
│       └── summary_functions.R       [agent]     the extracted function
├── tests/
│   ├── fixtures/
│   │   ├── make_reference.R          [provided]  captures the reference
│   │   └── metabolic_summary.rds     [you]       generated in Step 0, then frozen
│   └── test_summary_functions.R      [agent]     compares function against reference
└── renv.lock                         [provided]  pins tidyverse, here, testthat, renv
```

The single most important property of this layout: the `.rds` under tests/fixtures/ is generated by you and by nobody else. Everything else in the trust argument follows from that.

## What the template already provides (read, do not write)

Recall from the GitHub session that GitHub Actions runs automated workflows in response to repository events such as a push or a pull request. Two such workflows sit in .github/workflows/, both placed there for you.

- `r-checks.yml` — the independent check

This runs the tests on pushes to main and on every pull request. It is the check that neither you nor the agent can fake, because it runs on GitHub's infrastructure from a clean checkout. A green tick here is the signal you wait for at the end.

`copilot-setup-steps.yml` — the agent's environment

This is the file the original tutorial glossed, and it is worth your attention, because it is what makes the agent's environment resemble your IDE. The cloud agent runs the tests in the environment this file builds. If that environment does not restore renv.lock, the agent's tests will fail (or, worse, pass) for reasons that have nothing to do with the refactor, and the resemblance to your local setup is lost.

Two requirements are load-bearing:

- The job must be named exactly copilot-setup-steps. GitHub looks for that job name specifically; rename it and the agent silently falls back to a bare environment without your packages.
The job must restore the lockfile, so the agent computes against the same pinned package versions that r-checks will later use.

- A conventional form, using the r-lib/actions setup actions, looks like the following. Check this against your actual template file rather than copying it blindly; this reproduces the standard pattern, not necessarily your exact configuration.


```
name: Copilot setup steps

on: workflow_dispatch

jobs:

  copilot-setup-steps:
    runs-on: ubuntu-latest
    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: "4.5.2"      # match the version in renv.lock
          use-public-rspm: true
          
      - name: Set up renv
        uses: r-lib/actions/setup-renv@v2
        env:
          RENV_CONFIG_PAK_ENABLED: "true"

      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2

```

`setup-renv` reads `renv.lock` and installs exactly the pinned versions, so the agent's tidyverse, here and testthat match yours. The practical consequence: when you reach Step 4, the agent's own test run and the independent r-checks run should agree. If they disagree, suspect the environment before suspecting the code.

### The two scripts and the skill

The repository holds two short scripts. `R/01_load_data.R` loads the data and runs `glimpse()`; `R/02_analysis.R` sources it and computes the summary you will functionalise:

```{r}
#| eval: false
# R/02_analysis.R — the original, trusted version
library(tidyverse)
library(here)

# 01_load_data.R reads the CSV into `voles` and runs glimpse()
source(here::here("R", "01_load_data.R"))

metabolic_summary <- voles |>
  group_by(treatment_arm, sex) |>
  summarise(
    mean_rate = mean(metabolic_rate_post),
    sd_rate   = sd(metabolic_rate_post),
    n         = n(),
    .groups   = "drop"
  )
```

And a skill telling the agent how to behave for this kind of job:

```markdown
# .github/skills/characterisation-test/SKILL.md
---
name: characterisation-test
description: >-
  Use for background (cloud agent) tasks that move a computation from a script into a
  function in R/functions/ and verify it reproduces a saved reference. The reference is
  captured by the investigator, never by the agent.
---

# Functionalise and characterise

## What you must not do
Do not create, regenerate, or modify the reference output, and do not touch any file under
tests/fixtures/. If the fixture is missing, stop and ask the investigator to capture it. A
reference you make or alter from your own code proves nothing.

## Procedure
1. Write the computation as a function in R/functions/, with a roxygen2 block, argument
   checks, and one @examples entry. Reproduce the original output exactly, including column
   and row order.
2. Add a test that loads the fixture, runs the function, and compares with expect_equal().
3. Update the originating script (R/02_analysis.R) to call the new function in place of the
   inline computation, so the script and the function do not drift apart. Do not re-run any
   reference-capture script and do not touch tests/fixtures/.

## Verification
Run the test in your own environment first. Then push and wait for r-checks. If it fails,
read the log, fix the specific cause, and push again. Done means r-checks is green.

## If a failure persists after three attempts
Stop and comment on the pull request explaining what you tried and the remaining error.
Do not make speculative changes.
```

### Before you start: one repository setting

This repository holds no secrets, so you can turn on the setting that lets the Copilot coding agent's workflows run without manual approval, which lets you watch the checks run live. By default GitHub blocks them, treating the agent as an outside contributor, and you would otherwise sit at an "Approve and run" gate without understanding why.

1. In the repository, go to Settings → Actions → General.

2. Under Fork pull request workflows, allow GitHub Actions to run on pull requests from the agent.

3. Save.

:::{.callout-warning}

This is a convenience setting and you should be actively Approving runs in a standard workflow

:::


## Step 0 — You capture the reference (this stays with you)

Before the agent is involved at all, run `R/02_analysis.R` yourself and confirm
`metabolic_summary` looks as it should. This is the output you want to preserve. Then
run the capture script below, once, and commit the resulting `.rds`. This frozen
output is the trust anchor, and the agent never touches it. Run the capture from
within the project, so `renv` is active and the reference reflects the same locked
package versions that `r-checks` will restore. Capture it now, before the refactor:
afterwards `02_analysis.R` computes the summary through the very function under test,
so re-capturing from it would compare the function against itself. For this exercise
you capture once and leave the reference frozen.

```{r}
#| eval: false
# tests/fixtures/make_reference.R
# Capture the trusted reference once, from within the project so renv is active,
# so the reference reflects the locked package versions r-checks restores.
# Run this BEFORE the agent refactors 02_analysis.R, and capture only once: after
# the refactor that script computes the summary via the function under test, so a
# later capture would compare the function against itself.

if (!requireNamespace("renv", quietly = TRUE) || is.null(renv::project())) {
  stop(
    "Run this from within the project, with renv active, so the reference ",
    "matches the locked environment used by r-checks.",
    call. = FALSE
  )
}

source(here::here("R", "02_analysis.R"))
saveRDS(
  metabolic_summary,
  here::here("tests", "fixtures", "metabolic_summary.rds")
)
```

## Step 1 — Read the skill

Open `.github/skills/characterisation-test/SKILL.md` and find the one line that matters: the agent must not create, regenerate or modify the reference, and must not touch anything under `tests/fixtures/`. That single instruction is what stops the verification becoming circular, and it is your first aim in miniature, a few words of instruction setting a boundary the agent will respect.

## Step 2 — Write a thin issue and hand it over

```text
Title: Functionalise the metabolic-rate summary and verify it is unchanged

R/02_analysis.R computes metabolic_summary. Move that computation into a documented
function in R/functions/, have the script call it, and add a test that checks the function
reproduces the saved reference in tests/fixtures/metabolic_summary.rds.

Use the characterisation-test skill and follow the project conventions.
Push, make r-checks pass, and tell me when it is ready to review.
```

Notice how little you wrote. The return shape, the roxygen, the naming and the
assertion style are absent, because the skill and your standing instructions
already hold them. That is the lesson: short instructions work because the
architecture carries the standards.

## Step 3 — Watch it work, and probably stumble

The agent moves the computation into a function, writes the test, runs it in its
own environment, and pushes. The likely first failure is instructive. A tidy
refactor often switches `group_by()` for the modern `.by`, and the two do not
order their output the same way: `group_by()` sorts the grouping keys, whereas
`.by` returns groups in order of first appearance. The numbers are right, but
the rows are arranged differently, so `expect_equal()` reports a mismatch. Watch
the agent read that, add an `arrange()` to match the original order, and go
green. If it passes first time, walk through the prepared failed-then-fixed pull
request instead, so the recover-from-failure loop is seen either way. The point
to land is that improving code during a refactor can quietly change the output,
and the characterisation test is what catches it.

## Step 4 — Review the green pull request

```{r}
#| eval: false
# R/functions/summary_functions.R
#' Summarise post-treatment metabolic rate by treatment arm and sex
#'
#' @param data A data frame containing the columns treatment_arm, sex and
#'   metabolic_rate_post.
#' @return A tibble with one row per treatment_arm and sex, giving the mean,
#'   standard deviation and n.
#' @examples
#' demo <- tibble::tibble(
#'   treatment_arm = c("Control", "Control", "Low_dose", "Low_dose"),
#'   sex = c("F", "F", "M", "M"),
#'   metabolic_rate_post = c(1.2, 1.3, 1.6, 1.7)
#' )
#' summarise_metabolic_rate(demo)
summarise_metabolic_rate <- function(data) {
  stopifnot(is.data.frame(data))
  required <- c("treatment_arm", "sex", "metabolic_rate_post")
  missing_cols <- setdiff(required, names(data))
  if (length(missing_cols) > 0) {
    stop(
      "`data` is missing required column(s): ",
      paste(missing_cols, collapse = ", "),
      call. = FALSE
    )
  }

  data |>
    dplyr::summarise(
      mean_rate = mean(metabolic_rate_post),
      sd_rate   = sd(metabolic_rate_post),
      n         = dplyr::n(),
      .by = c(treatment_arm, sex)
    ) |>
    dplyr::arrange(treatment_arm, sex)
}
```

The tests: 

```{r}
#| eval: false
# tests/test_summary_functions.R
library(testthat)
source(here::here("R", "functions", "summary_functions.R"))

test_that("summarise_metabolic_rate reproduces the original summary", {
  reference <- readRDS(here::here("tests", "fixtures", "metabolic_summary.rds"))
  voles <- readr::read_csv(
    here::here("data", "voles_metabolism.csv"),
    show_col_types = FALSE
  )
  expect_equal(summarise_metabolic_rate(voles), reference, tolerance = 1e-8)
})
```

And `R/02_analysis.R`, now a thin caller rather than the inline pipeline:

```{r}
#| eval: false
# R/02_analysis.R — after the refactor
library(tidyverse)
library(here)

source(here::here("R", "01_load_data.R"))
source(here::here("R", "functions", "summary_functions.R"))

metabolic_summary <- summarise_metabolic_rate(voles)
```

Read four things. 

1. The green tick, in the Checks tab or at the bottom of the Conversation tab. Read it as evidence that the tests pass reproducibly on locked infrastructure. It is not evidence that the test is the right test; that is what the remaining items establish.

2. The test body, in the Files changed tab. Confirm it reads tests/fixtures/metabolic_summary.rds (your reference) and compares against it, rather than building its own reference inside the test.

3. R/02_analysis.R, also in Files changed. Confirm it now calls summarise_metabolic_rate() and no longer carries the old inline pipeline, so the function has genuinely replaced the code it was extracted from.

4. The list of changed files, at the left of the Files changed tab. Confirm tests/fixtures/metabolic_summary.rds does not appear anywhere in it. This is the most important item. Because the .rds is binary, git cannot show you what changed inside it, so the only safeguard is that it must not appear in the diff at all. If it appears, the agent has touched your trust anchor and the verification is circular, whatever the green tick says. In that case, reject the pull request and check the agent's commits against the skill's prohibition.



## Debrief — when is this worth it?

Use a CI agent task when the work is tedious, has a clear pass-or-fail check, and
you are content to review afterwards rather than steer live. Functionalising a
script step under a characterisation test fits all three. Deciding what the
script should compute in the first place, and judging whether the result is
scientifically sensible, do not, and stay with you.

## Check your understanding

**A green `r-checks` tick on the pull request tells you that:**

`r longmcq(c(
  "the function is correct and the result is scientifically sensible",
  answer = "the tests pass reproducibly on clean, locked infrastructure, but not that the test compares against an honest reference",
  "nothing useful, because the agent controls the outcome",
  "both that the result is reproducible and that the verification is non-circular"
))`

**True or false: the reference fixture is captured by the investigator rather than the agent because a reference generated from the agent's own output would make the verification circular.** `r torf(TRUE)`



:::{.task}
::::{.task-header}
Automated code review
::::
::::{.task-container}

Discuss what parts of code review you would hand over to an Agent.

What would the skill look like? 



::::
:::




## What comes next

This workshop is the foundation for a separate strand of work on Agent Skills. Three general-purpose skills are in development: tidyverse-style code optimisation, the publication-figure skill you saw demonstrated, and a code-review skill that applies a structured checklist to your work before submission. These will be released with documentation and worked examples. The `characterisation-test` skill you used with the cloud agent in Part 6 is a narrower, task-specific example of the same architecture, and it ships in the workshop template today. The architecture is intended to be modular; you can add your own skills as you identify recurring procedures in your work.


## Self-study pointers

For the underlying tools and conventions:

- The GitHub Copilot documentation is the canonical source for current behaviour and capabilities, and it changes faster than written materials can keep up. The model selector, the in-IDE Chat panel, and the cloud-agent assignment feature have all changed materially within the last twelve months.

- The TADA guidelines for analytical code sharing (Ivimey-Cook et al., 2025, EcoEvoRxiv preprint) cover the reproducibility scaffolding the skills assume.

- The tidyverse style guide (style.tidyverse.org) is the convention base layer; most of what you specify in your `copilot-instructions.md` will sit on top of it.



## Final note

AI tools accelerate code production. They do not accelerate understanding. The two diverge unless you actively close the gap by reading suggestions, inspecting data, writing one-sentence explanations of fixes, and reviewing what the agent has done. The verification habit is not a checklist item; it is the only thing that distinguishes AI-assisted research from research that has accepted unverified output. Maintain it, and the tools make you faster. Drop it, and the tools make you wrong faster.


