# Block 5: From Chat to Agent  {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```


In Block 4 you used Chat in Ask mode: you asked a question, it answered, and nothing on your machine or in your repository changed unless you copied the answer out yourself. This block introduces the shift that the rest of the workshop depends on, from an assistant that *advises* to an agent that *acts*. The configuration files you write in Block 5 exist because of this shift, so it is worth being precise about what changes, where it changes, and why it matters.

*⏱ ~30 minutes*


## Three surfaces, before we start

This workshop puts the same ideas in front of you through three different tools, and they do not all offer the same modes or honour the same files. You met this distinction in Block 4 for context; it returns here for actions, and it is the single thing most likely to make an exercise behave differently for the person next to you. Hold the three apart from the outset.

- **Web-based GitHub Chat** on github.com. It works in the context of the repository you open it from and can read its committed files, but it has no connection to your machine and runs no local code. Its agent is the **cloud agent**, which acts on the remote and returns a pull request. This is the surface that behaves the same for everyone, which is why several exercises specify it.

- **Positron Assistant**. The AI feature built into Positron, which for our purposes is the one most closely aligned with GitHub Copilot: it signs in with your GitHub account by default and reads Copilot's instruction files through the Copilot path. It can run in your local session and, in agent mode, act on your files. Its known weakness, established in Block 5, is that it does not reliably load skills.

- **Posit Assistant**. The newer tool, in preview, unifying Positron Assistant and Databot across Positron and RStudio. It inverts the trade-off: it has a native skills hook but does not read Copilot's `.github/` files. For the mode distinctions in this block it behaves as an agent that can act locally, but where it looks for configuration differs, and that difference is the subject of Block 5 rather than this one.

The blunt version: the cloud agent lives on github.com and is the same for everyone; Positron Assistant is the Copilot-aligned local agent; Posit Assistant is a separate local agent with its own configuration discovery. Where a step depends on which one you are using, it says so.


## Ask, Edit, Agent

Beneath the surface differences, GitHub Copilot Chat offers more than one *mode*, and they differ in what the assistant is permitted to do, not in how clever it is. The underlying model is the same; the leash is not.

- **Ask mode** answers in prose. It can read context and write code into the reply, but it changes nothing for you. Everything it suggests, you apply by hand. This is the mode you used throughout Block 4, and it is available on every surface, including web-based GitHub Chat.

- **Edit mode** proposes changes directly to named files. You give it a set of files to work on, it returns a diff across them, and you accept or reject the changes. It acts on your files, but only the ones you nominate, and only by editing them. Edit mode is a local-IDE feature; it is not the web Chat surface.

- **Agent mode** is given a goal rather than a file list. It decides which files to read, which to change, and what commands to run to reach the goal, then carries those steps out, often in a loop: make a change, run something, read the result, change again.

Agent mode runs in two distinct places, and the difference is where the changes land:

- The **in-IDE agent** (Positron Assistant, or Posit Assistant, or VS Code) works on the files on your own machine and runs commands in your local session, so its actions appear in your working directory in real time.

- The **cloud agent** on github.com works on your repository on the remote, in an environment you do not see, and delivers its work as a pull request you review and merge.

The progression is one of widening autonomy: Ask tells you, Edit changes what you point at, Agent decides what to change and does it. A short way to hold the distinction is that in Ask mode you are the hands and the agent is the advisor; in Agent mode the agent is both.

| Mode | Reads context | Writes code | Edits your files | Runs commands | Decides scope | Where it runs |
|---|---|---|---|---|---|---|
| Ask | Yes | Into the reply only | No | No | No | Any surface, incl. web |
| Edit | Yes | Yes | Files you nominate | No | No | Local IDE only |
| Agent | Yes | Yes | Any it judges relevant | Yes | Yes | In-IDE or cloud agent |


:::{.callout-important}

## Which modes you can actually run

These modes are not equally available on every plan, and the cloud agent in particular is gated.

- **In-IDE Chat (Ask, Edit, and agent mode)** is usable on the free Copilot tier, subject to monthly limits on the number of requests and on premium model use. For this workshop the free tier is enough to follow the exercises, though a heavy session can reach the cap; if your requests stop being answered, a limit rather than a bug is the likely cause. This applies to Positron Assistant signed in with a free GitHub account; Posit Assistant's own limits depend on the provider key you give it.

- **The cloud agent on github.com** currently requires a paid plan: GitHub Pro (which includes a 30-day free trial) or an eligible GitHub Education account. It is not part of the free tier. If you are on the free plan with no trial active, you can run agent mode inside your IDE but not the cloud agent on the website.

Plan terms change, so confirm your own access before relying on it rather than taking these limits as fixed. The practical consequence for the exercise below: everyone can run the in-IDE route, and the cloud-agent route on github.com is only open to those on Pro, a Pro trial, or Education. The exercise is written so the in-IDE route is the default and the cloud agent is the variant, not the reverse.

:::


## Why this is the moment configuration starts to matter

In Ask mode, a vague or unconventional answer costs you a re-read and a re-prompt. You are the filter: nothing reaches your files except what you choose to paste. An agent removes that filter. When the assistant edits files and runs commands on a goal you stated once, every convention it does not know, it will guess, and every guess lands in your project rather than in a reply you can ignore.

This is the practical reason the persistent-context mechanisms in Block 5 exist. A `copilot-instructions.md` that says you use the native pipe and write outputs only to a dedicated directory is mildly useful to an advisor and load-bearing for an agent, because the agent acts on those conventions unsupervised. The looser the leash, the more the standing instructions matter. Put plainly: in Ask mode you can afford to correct the assistant after the fact; in Agent mode you want the conventions in place before it starts, because "after the fact" now means after your files have changed.

This is also where the three surfaces stop being interchangeable. The instructions you write in Block 5 are honoured by web-based GitHub and by Positron Assistant, because both read the Copilot path, but Posit Assistant will not find them unless you place them under its own directories. An agent that cannot see your conventions does not become cautious; it proceeds anyway, on assumptions. So the surface you run the agent on decides whether the standing instructions you wrote are even in force, which is a reason to know which tool you are in before you hand it a goal.


:::{.callout-warning}

Autonomy and trust are not the same thing. An agent that can run commands can run a *wrong* command. In the in-IDE agent that command executes in your local session as it works; on the cloud agent it executes in the remote environment before the pull request reaches you. Either way it acts before you have read it. Two habits limit the damage, and both are reasons the Block 5 instructions file matters: forbid destructive operations in standing instructions (no `rm(list = ls())`, no overwriting raw data, outputs only to dedicated directories), and keep the agent working against version control so any change it makes is a diff you can inspect and revert. Never run an agent with uncommitted work you are not prepared to lose.

:::


## Exercise: the same task, three modes

This exercise makes the difference concrete by giving the same kind of task to each mode in turn, growing the scope as the autonomy grows. Steps 1 and 2 use a deliberately small change so the mode, not the task, is what you are watching; step 3 hands the agent a larger, end-to-end job that a single edit could not cover, because that is where an agent earns its name. Have your forked workshop repository to hand, committed and clean, so that anything the agent changes is a diff you can inspect.

**1. Ask mode.** This works on any surface. In Ask mode, prompt:

```
In R/clean_voles.R, I want to add a column mass_change that is
mass_post_g minus mass_pre_g. How do I do that?
```

Read the answer. It will tell you the line to add and roughly where, and possibly explain the `mutate()` call. Notice what has *not* happened: the file is unchanged. To apply this you would copy the line into the script yourself.

**2. Edit mode.** This is a local-IDE step; if you are on web-based GitHub, read it and move on. In Positron Assistant (or VS Code), switch to Edit mode, nominate `R/clean_voles.R` as the working file, and give the same instruction as an instruction rather than a question:

```
Add a column mass_change that is mass_post_g minus mass_pre_g.
```

Now you are offered a diff against that one file. Read the diff before accepting it. The assistant has written the change, but the decision to apply it is still yours, and its scope is fixed: it can only touch the file you named.

**3. Agent mode.** Now the task grows beyond a single edit. Rather than name a change, you state an outcome that spans the pipeline: a new derived variable, a plot that uses it, and the surrounding work needed to keep everything running. This is the kind of job that distinguishes an agent from an editor, because no one file is named and the agent must read the project, decide what to touch, and verify its own work.

This step has two routes. Use the in-IDE route if you are in Positron Assistant, Posit Assistant, or VS Code; use the cloud-agent route only if you are on GitHub Pro, a Pro trial, or Education (see the availability note above). The goal is the same for both; what differs is where the work happens and how it comes back to you.

*In-IDE agent (default route).* Switch the Chat panel to Agent mode and give it the goal:

```
In the voles pipeline, read data/voles_metabolism.csv through the existing
cleaning step, add a derived mass_change column (mass_post_g minus
mass_pre_g), and create a new script that produces an exploratory plot of
mass_change by treatment_arm. Make sure the cleaning step and the rest of
the pipeline still run end to end.
```

Watch what it does in your working directory. It cannot satisfy this from one edit: it must locate the cleaning step, decide whether the derived column belongs there or in a new script, write the plotting code in a new file, and run the pipeline to confirm nothing downstream broke. It works in a loop, editing then running then reading the result, and you see each change land in your working tree as it goes. This is precisely why you committed first: every file the agent writes or alters is a diff you can inspect, accept, or revert. Read the diff before you keep it; the agent's confidence that the pipeline "still runs" is its claim, and a claim is not a check until you have seen the output yourself.

*Cloud agent (Pro / Education route).* On github.com, hand the same goal to the cloud agent. The mechanics are worth seeing once, because a pull request is how an agent's work reaches a real project.

1. Open Copilot Chat on your repository, switch to agent mode, and give it the same goal as above. (You can also assign work through a GitHub Issue, but for this exercise prompting the agent directly is enough.)

2. The agent starts a session on the remote, in an environment you do not see. It creates a new branch to work on, rather than touching your default branch, so nothing it does affects your main line of work until you decide it should.

3. It then works through the goal on that branch, committing as it goes: reading the data, editing the cleaning step, adding the plotting script, and running the pipeline to check it. You can follow this progress in the session log while it runs.

4. When it finishes, it opens a **pull request** from its branch. A pull request is a proposed merge: a named, reviewable bundle of commits asking to be brought into your main branch. This is the unit you review, not the individual edits.

5. Open the pull request and read the **Files changed** tab. This is the diff: every line the agent added or removed, across every file it touched. Read it as you would a colleague's contribution, because that is what it is. Check that the derived column is computed correctly, that the plot uses the right variables, and that nothing unrelated was changed.

6. If you are satisfied, **merge** the pull request, which brings the changes into your main branch on the remote; then `git pull` locally to bring them to your machine. If you are not satisfied, you can comment on the pull request to ask the agent to revise it, or simply close it and discard the branch. Nothing reaches your main branch, or your local working tree, until you choose to merge.

The cloud agent's discipline is built into this flow: it never edits your working files directly, and its work arrives as a proposal you approve rather than a change you discover. The in-IDE agent gives you speed and immediacy; the cloud agent gives you a review gate by default.

Whichever route you took, notice the autonomy you handed over: you specified an outcome and the agent chose the steps, the files, and the order. The difference between the two routes is not how much it decides, which is the same, but where its decisions take effect: live in your working tree, or on a branch behind a pull request you approve first.

**4. The point of the comparison.** Re-read the results together. What differed across the modes was how much you decided versus how much the assistant decided. In Ask mode you placed every character. In Edit mode you approved a specific change to one file you named. In Agent mode you approved an outcome and trusted the agent to choose the files, the steps, and the order, whether that ran live in your working tree or on a branch behind a pull request. The step-3 task was large enough that you could not have specified it edit by edit in advance, which is the point: an agent is for the jobs where stating the goal is tractable but stating every step is not. Ask yourself whether the agent structured the pipeline the way you would have, whether the plot answers the question you meant, and whether you would have caught it had it not. That question, scaled up to a task you cannot read line by line, is exactly why the standing instructions in Block 5 are written before the agent runs, not after, and why it matters that the surface you run on is one that reads them.


## Check your understanding

**What is the key difference between Ask mode and Agent mode in GitHub Copilot?**

`r longmcq(c(
  "Agent mode uses a more capable underlying model than Ask mode",
  answer = "Ask mode answers without changing anything, whereas Agent mode can decide which files to change and which commands to run, and carries those actions out",
  "Ask mode is free and Agent mode is always paid",
  "Agent mode can only be used on github.com, never in an IDE"
))`

**Where does the in-IDE agent apply its changes, compared with the cloud agent on github.com?**

`r longmcq(c(
  "Both apply changes only as a pull request you review before merging",
  answer = "The in-IDE agent edits the files in your local working directory in real time, whereas the cloud agent works on the remote and returns a pull request",
  "Both edit your local files directly with no review step",
  "The in-IDE agent cannot run commands, only the cloud agent can"
))`

**Why do persistent instruction files matter more once you are using an agent than when you are using Ask mode?**

`r longmcq(c(
  "Agents cannot read pasted context, so the file is the only way to inform them",
  answer = "An agent acts on its assumptions by editing files and running commands unsupervised, so conventions it does not know are applied to your project rather than merely appearing in a reply you can discard",
  "Instruction files are required for Agent mode to function at all",
  "Ask mode ignores instruction files entirely"
))`
