# Block 6: Functionalise a script step and prove the result is unchanged {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```


Suppose you have a computation that already works and you want the cloud agent to
tidy it into a reusable function. The agent can do that in minutes. The difficulty
is verification: a refactor that "improves" code can quietly change its output, so
how do you confidently check that the function still does what the script did? The
answer this part teaches is to freeze the original result before the agent touches
anything. You run the analysis once, save its output to a reference file (an
`.rds`), and then require the agent's function to reproduce that reference exactly.
You hold the reference; the agent does the refactor and proves that it matches. The
reference is the check, and because you captured it, the check cannot be quietly
redefined by the agent.

*⏱ ~30 minutes*


## Recap: in-IDE Agent versus cloud agent

The distinction from Part 1 matters here. In-IDE Agent mode, available in Positron and VS Code, edits files in your editor in real time while you watch. The cloud agent runs on GitHub's servers, takes an issue as input, creates a branch and a pull request without you watching, and notifies you when it has finished. This block uses the cloud agent throughout, because it is platform-agnostic and exposes the full software-engineering workflow the workshop is building towards. RStudio users are therefore on equal footing in this part.

## The repository layout
Before any code, here is the full layout for this part. Each file is marked by who produces it: [provided] ships in the template, [you] is the one file you generate in Step 0, and [agent] is what the cloud agent creates or rewires. The three-way status is the conceptual spine of the whole exercise, so refer back to it as you go.

```
repo/
├── .github/
│   ├── workflows/
│   │   ├── copilot-setup-steps.yml   [provided]  builds the agent's R environment
│   │   └── r-checks.yml              [provided]  independent CI check on push and PR
│   └── skills/
│       └── characterisation-test/
│           └── SKILL.md              [provided]  tells the agent how to behave
├── data/
│   └── voles_metabolism.csv          [provided]  the dataset
├── R/
│   ├── 01_load_data.R                [provided]  loads the CSV, runs glimpse()
│   ├── 02_analysis.R                 [provided → rewired by agent]
│   └── functions/
│       └── summary_functions.R       [agent]     the extracted function
├── tests/
│   ├── fixtures/
│   │   ├── make_reference.R          [provided]  captures the reference
│   │   └── metabolic_summary.rds     [you]       generated in Step 0, then frozen
│   └── test_summary_functions.R      [agent]     compares function against reference
└── renv.lock                         [provided]  pins tidyverse, here, testthat, renv
```

The single most important property of this layout: the `.rds` under tests/fixtures/ is generated by you and by nobody else. Everything else in the trust argument follows from that.

## What the template already provides (read, do not write)

Recall from the GitHub session that GitHub Actions runs automated workflows in response to repository events such as a push or a pull request. Two such workflows sit in .github/workflows/, both placed there for you.

- `r-checks.yml` — the independent check

This runs the tests on pushes to main and on every pull request. It is the check that neither you nor the agent can fake, because it runs on GitHub's infrastructure from a clean checkout. A green tick here is the signal you wait for at the end.

`copilot-setup-steps.yml` — the agent's environment

This is the file the original tutorial glossed, and it is worth your attention, because it is what makes the agent's environment resemble your IDE. The cloud agent runs the tests in the environment this file builds. If that environment does not restore renv.lock, the agent's tests will fail (or, worse, pass) for reasons that have nothing to do with the refactor, and the resemblance to your local setup is lost.

Two requirements are load-bearing:

- The job must be named exactly copilot-setup-steps. GitHub looks for that job name specifically; rename it and the agent silently falls back to a bare environment without your packages.
The job must restore the lockfile, so the agent computes against the same pinned package versions that r-checks will later use.

- A conventional form, using the r-lib/actions setup actions, looks like the following. Check this against your actual template file rather than copying it blindly; this reproduces the standard pattern, not necessarily your exact configuration.


:::{.callout-warning}

This setup steers towards reproducibility because setup-renv reads renv.lock and installs the exact pinned versions, so the agent computes against the same package versions that r-checks.yml will later use from a clean checkout. This is what makes the agent's environment resemble your local IDE and is why the agent's test run and the independent check should agree. 

Its also what might make this workflow more fragile if the lockfile is out of step or does not restore packages

:::

```
name: Copilot setup steps

on: workflow_dispatch

jobs:

  copilot-setup-steps:
    runs-on: ubuntu-latest
    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: "4.5.2"      # match the version in renv.lock
          use-public-rspm: true
          
      - name: Set up renv
        uses: r-lib/actions/setup-renv@v2
        env:
          RENV_CONFIG_PAK_ENABLED: "true"

      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2

```

`setup-renv` reads `renv.lock` and installs exactly the pinned versions, so the agent's tidyverse, here and testthat match yours. The practical consequence: when you reach Step 4, the agent's own test run and the independent r-checks run should agree. If they disagree, suspect the environment before suspecting the code.

### The two scripts and the skill

The repository holds two short scripts. `R/01_load_data.R` loads the data and runs `glimpse()`; `R/02_analysis.R` sources it and computes the summary you will functionalise:

```{r}
#| eval: false
# R/02_analysis.R — the original, trusted version
library(tidyverse)
library(here)

# 01_load_data.R reads the CSV into `voles` and runs glimpse()
source(here::here("R", "01_load_data.R"))

metabolic_summary <- voles |>
  group_by(treatment_arm, sex) |>
  summarise(
    mean_rate = mean(metabolic_rate_post),
    sd_rate   = sd(metabolic_rate_post),
    n         = n(),
    .groups   = "drop"
  )
```

And a skill telling the agent how to behave for this kind of job:

```markdown
# .github/skills/characterisation-test/SKILL.md
---
name: characterisation-test
description: >-
  Use for background (cloud agent) tasks that move a computation from a script into a
  function in R/functions/ and verify it reproduces a saved reference. The reference is
  captured by the investigator, never by the agent.
---

# Functionalise and characterise

## What you must not do
Do not create, regenerate, or modify the reference output, and do not touch any file under
tests/fixtures/. If the fixture is missing, stop and ask the investigator to capture it. A
reference you make or alter from your own code proves nothing.

## Procedure
1. Write the computation as a function in R/functions/, with a roxygen2 block, argument
   checks, and one @examples entry. Reproduce the original output exactly, including column
   and row order.
2. Add a test that loads the fixture, runs the function, and compares with expect_equal().
3. Update the originating script (R/02_analysis.R) to call the new function in place of the
   inline computation, so the script and the function do not drift apart. Do not re-run any
   reference-capture script and do not touch tests/fixtures/.

## Verification
Run the test in your own environment first. Then push and wait for r-checks. If it fails,
read the log, fix the specific cause, and push again. Done means r-checks is green.

## If a failure persists after three attempts
Stop and comment on the pull request explaining what you tried and the remaining error.
Do not make speculative changes.
```

### Before you start: one repository setting

This repository holds no secrets, so you can turn on the setting that lets the Copilot coding agent's workflows run without manual approval, which lets you watch the checks run live. By default GitHub blocks them, treating the agent as an outside contributor, and you would otherwise sit at an "Approve and run" gate without understanding why.

1. In the repository, go to Settings → Actions → General.

2. Under Fork pull request workflows, allow GitHub Actions to run on pull requests from the agent.

3. Save.

:::{.callout-warning}

This is a convenience setting and you should be actively Approving runs in a standard workflow

:::


## Step 0 — You capture the reference (this stays with you)

Before the agent is involved at all, run `R/02_analysis.R` yourself and confirm
`metabolic_summary` looks as it should. This is the output you want to preserve. Then
run the capture script below, once, and commit the resulting `.rds`. This frozen
output is the trust anchor, and the agent never touches it. Run the capture from
within the project, so `renv` is active and the reference reflects the same locked
package versions that `r-checks` will restore. Capture it now, before the refactor:
afterwards `02_analysis.R` computes the summary through the very function under test,
so re-capturing from it would compare the function against itself. For this exercise
you capture once and leave the reference frozen.

```{r}
#| eval: false
# tests/fixtures/make_reference.R
# Capture the trusted reference once, from within the project so renv is active,
# so the reference reflects the locked package versions r-checks restores.
# Run this BEFORE the agent refactors 02_analysis.R, and capture only once: after
# the refactor that script computes the summary via the function under test, so a
# later capture would compare the function against itself.

if (!requireNamespace("renv", quietly = TRUE) || is.null(renv::project())) {
  stop(
    "Run this from within the project, with renv active, so the reference ",
    "matches the locked environment used by r-checks.",
    call. = FALSE
  )
}

source(here::here("R", "02_analysis.R"))
saveRDS(
  metabolic_summary,
  here::here("tests", "fixtures", "metabolic_summary.rds")
)
```

## Step 1 — Read the skill

Open `.github/skills/characterisation-test/SKILL.md` and find the one line that matters: the agent must not create, regenerate or modify the reference, and must not touch anything under `tests/fixtures/`. That single instruction is what stops the verification becoming circular, and it is your first aim in miniature, a few words of instruction setting a boundary the agent will respect.

## Step 2 — Write a thin issue and hand it over

```text
Title: Functionalise the metabolic-rate summary and verify it is unchanged

R/02_analysis.R computes metabolic_summary. Move that computation into a documented
function in R/functions/, have the script call it, and add a test that checks the function
reproduces the saved reference in tests/fixtures/metabolic_summary.rds.

Use the characterisation-test skill and follow the project conventions.
Push, make r-checks pass, and tell me when it is ready to review.
```

Notice how little you wrote. The return shape, the roxygen, the naming and the
assertion style are absent, because the skill and your standing instructions
already hold them. That is the lesson: short instructions work because the
architecture carries the standards.

## Step 3 — Watch it work, and probably stumble

The agent moves the computation into a function, writes the test, runs it in its
own environment, and pushes. The likely first failure is instructive. A tidy
refactor often switches `group_by()` for the modern `.by`, and the two do not
order their output the same way: `group_by()` sorts the grouping keys, whereas
`.by` returns groups in order of first appearance. The numbers are right, but
the rows are arranged differently, so `expect_equal()` reports a mismatch. Watch
the agent read that, add an `arrange()` to match the original order, and go
green. If it passes first time, walk through the prepared failed-then-fixed pull
request instead, so the recover-from-failure loop is seen either way. The point
to land is that improving code during a refactor can quietly change the output,
and the characterisation test is what catches it.

## Step 4 — Review the green pull request

```{r}
#| eval: false
# R/functions/summary_functions.R
#' Summarise post-treatment metabolic rate by treatment arm and sex
#'
#' @param data A data frame containing the columns treatment_arm, sex and
#'   metabolic_rate_post.
#' @return A tibble with one row per treatment_arm and sex, giving the mean,
#'   standard deviation and n.
#' @examples
#' demo <- tibble::tibble(
#'   treatment_arm = c("Control", "Control", "Low_dose", "Low_dose"),
#'   sex = c("F", "F", "M", "M"),
#'   metabolic_rate_post = c(1.2, 1.3, 1.6, 1.7)
#' )
#' summarise_metabolic_rate(demo)
summarise_metabolic_rate <- function(data) {
  stopifnot(is.data.frame(data))
  required <- c("treatment_arm", "sex", "metabolic_rate_post")
  missing_cols <- setdiff(required, names(data))
  if (length(missing_cols) > 0) {
    stop(
      "`data` is missing required column(s): ",
      paste(missing_cols, collapse = ", "),
      call. = FALSE
    )
  }

  data |>
    dplyr::summarise(
      mean_rate = mean(metabolic_rate_post),
      sd_rate   = sd(metabolic_rate_post),
      n         = dplyr::n(),
      .by = c(treatment_arm, sex)
    ) |>
    dplyr::arrange(treatment_arm, sex)
}
```

The tests: 

```{r}
#| eval: false
# tests/test_summary_functions.R
library(testthat)
source(here::here("R", "functions", "summary_functions.R"))

test_that("summarise_metabolic_rate reproduces the original summary", {
  reference <- readRDS(here::here("tests", "fixtures", "metabolic_summary.rds"))
  voles <- readr::read_csv(
    here::here("data", "voles_metabolism.csv"),
    show_col_types = FALSE
  )
  expect_equal(summarise_metabolic_rate(voles), reference, tolerance = 1e-8)
})
```

And `R/02_analysis.R`, now a thin caller rather than the inline pipeline:

```{r}
#| eval: false
# R/02_analysis.R — after the refactor
library(tidyverse)
library(here)

source(here::here("R", "01_load_data.R"))
source(here::here("R", "functions", "summary_functions.R"))

metabolic_summary <- summarise_metabolic_rate(voles)
```

Read four things. 

1. The green tick, in the Checks tab or at the bottom of the Conversation tab. Read it as evidence that the tests pass reproducibly on locked infrastructure. It is not evidence that the test is the right test; that is what the remaining items establish.

2. The test body, in the Files changed tab. Confirm it reads tests/fixtures/metabolic_summary.rds (your reference) and compares against it, rather than building its own reference inside the test.

3. R/02_analysis.R, also in Files changed. Confirm it now calls summarise_metabolic_rate() and no longer carries the old inline pipeline, so the function has genuinely replaced the code it was extracted from.

4. The list of changed files, at the left of the Files changed tab. Confirm tests/fixtures/metabolic_summary.rds does not appear anywhere in it. This is the most important item. Because the .rds is binary, git cannot show you what changed inside it, so the only safeguard is that it must not appear in the diff at all. If it appears, the agent has touched your trust anchor and the verification is circular, whatever the green tick says. In that case, reject the pull request and check the agent's commits against the skill's prohibition.



## Debrief — when is this worth it?

Use a CI agent task when the work is tedious, has a clear pass-or-fail check, and
you are content to review afterwards rather than steer live. Functionalising a
script step under a characterisation test fits all three. Deciding what the
script should compute in the first place, and judging whether the result is
scientifically sensible, do not, and stay with you.

## Check your understanding

**A green `r-checks` tick on the pull request tells you that:**

`r longmcq(c(
  "the function is correct and the result is scientifically sensible",
  answer = "the tests pass reproducibly on clean, locked infrastructure, but not that the test compares against an honest reference",
  "nothing useful, because the agent controls the outcome",
  "both that the result is reproducible and that the verification is non-circular"
))`

**True or false: the reference fixture is captured by the investigator rather than the agent because a reference generated from the agent's own output would make the verification circular.** `r torf(TRUE)`



:::{.task}
::::{.task-header}
Automated code review
::::
::::{.task-container}

Discuss what parts of code review you would hand over to an Agent.

What would the skill look like? 



::::
:::




## What comes next

This workshop is the foundation for a separate strand of work on Agent Skills. Three general-purpose skills are in development: tidyverse-style code optimisation, the publication-figure skill you saw demonstrated, and a code-review skill that applies a structured checklist to your work before submission. These will be released with documentation and worked examples. The `characterisation-test` skill you used with the cloud agent in Part 6 is a narrower, task-specific example of the same architecture, and it ships in the workshop template today. The architecture is intended to be modular; you can add your own skills as you identify recurring procedures in your work.


## Self-study pointers

For the underlying tools and conventions:

- The GitHub Copilot documentation is the canonical source for current behaviour and capabilities, and it changes faster than written materials can keep up. The model selector, the in-IDE Chat panel, and the cloud-agent assignment feature have all changed materially within the last twelve months.

- The TADA guidelines for analytical code sharing (Ivimey-Cook et al., 2025, EcoEvoRxiv preprint) cover the reproducibility scaffolding the skills assume.

- The tidyverse style guide (style.tidyverse.org) is the convention base layer; most of what you specify in your `copilot-instructions.md` will sit on top of it.



## Final note

AI tools accelerate code production. They do not accelerate understanding. The two diverge unless you actively close the gap by reading suggestions, inspecting data, writing one-sentence explanations of fixes, and reviewing what the agent has done. The verification habit is not a checklist item; it is the only thing that distinguishes AI-assisted research from research that has accepted unverified output. Maintain it, and the tools make you faster. Drop it, and the tools make you wrong faster.


