# Block 6: Custom instructions and skills  {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
```


```{r}
#| include: false
library(tidyverse)
library(webexercises)
```


Block 5 argued that an agent acts on conventions it cannot see, and that the looser the leash, the more those standing conventions matter. This block is how you make the conventions visible to the assistant in advance, so you do not re-state them in every comment and every Chat session. It falls into two parts. **Part A** covers the always-on mechanisms, custom instructions and scoped instructions, which load automatically and behave the same for everyone on a surface that reads them; you will test these in Ask and Edit mode, the modes from Block 5 where you remain the filter. **Part B** moves to agent mode and to skills, which are invoked rather than always loaded, are far more surface-dependent, and introduce Plan mode as the review step that sits between an editor and a fully autonomous agent.

*⏱ ~60 minutes*


## Before you start: which assistant are you actually using?

:::{.callout-warning collapse="true"}

You met Positron Assistant and Posit Assistant in Block 5. This is the detail that matters now that you are writing configuration files, because the two read those files from different places, and confusing them is the single most common reason a configuration file appears to be ignored.

**Positron Assistant** is the Copilot-aligned tool. For the purposes of this block, its useful property is that it reads GitHub Copilot's instruction files through the Copilot path: the `.github/` conventions in Part A are honoured. Its limitation is skills. It does not reliably load them; whether a given request reaches the path that would honour a skill is not fully determined by any setting you can see, and the routing internals are undocumented, so a skill that fires for one user can be silently ignored for another with an apparently identical setup. Treat instruction files as dependable in Positron Assistant and skills as not.

**Posit Assistant** is the newer, cross-IDE tool, in preview across both Positron and RStudio. It inverts the trade-off. It loads skills automatically when they are relevant, but it searches its own location, under `.positai/skills/` or `.agents/skills/` (configurable via `skills.paths`), not Copilot's `.github/`. So the `copilot-instructions.md`, scoped instructions, and `.github/skills` in this block are not picked up by Posit Assistant unless you place or point them at one of its own directories.

The blunt rule of thumb: author for Copilot in `.github/`, and rely on instruction files rather than skills if you are in Positron Assistant. If you also use Posit Assistant, treat its skill discovery as a separate, parallel setup rather than assuming one folder serves both. Where this block depends on behaviour that differs between the two tools it says so, and where the behaviour is undocumented it says that too, because both are moving targets and guessing wastes your time.

:::


# Part A: Instructions and scoping {.unnumbered}

The two mechanisms in this part share one property that makes them dependable: they load automatically, without your asking. You do not invoke them, you do not name them, and you do not have to remember they exist. That is also why they are the right place to start, because everything you learn here works the same for everyone on a surface that reads the Copilot path, with none of the surface-dependence that complicates Part B.


## Providing context to models

GitHub Copilot, like any LLM client, provides several ways to supply context, distinguished by how they load.

**Workspace-level custom instructions** live in `.github/copilot-instructions.md`. Copilot loads this file automatically for every request on a surface that reads the Copilot path: web-based GitHub Chat, the cloud agent, code review, and Positron Assistant, as Block 5 set out. (Posit Assistant does not read it, per the orientation above.) Use it for things that are always true of your project: which packages you use, your naming conventions, your file structure, your statistical conventions. The file should be short and broad in scope; treat two pages as a ceiling, not a target.

**Scoped instructions** live in `.github/instructions/*.instructions.md`, with a YAML front-matter `applyTo` glob that restricts the file to particular paths (for example `applyTo: "**/*.qmd"` to apply only to Quarto documents). Use them for rules that apply to one kind of file but not the rest. They load automatically when Copilot works on a matching file, with one important limitation on GitHub.com: there they are currently honoured by the cloud agent and by code review, but not by in-IDE Chat or inline completion. If your work is mainly in the editor, confirm your surface before relying on this mechanism.

**Agent Skills** live in `.github/skills/<skill-name>/SKILL.md`, with a YAML front-matter `name` and `description`. Unlike the first two mechanisms, skills are not always loaded; they are invoked. Either you type `/skill-name`, or the assistant infers from the description that the skill is relevant. Skills are the subject of Part B; they are listed here only to complete the picture.

A fourth, related mechanism is custom prompt files (`.github/prompts/*.prompt.md`), invoked explicitly as slash commands. These are outside the scope of this workshop.

The mechanisms answer different questions. Custom instructions answer "how should code in this project be written?" Scoped instructions answer "how should this kind of file be written?" Skills answer "how should this specific task be executed?"


## How to write effective instructions

A `copilot-instructions.md` file is loaded into every interaction on a reading surface, and is read by the cloud agent before it starts work, so its length is a recurring cost and brevity is a virtue. A handful of accurate lines outperforms a long file.

A few principles carry most of the value:

- Write each instruction as a single, simple statement. If you have several points, use several lines rather than one dense paragraph; the model parses discrete statements more reliably than prose.

- Describe the conventions you actually follow, not the ones you aspire to. An instruction your own code routinely breaks teaches the model a rule the surrounding code contradicts, which makes its suggestions less consistent.

- Be concrete and positive. "Use the native pipe `|>`" and "snake_case for all object names" are actionable; "write clean, idiomatic code" is not.

- Avoid contradiction and overlap. When a workspace instruction and a scoped instruction conflict, Copilot's choice between them is not deterministic, so do not rely on one quietly overriding the other.

Some other set-up and principles can be found at the
[GitHub documentation on onboarding an instruction file](https://docs.github.com/en/copilot/how-tos/copilot-on-github/customize-copilot/add-custom-instructions/add-repository-instructions).

`r hide("Example")`

```
# Copilot Instructions: R Analytical Repository

This repository contains analytical R code for computationally reproducible
research. Most items below are written for you to follow as you generate code.
A short Verification section lists the few things that are also checked in CI,
because a silent failure there is costly.

## Canary
Start any response that writes or edits R code with this as the first line of
the first code block:
`# copilot-instructions: active v4`

## Priorities
When two instructions conflict and no test settles it, prefer in this order:
correctness, reproducibility, transparency, interpretability, maintainability.
Note the trade-off in the pull request description.

## Agent behaviour
Never wait for human input in committed code. Do not use `readline()`,
`menu()`, `file.choose()`, `browser()`, or `View()`. When an input is missing,
add a `# DEFAULT:` comment naming the assumption and the value you chose, then
carry on.

## Naming and style
- Use the native pipe `|>`, not `%>%`.
- Use snake_case for all objects and functions.
- Assign with `<-`, not `=`.
- Write `TRUE` and `FALSE`, never `T` and `F`.
- Suffix dataframe and tibble objects with `_df`; raw import objects use `_raw`
  instead and are never reassigned.
- Suffix fitted models with `_model`.
- Suffix ggplot objects with `_plot`.
- Namespace-qualify calls (`pkg::fun()`) inside `R/functions/`. In analysis
  scripts, load packages with `library()` at the top and use bare names below.
- Avoid hidden global state and partial argument matching.
- Abstract logic into a function once it appears in three or more places.

## Reproducibility
- Never use `setwd()` or `rm(list = ls())`.
- Build every path with `here::here()`; relative paths only.
- Assume each script runs from a fresh R session.
- Manage dependencies with `renv`.
- Set a seed before any stochastic operation.
- Save `sessioninfo::session_info()` to the output directory at the end of each
  top-level script.

## Dependencies (renv)
- Manage all dependencies with `renv`; the project library is the only source
  of truth.
- Never call `install.packages()`. Add a package with `renv::install()`, then
  `renv::snapshot(prompt = FALSE)`.
- When `renv.lock` already lists the packages you need, restore rather than
  install: `renv::restore(prompt = FALSE)`.
- Never edit `renv.lock` by hand.
- Set `options(renv.config.auto.snapshot = FALSE)` so the lockfile is never
  written as a side effect; snapshot only deliberately.
- Use `renv::snapshot(type = "all")` so the lockfile is reproducible across
  runs rather than depending on usage detection.

## Reading and writing data
- Always use functions from readr (`readr::read_*(col_types = ...)`), not
  `read.*`.
- Keep raw data files immutable; never overwrite anything in the raw data
  directory.
- Separate raw data, derived data, scripts, outputs, and documentation.
- Write outputs only to dedicated output directories, never outside the
  repository.
- Assign cleaned and derived results to new objects; do not overwrite the raw
  import object.

## Data integrity
- Never drop missing data silently. Filter with a stated reason, or use
  `tidyr::drop_na()` on named columns.
- Prefer `purrr::map_*()` and `vapply()` over `sapply()`.

## Statistical practice
- Distinguish exploratory from confirmatory analyses, in prose or chunk labels.
- Validate model assumptions with `performance::check_model()` for regression
  models and record the inspection.

## Figures
For any figure, quick or polished:
- Use tidyverse style, the native pipe, and British spelling.
- Build paths with `here::here()`.
- No pie charts, dual y-axes, 3D effects, or dynamite plots (bar plus error
  bar) when raw data exist.
- Never use `theme_grey()`; default to `theme_minimal()` or `theme_classic()`.
- For categorical groups, add `shape` or `linetype` alongside colour.
- Use `linewidth`, not `size`, for line and density layers.
- For figures meant for an audience, follow the pub-figures skill in full.

## Code chunks
- Keep chunks short: one logical step each (import, clean, transform, model,
  diagnose, or visualise), and no more than about 40 lines.
- Comment the rationale for the step, not the mechanics of each line.
- Include at least one comment per chunk.

## Quarto
For `.qmd` files: render from a clean session, give chunks informative labels,
and avoid hidden dependencies between chunks.

## Verification (also checked in CI)
These are enforced because a silent miss is costly:
- No `setwd(`, `%>%`, `install.packages(`, or absolute paths in committed code.
- `renv::status()` reports a synchronised library.
- Each script and `.qmd` runs or renders in a fresh `--vanilla` session without
  error.
- The raw data directory is unchanged by any run.
- Outputs resolve inside the repository.

## Specialist instructions and skills
Specialised instruction files hold domain standards and repository policy;
skills hold procedural workflows. When one applies, follow it alongside these
instructions. Skill selection follows each skill's own description.

```

`r unhide()`


## Activity: Build your copilot-instructions.md file

:::{.task}
::::{.task-header}
Build your copilot-instructions.md file
::::
::::{.task-container}

In the workshop template repository (which you forked at the start of the day), you will create `.github/copilot-instructions.md`. Rather than start from a blank file, generate a skeleton and then refine it. The refining is the real exercise.

**Step 1: Before and after.** This is a controlled before-and-after: the same prompt, run twice in the same mode, with the instructions file the only thing that changes between the runs. Unlike the Block 5 exercise, where the mode changed between runs, here the mode is held fixed (Ask) and the file is the single variable.

Run the prompt, in Ask mode, with no instructions file present:

```
Read in data/voles_metabolism.csv and make a quick exploratory plot of
mass_post_g by treatment arm.
```

> You can do this in the Positron IDE or by heading to the repository on GitHub.

Keep this answer. It is your baseline.

**Step 2: Draft a skeleton.**

*If you are in Positron*: open the assistant chat and prompt, "Create .github/copilot-instructions.md by reading the current workspace." The assistant reads the repository and drafts a file covering the structure, languages, and packages it detects.

*If you are in RStudio*: there is no Copilot Chat to generate the file for you, so create `.github/copilot-instructions.md` by hand. Either take the example above, or, if it is easier, push an empty file and use GitHub.com's web Chat to generate a draft once it is on the remote. Hand-authoring is fine; the file is short.

**Step 3: Trim it hard.** Generated instructions run long and generic. Cut the draft down to what is always true of the project and broadly useful, treating two pages as a ceiling, not a target. Delete anything the model padded or guessed at.

**Step 4: Make it yours.** The generator reads the workshop template, not your own work, and it cannot read your judgement. Edit the file so that it specifies, at minimum:

1. **One naming or style convention you actually hold** that the generator did not infer: the native pipe over `%>%`, `<-` over `=`, a suffix scheme for objects, whichever you genuinely follow. The test is whether your own existing code obeys it.

2. **One project-structure fact** the model cannot guess from a fresh template: where raw data live and that they are immutable, where outputs are written, where functions are sourced from. State it as a single declarative line.

3. **One statistical or reproducibility convention** that is always true of your work: that paths are built with `here::here()`, that a seed is set before any stochastic step, that exploratory and confirmatory analyses are labelled as such.

4. **A canary line** you can verify. Instruct Copilot to begin any response that writes R code with a fixed first comment, for example `# copilot-instructions: active`. This is the cheapest possible test of whether the file is being read at all: if the canary is absent, the file is not loading, and you are debugging the wrong thing.

Keep each as a single, simple statement. Four accurate lines you can verify are worth more than two pages of generated prose you cannot.

**Step 5: Commit and push.** Commit `.github/copilot-instructions.md` and push to your fork on GitHub. This is what makes the file available where it is read most reliably, and it is the point at which Positron and RStudio users converge: from here on, everyone works the same way.

**Step 6: Verify.** Open a fresh chat session, on GitHub.com or in Positron, and run the same prompt as Step 1:

```
Read in data/voles_metabolism.csv and make a quick exploratory plot of
mass_post_g by treatment arm.
```

Read the answer against your Step 1 baseline. If your canary line appears, the file is loading. If the plot follows the conventions you wrote in Step 4, in particular the one only you could have supplied, the instructions are doing their job. If the canary is missing, the file is not being found: check that it sits at exactly `.github/copilot-instructions.md`, that it is committed and pushed, and that instruction files are enabled in your settings.

::::
:::

Compare your answer with what someone else gets from their own instructions file. The differences are the personality of your standards.


## Activity: Write a scoped instruction for Quarto documents

A scoped instruction is a file of rules that applies only to certain files in your project. It lives in `.github/instructions/` and carries an `applyTo` glob in its front matter that names the paths it covers, for example `applyTo: "**/*.qmd"` to apply only to Quarto documents. When Copilot works on a file that matches the glob, the rules load automatically; when it works on any other file, they do not.

:::{.callout-note}

## When to use a scoped file

The test for whether something belongs in a scoped file rather than the
always-on `copilot-instructions.md` is simple: would the rule be wrong, not
merely irrelevant, in the files the glob excludes? If a rule is true
everywhere (use the native pipe, use snake_case), it belongs in the always-on
file; putting it in a scoped file gains nothing and adds the scoped mechanism's
reliability problems. A rule earns a scoped file only when it is a property of
one file type and would mislead the model elsewhere.

:::

Quarto `.qmd` files are a clear case for an R user, because they carry concerns a plain `.R` script does not have at all: chunk options, how code is echoed into the rendered document, figure captions and dimensions, and a working-directory trap specific to rendering. An instruction to set `echo: false` is correct inside a `.qmd` and meaningless inside `R/02_analysis.R`.

:::{.task}
::::{.task-header}
Create a scoped instruction file
::::
::::{.task-container}

Create `.github/instructions/quarto.instructions.md` with the following
content:

```
---
applyTo: "**/*.qmd"
---

# Quarto document conventions

These rules apply to Quarto source files only. They concern how code is
presented and rendered in the document, not the analysis logic itself.

## Chunk options
- Set chunk options with the hashpipe (`#|`) inside the chunk, not in the
  `{r}` header. For example `#| echo: false`, not `{r, echo=FALSE}`.
- Give every code chunk a label (`#| label: fig-mass-by-arm`) so cross-
  references and cached results are stable.
- Every chunk that produces a figure must set `#| fig-cap:` with a caption;
  do not emit an uncaptioned figure into a rendered document.

## Echo and output for a written report
- Default to `#| echo: false` so the rendered document shows results, not
  source, unless the document is explicitly a code walkthrough.
- Use `#| warning: false` and `#| message: false` on chunks that load
  packages or fit models, so loading noise does not appear in the output.

## Figures
- Label figure chunks with a `fig-` prefix and reference them in prose with
  `@fig-...`, not by a hardcoded number.
- Set figure dimensions explicitly with `#| fig-width:` and `#| fig-height:`
  rather than relying on defaults.

## Reproducibility
- Resolve paths with `here::here()` inside chunks; never assume the working
  directory is the document's folder, because Quarto and the R session may
  differ on this.

```

**Test it.** With the scoped file committed, open a `.qmd` file and, in Edit mode, ask Copilot to add a code chunk that reads the voles data and plots `mass_post_g` by `treatment_arm`. A correct response uses hashpipe chunk options, labels the chunk with a `fig-` prefix, sets a `fig-cap:`, and resolves the path with `here::here()`. Then open a plain `.R` script and ask for the same plot: the hashpipe and caption rules should not appear, because the scoped file does not apply there. That contrast, the same request producing different output in the two file types, is the demonstration.

::::
:::

Part A ends here. Everything so far has loaded automatically and left you, in Ask and Edit mode, as the filter on what reaches your files. Part B removes that filter.


# Part B: Agent mode, Plan mode, and skills {.unnumbered}

Part A's mechanisms loaded whether or not you thought about them. Skills are different: they are invoked, not always loaded, and they live in an agent's world, where the assistant acts rather than advises. Before the skill itself, one more mode completes the set you began in Block 5.


## Plan mode: the review step before action

Block 5 set out three modes: Ask, which answers; Edit, which changes files you name; and Agent, which decides scope and acts. There is a fourth, **Plan mode**, and it sits between Edit and Agent in the amount of autonomy you hand over.

Plan mode is read-only. Given a goal, it does not change anything; it produces an ordered, step-by-step plan of what it *would* do, which files it would touch, and in what order, and waits for you to approve, amend, or reject that plan before any of it executes. Where Edit shows you a diff for one file you nominated, Plan shows you the intended shape of a whole multi-step job before a single file changes. It is the review gate for work too large to specify edit by edit but too consequential to hand straight to an autonomous agent.

The progression is now complete, and it is one of widening autonomy with a review point inserted before the riskiest step:

| Mode | What you get back | What changes | Your control point |
|---|---|---|---|
| Ask | Prose and code in the reply | Nothing | You apply everything by hand |
| Edit | A diff for files you named | Those files, on accept | You accept or reject the diff |
| Plan | An ordered plan of intended steps | Nothing, until you approve | You approve the plan before execution |
| Agent | Executed changes across the project | Files and state, as it works | You review the result, or the pull request |

Plan mode is available in Positron Assistant and Posit Assistant. The natural workflow for any non-trivial agent task is Plan first, read the plan, then let it execute, rather than handing a goal straight to Agent mode and discovering its choices after the fact.

:::{.task}
::::{.task-header}
Plan, review, then execute
::::
::::{.task-container}

In Positron Assistant or Posit Assistant, with your forked repository committed and clean, switch to **Plan mode** and give it a goal that spans more than one step:

```
In the voles pipeline, add a derived mass_change column (mass_post_g minus
mass_pre_g) to the cleaning step, then create a new script that plots
mass_change by treatment_arm, and make sure the pipeline still runs end to
end.
```

Read the plan it returns before doing anything else. It should name the files it intends to touch, the order it will work in, and how it will check the pipeline. This is the moment to catch a wrong assumption: if it proposes editing the raw data, or putting the plot in the wrong place, or skipping the end-to-end check, you say so now, while nothing has changed. Amend the plan in conversation, then approve it and let it execute. Compare the experience with handing the same goal straight to Agent mode in Block 5: the difference is that you saw the intent before the action, not the action before the review.

::::
:::


## Agent skills

A skill is a packaged procedure the assistant loads when a task calls for it, rather than a set of rules attached to a file type or applied to every request. For Copilot it lives in `.github/skills/<skill-name>/SKILL.md` and carries a `name` and a `description` in its front matter; the description is what the assistant reads to decide the skill is relevant, or you invoke it directly by name. Because a skill loads only when needed, it can be longer and carry supporting material, templates, tables, and reference files that an always-on instruction file could not afford to include on every request.

A caution before the demonstration, following from the orientation at the start of this block. Skills are the one mechanism here that does not transfer cleanly across the tools, and this is the single reason the skills exercise behaves differently from desk to desk:

- **Positron Assistant does not reliably load skills.** Whether a request reaches the path that would honour a skill is undocumented and inconsistent, so the `pub-figures` skill below should not be expected to fire there. Treat any success as a bonus, not the plan.

- **Posit Assistant loads skills automatically, but from its own location.** Its skills must live under `.positai/skills/` (or another path in `skills.paths`), not Copilot's `.github/skills`. A skill authored once for Copilot is not available to Posit Assistant unless you also place it under `.positai/skills/`.

- **GitHub Copilot loads skills from `.github/skills/` on its own surface**, the cloud agent and CI, which is where the exercise below is written to run.

So a skill authored once for Copilot is not automatically available to either Positron tool. The reliable place to run the demonstration is the GitHub.com surface.

The skill demonstrated is `pub-figures`, which governs the production of every figure in the project that a human will read. It encodes a chart-selection table, a list of unconditional prohibitions (no pie charts, no dual y-axes, no dynamite plots when raw data are available), an accessible colour scheme, audience-specific styling, a mandatory colour-vision-deficiency check, and a checklist.

:::{.callout-warning}

Run this exercise where the skill will actually load. With GitHub Copilot, that is a `.github/skills/pub-figures` directory executed on GitHub.com (the cloud agent), as in Block 5, not the editor. If you want to reproduce it inside your IDE instead, that path is Posit Assistant with the skill placed under `.positai/skills/pub-figures/`; Positron Assistant will not reliably load it, and the Copilot editor will not either. The steps below are written for the GitHub.com surface.

:::

**1. Default output.** In a fresh Chat session (with no skill loaded), ask:

```
Produce a ggplot2 boxplot of mass_post_g by treatment_arm for the voles
dataset.
```

The result is a default-styled boxplot: grey background, ggplot2 default colours, no consideration of accessibility, no annotation of sample sizes. It runs and answers the literal question, and that is all.

**2. With the skill.** Open a new Chat session, select agent mode invoke the skill (`/pub-figures`), and ask the same question. The skill requires the agent to elicit the audience before producing code. It will ask whether the figure is for a journal or lab report, a poster, a talk, or public engagement. Answer "journal/lab report" and confirm the plan it produces.

The skill folder looks like this:

```
your-repo/
└── .github/
    └── skills/
        └── pub-figures/
            ├── SKILL.md
            ├── raincloud-template.R
            ├── emmeans-bridge.R
            ├── save-figure.R
            └── theme-project.R
```

The output is materially different. A raincloud (`ggdist::stat_halfeye` plus a boxplot plus jittered points) should replace the bare boxplot, because the skill specifies raincloud as the default for discrete-continuous comparisons where raw data are available. The palette is colour-blind friendly. Axis labels carry units. The pre-flight checklist is reported in the chat.

**3. Iteration.** Ask the agent to add pairwise statistical annotation. The skill prohibits hallucinating p-values, so the agent should refuse unless a fitted model object is available. Provide a fitted `lm()` and the `emmeans` contrasts; the agent will use `ggpubr::stat_pvalue_manual()` to draw brackets and labels from the model-derived data. The four-element figure legend appears at the bottom of the output.


## How to write skills

You will receive these skills rather than author them in this workshop, but it is worth seeing how a skill is built, because the same structure applies when you write your own, because it explains why the demonstration above behaved as it did, and because it explains why the `characterisation-test` skill in Block 6 is able to constrain the cloud agent so tightly.

A skill is a folder, `.github/skills/<skill-name>/SKILL.md`, with a YAML front-matter block and a Markdown body. Two parts decide whether it works:

- The `description` in the front matter is the trigger. It is the text the agent reads to decide whether the skill is relevant to a request, so it must name the condition under which the skill applies, specifically. "Used when producing any ggplot2 figure a human will read" loads when it should; "for plots" loads erratically. A skill invoked by name with `/skill-name` does not depend on the description, but a skill you want loaded automatically does.

- The body is a procedure: when to use the skill, the ordered steps, any hard prohibitions, and a final checklist. Because a skill is read only when invoked, it can be long without slowing ordinary interactions, which is why the `pub-figures` skill can carry a chart-selection table, a prohibition list, and an accessibility check at no cost to a routine question. Supporting files such as templates or reference tables can sit in the skill folder alongside `SKILL.md`.


Check out the github page where Posit has written a series of skills [here](https://github.com/posit-dev/skills/tree/main). You will see they are .md files with supporting templates and these can be used by a range of LLM providers.

## Check your understanding

**An Agent Skill differs from a workspace-level custom instructions file in that:**

`r longmcq(c(
  "Skills only work in Positron and VS Code, custom instructions work in all IDEs",
  answer = "Custom instructions are loaded automatically for every interaction on a reading surface, whereas skills are invoked by name or loaded only when judged relevant",
  "Skills are paid features and custom instructions are free",
  "Custom instructions can include code but skills cannot"
))`

**What distinguishes Plan mode from Agent mode?**

`r longmcq(c(
  "Plan mode uses a cheaper model and so produces lower-quality code",
  answer = "Plan mode is read-only and returns an ordered plan for you to approve before anything executes, whereas Agent mode carries out the work and you review the result",
  "Plan mode can only be used on GitHub.com, Agent mode only in an IDE",
  "Plan mode edits files one at a time, Agent mode edits them all at once"
))`
