# Block 8: Functionalise a script step and prove the result is unchanged {.unnumbered}

```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)
library(here)
library(webexercises)
```

You have a computation that already works and you want to tidy it into a reusable
function. An agent can do that in seconds; doing it yourself takes a little longer.
Either way the hard part is not the refactor, it is the verification: a change that
"improves" code can quietly alter its output, so how do you confirm the function
still does what the script did?

The method this block teaches is to freeze the original result before anything is
touched. You run the analysis once, save its output to a reference file (an `.rds`),
and then require the new function to reproduce that reference exactly. This is a
*characterisation test*: it does not assert the answer is correct, only that it is
*unchanged*. That distinction is the whole point, and Step 4 returns to it.

The block has a shared spine and then three paths that differ only in who writes the
function and the test:

- **Path A — by hand.** You write the function and the test yourself. Slowest, and the
  one that teaches the verification logic most directly. Start here if testthat is new.
- **Path B — delegate in two steps.** You instruct an in-IDE agent to do the refactor
  and the test, in two separate prompts, and you verify between them.
- **Path C — encode it as a skill.** Conceptual only. You see how the trust rule, once
  understood, can be bundled into a reusable instruction for an agent, and why that is
  a sharper knife than it first appears.

Paths A and B are the substance. Path C only makes sense once you have done one of
them, because a skill hides the very boundary the earlier paths make you type out.

*⏱ ~45 minutes*

## The repository layout

```
repo/
├── .github/
│   └── workflows/
│       ├── r-checks.yml              the independent check (runs on GitHub)
│       └── copilot-setup-steps.yml   the cloud agent's environment
├── data/
│   └── voles_metabolism.csv          the dataset (contains one NA, on purpose)
├── scripts/
│   ├── 01_load_data.R                loads the CSV, runs glimpse()
│   ├── 02_analysis.R                 the original script; later calls the function
│   └── functions/
│       └── summary_functions.R       the extracted function (you or the agent)
├── tests/
│   ├── make_reference.R          captures the reference
│   └── metabolic_summary.rds     you generate this in Step 0, then freeze it
│   └── test_summary_functions.R      compares the function against the reference
└── renv.lock                         pins tidyverse, here, testthat, renv
```

The single most important property of this layout: the `.rds` under `tests/`
is generated by you and by nobody else. Everything else in the trust argument follows from that, and it holds identically across all three paths.

## The original script

`R/01_load_data.R` reads the CSV into `voles` and runs `glimpse()`. Its `col_types`
specification is shown in full because the reference and the test must read the data
identically; an unspecified type here is the silent mismatch the environment section
warns about.

```{r}
#| eval: false
# scripts/01_load_data.R
library(tidyverse)
library(here)

voles <- readr::read_csv(
  here::here("data", "voles_metabolism.csv"),
  col_types = readr::cols(
    treatment_arm       = readr::col_character(),
    sex                 = readr::col_character(),
    metabolic_rate_post = readr::col_double()
  )
)

glimpse(voles)
```

`R/02_analysis.R` sources it and computes the summary you will functionalise:

```{r}
#| eval: false
# scripts/02_analysis.R — the original, trusted version
library(tidyverse)
library(here)

source(here::here("scripts", "01_load_data.R"))

metabolic_summary <- voles |>
  group_by(treatment_arm, sex) |>
  summarise(
    mean_rate = mean(metabolic_rate_post),
    sd_rate   = sd(metabolic_rate_post),
    n         = n(),
    .groups   = "drop"
  )
```

Two things to note now and leave alone. First, the dataset contains one missing value
in `metabolic_rate_post` and this computation has no `na.rm`, so one cell of the result
will be `NA`. That is deliberate and it matters in Step 4; do not "fix" it yet. Second,
`group_by()` sorts its grouping keys in the output, whereas the `.by` argument
introduced in Path A does not. The supplied `voles_metabolism.csv` is ordered so that
the first appearance of the grouping keys does not match their sorted order, which means
the unpatched `.by` function produces a genuinely different row arrangement. That fact
returns in Step 2.

## Step 0 (spine) — You capture the reference

Before anything else, run `R/02_analysis.R` yourself and confirm `metabolic_summary`
looks as you expect, including the one `NA` cell, which is expected. Then run the
capture script below once and keep the resulting `.rds`. This frozen output is the
trust anchor. Capture it now, before any refactor: afterwards `02_analysis.R` computes
the summary through the very function under test, so re-capturing would compare the
function against itself.

```{r}
#| eval: false
# tests/fixtures/make_reference.R
# Run this BEFORE refactoring, and capture only once.
source(here::here("scripts", "02_analysis.R"))
saveRDS(
  metabolic_summary,
  here::here("tests", "metabolic_summary.rds")
)
```

You now hold the reference. The three paths differ only in who produces the function
and the test that must reproduce it. Pick one path, then rejoin at the environment
section and Step 4.

## Path A — Write the function and test yourself

Write the function in `functions/summary_functions.R`. A first attempt that a tidy
refactor naturally reaches for uses the modern `.by` argument:

`r hide("Possible solution")`

```{r}
#| eval: false
# scripts/functions/summary_functions.R — first attempt, deliberately incomplete
summarise_metabolic_rate <- function(data) {
  data |>
    dplyr::summarise(
      mean_rate = mean(metabolic_rate_post),
      sd_rate   = sd(metabolic_rate_post),
      n         = dplyr::n(),
      .by = c(treatment_arm, sex)
    )
}
```

`r unhide()`

Above is a test example that will trigger a deliberate issue!

Write the test of the function above and run it:

```{r}
#| eval: false
# tests/test_summary_functions.R
library(testthat)
source(here::here("functions", "summary_functions.R"))

test_that("summarise_metabolic_rate reproduces the original summary", {
  reference <- readRDS(here::here("tests", "fixtures", "metabolic_summary.rds"))
  voles <- readr::read_csv(
    here::here("data", "voles_metabolism.csv"),
    col_types = readr::cols(
      treatment_arm       = readr::col_character(),
      sex                 = readr::col_character(),
      metabolic_rate_post = readr::col_double()
    )
  )
  expect_equal(summarise_metabolic_rate(voles), reference, tolerance = 1e-8)
})
```

`r hide("Explanation")`

Run it with `testthat::test_file(here::here("tests", "test_summary_functions.R"))`.
It fails, and not on the numbers. Because the supplied dataset is ordered so that first
appearance differs from sorted order (noted above), `.by` returns the rows in a
different arrangement from the `group_by()` that produced your reference. The values are
right; the rows are arranged differently, and `expect_equal()` reports a mismatch.
Add `dplyr::arrange(treatment_arm, sex)` to the end of the pipeline and it passes.
That failure is the lesson in miniature: a change that improves the code altered the
output, and the characterisation test caught it.

`r unhide()`

While you are writing the test by hand, resist one specific temptation. Do not type
the expected numbers into the test as literals. If you hard-code the values you saw,
the test no longer compares against your frozen reference; it compares the function
against numbers you transcribed, which proves nothing about whether the refactor
preserved the original computation. The test must *read the `.rds`*. That is the one
non-negotiable line, and writing it yourself is how you learn why.

The `col_types` specification is deliberate, and it matches the one in
`scripts/01_load_data.R` exactly. Letting `read_csv` guess types can produce a column type
that differs from the one stored in the reference, causing a mismatch that has nothing
to do with the refactor. Column types are the finest-grained version of a broader
concern, the environment the test runs in, which the environment section below
addresses. When a characterisation test fails, suspect this kind of difference before
suspecting the code.

The function you should converge on, with a roxygen2 block and an argument check:

`r hide("Possible solution")`

```{r}
#| eval: false
# R/functions/summary_functions.R — final
#' Summarise post-treatment metabolic rate by treatment arm and sex
#'
#' @param data A data frame with columns treatment_arm, sex and
#'   metabolic_rate_post.
#' @return A tibble with one row per treatment_arm and sex, giving the mean,
#'   standard deviation and n.
summarise_metabolic_rate <- function(data) {
  stopifnot(is.data.frame(data))
  required <- c("treatment_arm", "sex", "metabolic_rate_post")
  missing_cols <- setdiff(required, names(data))

    if (length(missing_cols) > 0) {
    stop(
      "`data` is missing required column(s): ",
      paste(missing_cols, collapse = ", "),
      call. = FALSE
    )
  }

  dplyr::group_by(data, treatment_arm, sex) |>
    dplyr::summarise(
      mean_rate = mean(metabolic_rate_post),
      sd_rate = sd(metabolic_rate_post),
      n = dplyr::n(),
      .groups = "drop"
    )
}
```

`r unhide()`

Finally, edit `R/02_analysis.R` to call `summarise_metabolic_rate()` instead of the
inline pipeline, so the function genuinely replaces the code it came from. Rejoin at the environment section.

## Path B — Delegate the two steps to an agent

Use this path if you want the agent to do the work while you verify. The key
discipline is to split it into two prompts, not one. A single prompt that emits both
the refactor and the test lets one misunderstanding (about row order, say) propagate
into both files at once, producing a green result that proves nothing. Separating the
prompts keeps the verifier from inheriting the refactor's assumptions, which is the
testing equivalent of keeping the `.rds` out of the agent's hands.

**Prompt 1 — the refactor only:**

```text
Move the computation in scripts/02_analysis.R into a documented function in
scripts/functions/summary_functions.R, with a roxygen2 block and an argument check.
Have 02_analysis.R call the new function instead of the inline pipeline.
Do not write any tests yet, and do not touch anything under tests/.
```

Watch the agent edit the files in your editor and read what it writes to
`R/functions/` as it appears. If it uses `.by` without a trailing `arrange()`, leave
that for now; the next step will expose it.

**Prompt 2 — the test only:**

```text
Add tests/test_summary_functions.R that loads the reference
tests/fixtures/metabolic_summary.rds, runs summarise_metabolic_rate(), and
compares with expect_equal().

Do not create, regenerate or modify anything under tests/fixtures/. I captured
that reference myself; a reference you generate proves nothing.
```

That last paragraph is the entire trust argument in three sentences. Before you trust
the green tick, apply this checklist to whatever the agent wrote, reading the lines
yourself:

1. The test *reads* `tests/metabolic_summary.rds` and compares against it. It
   does not hard-code expected numbers inside the test. If it does, the verification
   is circular whatever colour the result is.
2. The test calls `summarise_metabolic_rate()`, the actual function under test, not a
   reimplementation of the pipeline inside the test body.
3. Nothing under `tests/` was created or changed by the agent.

Now run the test with
`testthat::test_file(here::here("tests", "test_summary_functions.R"))`. If the agent's
function used `.by` without `arrange()`, it fails on row order, not values, for the
reason given in the spine. Have the agent add `dplyr::arrange(treatment_arm, sex)` and
run again. If it passed first time, force the failure yourself by removing the
`arrange()`, so you see the mechanism rather than taking it on trust. Rejoin at the
environment section.

## Path C — Encode the workflow as a skill (concept)

This path is conceptual; you will not build a runnable file here. Notice that the trust
paragraph you typed in Path B is a *durable convention*: it says the same thing every
time, regardless of the analysis. Anything you find yourself retyping into every prompt
is a candidate for a skill, a short standing instruction an agent loads and follows
without being told afresh. A skill for this workflow would say, in essence: when asked
to functionalise a script, capture the reference only if the human has not provided
one, never regenerate a fixture under `tests/fixtures/`, refactor and test in separate
steps, and compare against the human-held reference with `expect_equal()`.

The benefit is obvious: you stop retyping the boundary. The hazard is exactly the same
sentence read the other way. Once the rule lives in a skill, you no longer type it, no
longer see it, and a colleague using your skill may never know the entire trust
argument rests on a single line forbidding the agent from generating its own reference.
The boundary becomes invisible, and an invisible boundary is precisely what Step 4
warns against in the statistical domain. A skill is therefore safe to use only once you
have internalised what Paths A and B make explicit; it is a compression of that
understanding, not a substitute for it. If you build one later, keep the
"reference-stays-with-the-human" clause as a visible, commented line, so the boundary
remains legible to whoever reads the skill next.

## The environment the test runs in (spine)

A characterisation test is only as stable as the environment it runs in. The
`col_types` specification in Path A handles the finest-grained version of that, a
single column read; the rest of the environment, the R version and the entire package
set, is handled by the two GitHub Actions workflows the template ships. Read them; do
not write them.

Recall from the GitHub session that GitHub Actions runs automated workflows in response
to repository events such as a push or a pull request. Two such workflows sit in
`.github/workflows/`, both placed there for you.

`r-checks.yml` is the independent check. It runs the tests on pushes to `main` and on
every pull request, from a clean checkout on GitHub's infrastructure. It is the check
that neither you nor the agent can fake, and a green tick here is the signal you wait
for at the end.

`copilot-setup-steps.yml` is the cloud agent's environment. The cloud agent runs the
tests in the environment this file builds, and if that environment does not restore
`renv.lock`, the agent's tests will fail (or, worse, pass) for reasons that have nothing
to do with the refactor, and the resemblance to your local setup is lost. Two
requirements are load-bearing, and they fail in importantly different ways:

- The job must be named exactly `copilot-setup-steps`. GitHub looks for that job name
  specifically; rename it and the agent silently falls back to a bare environment
  without your packages.
- The job must restore the lockfile, so the agent computes against the same pinned
  package versions that `r-checks` will later use.

Note the asymmetry between the two failure modes, because it determines which one you
must watch for. A misnamed job fails loudly and safely: the fallback environment lacks
your packages, so the tests typically error on the first missing function and you notice
at once. A lockfile that restores subtly wrong versions fails quietly and dangerously:
the tests run green against different package behaviour and certify invariance against
the wrong baseline, which is exactly the failure the held `.rds` was meant to rule out.
The self-announcing error is the lesser hazard; the silent green is the one that defeats
your trust anchor.

A conventional form, using the `r-lib/actions` setup actions, looks like the following.
Check this against your actual template file rather than copying it blindly; this
reproduces the standard pattern, not necessarily your exact configuration.

:::{.callout-warning}
This setup steers towards reproducibility because `setup-renv` reads `renv.lock` and
installs the exact pinned versions, so the agent computes against the same package
versions that `r-checks.yml` will later use from a clean checkout. This is what makes
the agent's environment resemble your local IDE and is why the agent's test run and the
independent check should agree. It is also what might make this workflow more fragile if
the lockfile is out of step or does not restore packages.
:::

```yaml
name: Copilot setup steps
on: workflow_dispatch
jobs:
  copilot-setup-steps:
    runs-on: ubuntu-latest
    steps:
      - name: Check out repository
        uses: actions/checkout@v4
      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: "4.5.2"      # match the version in renv.lock
          use-public-rspm: true

      - name: Set up renv
        uses: r-lib/actions/setup-renv@v2
        env:
          RENV_CONFIG_PAK_ENABLED: "true"
      - name: Install Quarto
        uses: quarto-dev/quarto-actions/setup@v2
```

`setup-renv` reads `renv.lock` and installs exactly the pinned versions, so the agent's
tidyverse, here and testthat match yours. The final `Install Quarto` step is inherited
from the wider template that renders the book; this chapter's tests are plain R and do
not require it, so do not be surprised that nothing here uses it. The practical
consequence: when you reach Step 4, the agent's own test run and the independent
`r-checks` run should agree. If they disagree, suspect the environment before suspecting
the code.

