---
title: "A 50-Minute Tour: Parquet, SQL, and data.table"
format: html
---

```{r workshop-setup}
#| include: false
knitr::opts_chunk$set(echo      = TRUE,
                      eval      = FALSE,
                      message   = FALSE,
                      warning   = FALSE,
                      fig.align = "center")
```

This is a self-paced session for R users who know `dplyr` and the native pipe
(`|>`) but have not met Parquet, SQL, or `data.table`. It is deliberately short and
deliberately shallow: you will not write any of the three fluently by the end. You
will leave able to do one thing, and recognise three tools that do it.

The one thing is this: **read only the part of a file you actually need, rather than
loading the whole file and then throwing most of it away.** That single habit is
what keeps a large file from filling your memory and crashing the session. Parquet,
SQL, and `data.table` are three different ways to apply it. Seeing the same idea in
all three is the point; mastering any one of them is a job for a longer session.

### Learning outcomes {-}

By the end you should be able to:

1. Explain why a CSV read loads the whole file but a Parquet read need not.
2. Read selected columns from a Parquet file and measure that it reads less.
3. Query a database with `dplyr`, finish with `collect()`, and read the SQL that
   `show_query()` produces.
4. Recognise the `data.table` `DT[i, j, by]` template and map it to `dplyr`.
5. Decide, for a file of your own, which of the three tools fits the situation.

:::{.callout-note}
**Before you start.** Install the packages and confirm the data files are present.

```{r install}
#| eval: false
install.packages(c("readr", "dplyr", "arrow", "bench",
                   "DBI", "RSQLite", "dbplyr", "data.table"))
```

If `install.packages("arrow")` fails rather than completing, you can still do
Blocks 3 and 4; only Block 2 needs it. Three data files are provided, all holding
the **same** one-million-row dataset: `bird_survey.csv`, `bird_survey.parquet`, and
`bird_survey.sqlite`. Put them in your working directory and check:

```{r check-files}
#| eval: false
file.exists("bird_survey.csv")      # should be TRUE
file.exists("bird_survey.parquet")  # should be TRUE
file.exists("bird_survey.sqlite")   # should be TRUE
```

The dataset has 15 columns. The ones used below are `species`, `habitat`, `year`
(values 2000 to 2023), and `count`.
:::

## Running order

| Block | Content | Time |
|---|---|---|
| 1 | The one idea: read less of the file | ≈8 min |
| 2 | Parquet: columns you don't ask for are never read | ≈12 min |
| 3 | SQL: querying without loading | ≈12 min |
| 4 | `data.table`: the fast in-memory option | ≈10 min |
| 5 | Close: which to reach for, and a checklist | ≈8 min |

---

## 1. The one idea: read less of the file

A CSV stores data one row at a time, as plain text. To read it, R parses every
value on every line into memory. For a small file this is instant and you never
think about it. For a file approaching the size of your memory, it is the reason
the session dies: R tries to hold the whole thing at once.

There are two independent ways to read less. You can **store the file in a format
that lets you read only some columns** (that is Parquet, Block 2). Or you can
**describe the query first and let an engine run it, returning only the result**
rather than the whole table (that is SQL and `data.table`, Blocks 3 and 4). The
best tools do both.

The word that recurs is `collect()`. When you query a database or a lazy Parquet
dataset through `dplyr`, nothing is read until you call `collect()`. Before it, you
hold a *description* of the query; after it, you hold an ordinary tibble with just
the answer. Watch for that word in every block below.

:::{.callout-tip}
**I/O is the slow part.** "I/O" means reading from and writing to disk, which is far
slower than computing on data already in memory. Every technique here is, at heart,
a way to move fewer bytes off the disk. If you remember nothing else, remember that
the goal is to read fewer bytes, not to find a faster reading function.
:::

---

## 2. Parquet: columns you don't ask for are never read

Parquet stores data one *column* at a time, in compressed binary form, with a small
footer recording each column's name, type, and range. Because each column sits in
its own block, asking for two columns reads only those two blocks. The rest of the
file is never touched. This is the whole advantage, and CSV cannot do it: a CSV read
always parses every column on every line, even the ones you discard.

### Exercise 2: does selecting columns speed up the read?

**Predict first.** You will read 2 of the 15 columns from each format. Write down,
before you run it, whether you expect selecting columns to be *faster* for CSV, for
Parquet, for both, or for neither.

```{r ex-parquet}
#| eval: false
library(readr)
library(arrow)
library(bench)

bench::mark(
  csv_all      = read_csv("bird_survey.csv", show_col_types = FALSE),
  csv_two      = read_csv("bird_survey.csv",
                          col_select = c(species, count),
                          show_col_types = FALSE),
  parquet_all  = read_parquet("bird_survey.parquet"),
  parquet_two  = read_parquet("bird_survey.parquet",
                              col_select = c(species, count)),
  iterations = 5,
  check      = FALSE
)[, c("expression", "median")]
```

`r hide("Solution")`

The two CSV times should be close to each other: a CSV is row-oriented, so the
parser reads every value regardless and only *then* drops the columns you didn't
ask for. Selecting columns saves memory in R but not disk reading. The two Parquet
times should differ, with the two-column read faster: Parquet reads only the blocks
for the columns you named, so fewer columns means genuinely fewer bytes off the
disk. This is the reason to convert a large, code-queried file to Parquet.

The numbers are noisy over five iterations, so judge the *direction* of the
difference, not the exact figures. On this million-row file the Parquet gap is
real but modest; it widens as the file grows and as you skip more columns.

`r unhide()`

To convert a CSV you already have, once:

```{r convert-parquet}
#| eval: false
read_csv("bird_survey.csv", show_col_types = FALSE) |>
  write_parquet("bird_survey.parquet")
```

:::{.callout-note}
Parquet is binary, so you cannot open it in a text editor or a basic spreadsheet.
Keep CSV for files a person edits by hand or opens in Excel. Choose Parquet for a
large file you query from code.
:::

---

## 3. SQL: querying without loading

SQL (Structured Query Language) is the standard language for asking questions of
data in a database. A database file sits on disk, and you query it without loading
it into R. The good news for you: you can write the query in `dplyr`, and the
`dbplyr` package translates it to SQL automatically. The only new step is
`collect()` at the end, to pull the (small) result back into R.

We use SQLite, a database that lives in a single file and needs no server. The
provided `bird_survey.sqlite` holds one table, `surveys`, with the same data as the
CSV and Parquet files.

### Exercise 3: query a database with dplyr, then read the SQL

```{r ex-sql}
#| eval: false
library(DBI)
library(dplyr)
library(dbplyr)

# Connect to the single-file SQLite database
con <- dbConnect(RSQLite::SQLite(), "bird_survey.sqlite")

dbListTables(con)                  # should list "surveys"
surveys <- tbl(con, "surveys")     # a lazy handle, NOT a tibble yet
surveys                            # prints first rows; note the row count is "??"
```

The `??` in the printed dimensions is the lesson: R knows the columns but has not
counted the rows, because it has not read the table. Now query it, exactly as you
would a data frame, and finish with `collect()`:

```{r ex-sql-query}
#| eval: false
result <- surveys |>
  filter(year >= 2020, habitat == "woodland") |>
  group_by(year, species) |>
  summarise(total = sum(count), .groups = "drop") |>
  collect()                        # runs the query, returns a tibble
```

**Predict first:** what does the same pipeline *without* `collect()` return? Run
both and check the class of each.

```{r ex-sql-class}
#| eval: false
plan <- surveys |>
  filter(year >= 2020, habitat == "woodland") |>
  group_by(year, species) |>
  summarise(total = sum(count), .groups = "drop")

class(plan)     # a lazy query object
class(result)   # a tibble

plan |> show_query()   # see the SQL your dplyr became

dbDisconnect(con)
```

`r hide("Solution")`

Without `collect()` you hold a lazy query object, not data: `class(plan)` shows
database/`tbl` classes (such as `tbl_SQLiteConnection`), while `class(result)`
shows `tbl_df`. `show_query()` prints the generated SQL, which selects `year`,
`species`, and a `SUM(count)`, filters with a `WHERE` clause, and groups with
`GROUP BY`. Reading that SQL beside the `dplyr` you wrote is the gentlest way to
start learning it: `WHERE` is `filter()`, `SELECT` is `select()`, `GROUP BY` is
`group_by()`.

`r unhide()`

:::{.callout-warning}
One real difference between the two engines: `dplyr`'s `sum()` returns `NA` if any
value is missing, whereas SQL's `SUM` ignores missing values by default. Add
`na.rm = TRUE` on the R side if you need them to agree, and always check on a small
subset before trusting that two engines give the same answer.
:::

---

## 4. data.table: the fast in-memory option

Parquet and SQL both help when a file is large relative to your memory. `data.table`
solves a different problem: a table that *does* fit in memory, but where `dplyr`
operations feel slow. It is a drop-in alternative to the data frame with its own
compact syntax, and it is widely used for speed on large in-memory tables.

The syntax is one bracket, read as `DT[i, j, by]`: `i` chooses rows (like
`filter`), `j` chooses or computes columns (like `select` or `summarise`), and `by`
groups (like `group_by`). You do not need to write it fluently today; you need to
recognise it.

### Exercise 4: the same operation, three ways

Here is one operation, expressed in `dplyr`, in `data.table`, and in SQL, so you can
see they are the same idea. It matches the question from Block 3: woodland records,
total count per species, for years 2020 onward. Run the first two; read the third.

```{r ex-dt}
#| eval: false
library(data.table)
library(dplyr)

# Load the CSV both ways
df <- readr::read_csv("bird_survey.csv", show_col_types = FALSE)  # a tibble
dt <- fread("bird_survey.csv")                                    # a data.table

# dplyr
df |>
  filter(year >= 2020, habitat == "woodland") |>
  group_by(species) |>
  summarise(total = sum(count))

# data.table: the same, in one bracket
dt[year >= 2020 & habitat == "woodland",
   .(total = sum(count)),
   by = species]
```

The SQL form of the same question, for recognition only:

```sql
SELECT species, SUM(count) AS total
FROM surveys
WHERE year >= 2020 AND habitat = 'woodland'
GROUP BY species
```

`r hide("Solution")`

All three keep woodland rows from 2020 onward, group by species, and total the
count. In `data.table`, the `i` slot (`year >= 2020 & habitat == "woodland"`) is
the filter, the `j` slot (`.(total = sum(count))`) computes the summary, and
`by = species` groups. The three results should match. The mapping to recognise:

| | rows (`filter`) | summary (`summarise`) | group |
|---|---|---|---|
| `dplyr` | `filter(...)` | `summarise(...)` | `group_by(...)` |
| `data.table` | `DT[i, ...]` | `DT[, j, ]` | `by =` |
| SQL | `WHERE` | `SELECT ... AS` | `GROUP BY` |

`fread()` is also simply a fast CSV reader, so even if you never adopt the bracket
syntax, `fread("file.csv")` is a useful function to know.

`r unhide()`

:::{.callout-note}
`data.table` works in memory, so it does not help with a file too large to load.
For that, you are back to Parquet (Block 2) or a database (Block 3). `data.table`
is for speed on data that already fits.
:::

---

## 5. Close: which to reach for

Three tools, three situations:

- **Parquet** when you have a large file you query from code by column, and you want
  reads to touch only the columns you need. Convert once with `write_parquet()`,
  then read with `read_parquet(..., col_select = ...)` or query lazily with
  `arrow::open_dataset()`.
- **SQL / a database** when data lives in, or belongs in, a database, especially
  several related tables or a store queried repeatedly. Query it in `dplyr` through
  `dbplyr`, finish with `collect()`, and use `show_query()` to learn the SQL.
- **`data.table`** when the data fits in memory but `dplyr` is too slow, or when you
  just want a fast CSV reader in `fread()`.

The thread through all three is the one idea from Block 1: read only what you need.
Parquet reads only the columns; a database returns only the query result;
`data.table` is the fast path once the data is in memory.

### Closing checklist

- [ ] Large file, queried from code, want fast column reads? **Parquet.**
- [ ] Hand-edited or opened in a spreadsheet? **Keep CSV.**
- [ ] Data in a database, or several related tables? **Query in `dplyr` via
      `dbplyr`, end with `collect()`.**
- [ ] Fits in memory but `dplyr` feels slow? **`data.table`** (and `fread()` for a
      fast read).
- [ ] When you meet SQL, read `WHERE` as `filter()` and `SELECT` as `select()`;
      `show_query()` shows the SQL your own `dplyr` becomes.
- [ ] Don't trust a general speed claim. Measure on your file and machine with
      `bench::mark()` first.

---

## Appendix: key terms {-}

**`collect()`**: runs a lazy query (database or Arrow) and returns a tibble. Before
it, a plan; after it, data in memory.

**Column projection**: reading only the columns you asked for. Parquet does this;
CSV cannot.

**`data.table`**: a fast in-memory alternative to the data frame, with the syntax
`DT[i, j, by]`. `fread()` is its CSV reader.

**`fread()`**: `data.table`'s fast CSV reader; usable on its own.

**I/O**: input/output, the work of reading and writing disk, usually the slow part.

**Parquet**: a compressed, columnar, binary file format; small and fast to query by
column, but not human readable.

**SQL**: Structured Query Language, the standard language for querying databases;
`SELECT`, `FROM`, `WHERE`, `GROUP BY` are its core verbs.

**`show_query()`**: prints the SQL that a `dplyr` pipeline against a database
generates; a gentle way to learn SQL.
