# Working with larger data {.unnumbered}


```{r}
#| include: false
source("R/booktem_setup.R")
source("R/my_setup.R")
library(tidyverse)

library(here)
```


Throughout this block we use the Portal mammals survey data, distributed as both a
CSV file and a SQLite database. The CSV is around 35,000 rows: small enough to
inspect, large enough to make timing differences visible.

```{r}
#| eval: false
library(tidyverse)
library(data.table)   # fread() and the data.table class
library(dtplyr)       # dplyr syntax, data.table speed
library(DBI)          # database connections
library(RSQLite)      # the SQLite driver
library(microbenchmark)

csv_name <- here::here("data", "surveys.csv")
```


```{r, eval = TRUE, echo = FALSE}
downloadthis::download_link(
  link = "https://raw.githubusercontent.com/UEABIO/physalia-R-course/refs/heads/2024/surveys.csv",
  button_label = "Download surveys data as csv",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```


## Why a data frame is sometimes the wrong tool

A `data.frame` (and its tidyverse refinement, the `tibble`) holds the whole
dataset in memory and copies it on most operations. For tens of thousands of rows
this is invisible. For tens of millions it becomes the dominant cost of an
analysis, and for data larger than available RAM it fails outright.

Two strategies address this. The first keeps the data in memory but uses a
structure optimised for speed and low copying: `data.table`. The second leaves the
data on disk in a database and pulls only what each query needs: a SQLite file
queried through `dbplyr`. The rest of this block works through each in turn.

:::{.callout-note}

The thresholds are approximate and hardware-dependent, but a useful rule of thumb:
under ~1 GB, a tibble is fine; from ~1 GB up to the limits of RAM, reach for
`data.table`; beyond RAM, or when the data are already in a database, use a
database backend.

:::

## Reading large files: `fread()`

There are three common ways to read a delimited file: base R's `read.csv()`,
`readr::read_csv()`, and `data.table::fread()`. For small files the choice is
immaterial. As files grow past roughly 100 MB the differences become substantial,
with `fread()` and `read_csv()` typically several times faster than `read.csv()`.

```{r}
#| eval: false
library(microbenchmark)

microbenchmark(
  base  = read.csv(csv_name),
  readr = readr::read_csv(csv_name, show_col_types = FALSE),
  dt    = data.table::fread(csv_name, showProgress = FALSE),
  times = 10,
  unit  = "ms"
)
```

The return types differ, and this matters. `read.csv()` returns a `data.frame`,
`read_csv()` returns a `tibble`, and `fread()` returns a `data.table`. The first
two behave identically for our purposes; the `data.table` is a different object
with its own syntax, covered below.

### Reading only part of a file

When a file does not fit in memory, the simplest fix is to read less of it.
`fread()` exposes two arguments for this. `nrows` caps the number of rows read, and
`skip` discards rows from the top — or, given a string, starts reading at the first
line matching it.

```{r}
#| eval: false
# First 500 rows only
fread(csv_name, nrows = 500, showProgress = FALSE)

# Start from the first line containing "1986"
fread(csv_name, skip = "1986", showProgress = FALSE)
```

The `select` argument restricts which columns are read, which often has a larger
effect on memory than restricting rows. `drop` is its complement.

```{r}
#| eval: false
three_columns <- fread(
  csv_name,
  select       = c("species_id", "sex", "hindfoot_length"),
  showProgress = FALSE
)
```

`readr` offers the same controls under different names: `n_max` for rows, `skip`
for an offset, and `col_select` for columns. Unlike older versions, `col_select`
now accepts column names directly, so the two packages are close to equivalent
here.

```{r}
#| eval: false
readr::read_delim(
  csv_name,
  delim      = ",",
  n_max      = 500,
  col_select = c(species_id, sex, hindfoot_length)
)
```

## `data.table` syntax in one minute

A `data.table` is indexed as `dt[i, j, by]`: filter rows in `i`, do something to
columns in `j`, grouped by `by`. It is terse and fast, but the grammar is not
dplyr's.

```{r}
#| eval: false
allData <- fread(csv_name, showProgress = FALSE)

# Filter rows
allData[year > 1988]

# Select columns (note: .() returns a data.table, not a vector)
allData[, .(species_id, year)]

# Aggregate by group
allData[, .(mean_hf = mean(hindfoot_length, na.rm = TRUE)), by = species_id]
```

You do not have to learn this grammar to get the speed. `dtplyr` lets you write
ordinary dplyr and translates it to `data.table` code under the hood.

## Driving `data.table` with dplyr: `dtplyr`

`dtplyr` is the bridge. You wrap a `data.table` in `lazy_dt()`, write the dplyr you
already know, and call `as_tibble()` (or `collect()`) at the end to trigger
execution and return a normal tibble.

```{r}
#| eval: false
library(dtplyr)

allData |>
  lazy_dt() |>
  filter(year > 1988) |>
  summarise(mean_hf = mean(hindfoot_length, na.rm = TRUE),
            .by = species_id) |>
  as_tibble()
```

The `lazy_dt()` step is the important change from older tutorials. Evaluation is
*lazy*: the pipeline builds up a translation but runs nothing until
`as_tibble()`/`collect()`. To see the `data.table` code that dtplyr generates, pipe
into `show_query()`.

```{r}
#| eval: false
allData |>
  lazy_dt() |>
  filter(year > 1988) |>
  summarise(mean_hf = mean(hindfoot_length, na.rm = TRUE), .by = species_id) |>
  show_query()
```

:::{.callout-tip}

Piping a bare `data.table` straight into a dplyr verb appears to work because
recent dtplyr quietly converts it to a `lazy_dt()` for you, with a warning. Do not
rely on this. Make the `lazy_dt()` step explicit so it is clear where the
translation begins and so the laziness is visible to anyone reading the code.

:::

There is a small, fixed translation overhead per verb (under a millisecond), which
is irrelevant for large data but means dtplyr is not worth it for tiny tables. And
because dtplyr preserves dplyr's copy-on-modify semantics, it is marginally slower
than hand-written `data.table`. For most analysis work that trade is well worth the
readability.

### Activity: translate a pipeline

:::{.task}
::::{.task-header}
the same question, three ways
::::
::::{.task-container}

Using `allData`, find the mean `hindfoot_length` per `species_id` for records after
1985, sorted from largest to smallest mean. Write it (a) as native `data.table`,
(b) as a `dtplyr` pipeline, then use `show_query()` to compare your dtplyr code with
the `data.table` it generates. Are they the same?

::::
:::


`r hide("Solution")`

```{r}
#| eval: false
# (a) native data.table
allData[year > 1985,
        .(mean_hf = mean(hindfoot_length, na.rm = TRUE)),
        by = species_id][order(-mean_hf)]

# (b) dtplyr
allData |>
  lazy_dt() |>
  filter(year > 1985) |>
  summarise(mean_hf = mean(hindfoot_length, na.rm = TRUE), .by = species_id) |>
  arrange(desc(mean_hf)) |>
  as_tibble()

# inspect the translation
allData |>
  lazy_dt() |>
  filter(year > 1985) |>
  summarise(mean_hf = mean(hindfoot_length, na.rm = TRUE), .by = species_id) |>
  arrange(desc(mean_hf)) |>
  show_query()
```

The generated query is close to the hand-written version. It will not be character
-for-character identical — dtplyr introduces intermediate steps to honour dplyr
semantics — but it uses the same `[i, j, by]` machinery, which is where the speed
comes from.

`r unhide()`

## When the data live on disk: databases

`data.table` still holds everything in memory. Once the data exceed RAM, or when
they already sit in a shared database, the better move is to leave them where they
are and query across the connection. We use SQLite here because it is a single
file with no server to configure, but the same code works against PostgreSQL,
MySQL, BigQuery, and others through the common `DBI` interface.

```{r}
#| eval: false
library(DBI)
library(RSQLite)

# A connection to a SQLite database file
mammals <- DBI::dbConnect(RSQLite::SQLite(),
                          here::here("data", "portal_mammals.sqlite"))

DBI::dbListTables(mammals)
# [1] "plots"   "species" "surveys"
```

`dbConnect()` opens the connection; `RSQLite::SQLite()` names the driver. The
returned object, `mammals`, is the handle we query against. Point a `tbl()` at one
of its tables to get something that behaves like a lazy tibble.

```{r}
#| eval: false
surveys <- dplyr::tbl(mammals, "surveys")
surveys
# # Source:   table<`surveys`> [?? x 9]
# # Database: sqlite [.../portal_mammals.sqlite]
#    record_id month   day  year plot_id species_id sex   ...
#        <int> <int> <int> <int>   <int> <chr>      <chr>
#  1         1     7    16  1977       2 NL         M
#  ...
```

Note the `[?? x 9]`: R reports nine columns but an unknown number of rows. It has
not counted them, because doing so would mean reading the whole table. This
laziness is the entire point — R fetches only what a query demands.

## Querying with `dbplyr`

`dbplyr` translates dplyr verbs into SQL automatically. In practice you write the
same pipeline you would for a local tibble, with one addition: nothing is fetched
until you call `collect()`.

```{r}
#| eval: false
library(dbplyr)

# How many males were recorded in plot 2 in 1978?
surveys |>
  filter(sex == "M", year == 1978, plot_id == 2) |>
  count() |>
  collect()                # <- pulls the result into a local tibble
```

Without `collect()`, the pipeline remains a query object and prints only a preview.
With it, the SQL runs on the database and the (usually small) result comes back as
an ordinary tibble you can plot or model. Filter and aggregate on the database
side, where the data are; `collect()` only the reduced result.

:::{.callout-note}

Most dplyr verbs translate, but not all of R is available inside a database.
`slice_head()` and friends are unsupported because database tables have no
inherent row order, and any function dbplyr cannot translate will error. When that
happens, `collect()` first and finish the work locally.

:::

### Seeing the generated SQL

Learning what dbplyr produces is a low-cost way to learn SQL itself. Pipe any
un-collected query into `show_query()`.

```{r}
#| eval: false
surveys |>
  filter(year %in% c(1998, 1999, 2000)) |>
  summarise(n = n(), .by = c(year, species_id)) |>
  show_query()
# <SQL>
# SELECT `year`, `species_id`, COUNT(*) AS `n`
# FROM `surveys`
# WHERE (`year` IN (1998, 1999, 2000))
# GROUP BY `year`, `species_id`
```

If you prefer to write SQL directly, `DBI::dbGetQuery()` sends a raw query and
returns a data frame in one step.

```{r}
#| eval: false
DBI::dbGetQuery(
  mammals,
  "SELECT year, species_id, COUNT(*) AS n
   FROM surveys
   WHERE year IN (1998, 1999, 2000)
   GROUP BY year, species_id"
)
```

### Activity: SQL to dplyr and back

:::{.task}
::::{.task-header}
read it, then write it
::::
::::{.task-container}

First, translate this SQL into a dbplyr pipeline:

```sql
SELECT record_id, hindfoot_length
FROM surveys
WHERE hindfoot_length > 20
```

Then, going the other way, write the dbplyr pipeline below and use `show_query()`
to predict the SQL before you run it:

```{r}
#| eval: false
surveys |>
  filter(year > 1977, year < 2000, hindfoot_length == 30) |>
  arrange(desc(species_id))
```

::::
:::


`r hide("Solution")`

```{r}
#| eval: false
# SQL -> dbplyr
surveys |>
  select(record_id, hindfoot_length) |>
  filter(hindfoot_length > 20) |>
  collect()

# The second pipeline generates SQL along these lines:
# SELECT *
# FROM `surveys`
# WHERE (`year` > 1977 AND `year` < 2000 AND `hindfoot_length` = 30)
# ORDER BY `species_id` DESC
```

Two things to notice. The chained `filter()` conditions become a single `WHERE`
joined by `AND`, and `arrange(desc(...))` becomes `ORDER BY ... DESC`. The mapping
is close enough that reading SQL and reading dplyr reinforce each other.

`r unhide()`

## Building a database from a CSV

If your data start life as a CSV that is too large to load, you can write it into a
SQLite file in chunks, then query it lazily as above. Read a block with `fread()`,
append it to the table, and repeat. `purrr::walk()` drives the loop for its side
effect (writing), returning nothing.

```{r}
#| eval: false
library(purrr)

db <- dbConnect(SQLite(), here::here("data", "surveys.sqlite"))

# Write one chunk: read `n` rows starting after `skip`, append to the table
write_chunk <- function(skip_rows, n = 5000) {
  chunk <- fread(csv_name, skip = skip_rows, nrows = n, showProgress = FALSE)
  dbWriteTable(db, "surveys", chunk, append = TRUE)
}

# Offsets for each chunk, then write them all
skips <- seq(0, by = 5000, length.out = 8)
walk(skips, write_chunk)

dbDisconnect(db)
```

In real use you would read column names from the header once and reuse them, since
`skip` discards the header along with the data rows. The pattern, though, is the
general one: never hold more than one chunk in memory at a time.

:::{.callout-note}

**Where this leaves us.** We can read selectively with `fread()`, run dplyr at
`data.table` speed through `lazy_dt()`, and query on-disk databases through
`dbplyr` while fetching only what we need with `collect()`. The common thread is
that the dplyr grammar carries across all three settings; what changes is where
the data sit and when the work actually runs.

:::

## Reading

- dtplyr: <https://dtplyr.tidyverse.org/>
- dbplyr: <https://dbplyr.tidyverse.org/>
- data.table introduction: <https://rdatatable.gitlab.io/data.table/>
- Handling large data files in R: <https://inbo.github.io/tutorials/tutorials/r_large_data_files_handling/>
