---
title: "Misc"
editor: visual
---

# Appendix: Dataset generation

The synthetic dataset `voles_metabolism.csv` is generated by the following script. The instructor runs this once and distributes the CSV.

```{r}
#| eval: false
library(tidyverse)

set.seed(2026)

n_per_cell <- 12
treatments <- c("Control", "Low_dose", "High_dose")

voles <- expand_grid(
  treatment_arm = treatments,
  sex = c("M", "F"),
  replicate = seq_len(n_per_cell)
) |>
  mutate(
    subject_id = sprintf("V%03d", row_number()),
    mass_pre_g = round(rnorm(n(), mean = 28, sd = 3), 1),
    mass_change = case_when(
      treatment_arm == "Control"   ~ rnorm(n(), 0.2, 0.8),
      treatment_arm == "Low_dose"  ~ rnorm(n(), 1.5, 1.0),
      treatment_arm == "High_dose" ~ rnorm(n(), 3.2, 1.2)
    ),
    mass_post_g = round(mass_pre_g + mass_change, 1),
    metabolic_rate_post = round(
      8.5 + 0.15 * mass_post_g +
        case_when(
          treatment_arm == "Control"   ~ 0,
          treatment_arm == "Low_dose"  ~ 0.8,
          treatment_arm == "High_dose" ~ 1.6
        ) +
        if_else(sex == "M", 0.5, 0) +
        rnorm(n(), 0, 0.4),
      2
    )
  ) |>
  select(subject_id, sex, treatment_arm,
         mass_pre_g, mass_post_g, metabolic_rate_post)

write_csv(voles, here::here("data", "voles_metabolism.csv"))
```

The seed `2026` produces a dataset with detectable effects of treatment on both mass change and metabolic rate, and a small additive effect of sex on metabolic rate. Tests on the dataset will recover these effects.

# Appendix: Workshop template repository

The fork-target repository should contain, at minimum:

```         
workshop-template/
├── .github/
│   ├── workflows/
│   │   ├── copilot-setup-steps.yml
│   │   └── r-checks.yml
│   └── skills/
│       └── characterisation-test/
│           └── SKILL.md
├── data/
│   └── voles_metabolism.csv
├── R/
│   ├── functions/
│   │   └── (empty; the agent populates this in Part 6)
│   ├── 01_load_data.R
│   └── 02_analysis.R
├── tests/
│   ├── fixtures/
│   │   └── make_reference.R
│   └── (test_summary_functions.R is added by the agent in Part 6)
├── outputs/
│   ├── figures/
│   └── tables/
├── renv/
│   └── activate.R
├── renv.lock
├── .Rprofile
├── .gitignore
└── README.md
```

Participants create `.github/copilot-instructions.md` themselves in Part 5; everything else above ships in the template.

The two analysis scripts are short. `R/01_load_data.R` contains only the load step and a `glimpse()` call, and `R/02_analysis.R` sources it and computes `metabolic_summary` (the `02_analysis.R` body is shown in Part 6, "What the template provides"):

```{r}
#| eval: false
library(tidyverse)
library(here)

voles <- read_csv(here::here("data", "voles_metabolism.csv"))

glimpse(voles)
```

The reference fixture `tests/fixtures/metabolic_summary.rds` is deliberately **not** shipped. Participants generate it by running `tests/fixtures/make_reference.R` in Part 6, Step 0, which is the point at which they take ownership of the trust anchor. The capture script itself is listed in Part 6.

`.github/skills/characterisation-test/SKILL.md` ships with the content shown in Part 6, "What the template provides". It is the only skill bundled in this template; the general-purpose course skills are distributed separately.

The two GitHub Actions workflows live in `.github/workflows/`. Their contents are project-specific: they set up R, restore `renv`, and run `testthat`. Paste your prepared versions into the placeholders below. The job in `copilot-setup-steps.yml` must be named exactly `copilot-setup-steps`, or the cloud agent will not pick it up.

``` yaml
# .github/workflows/copilot-setup-steps.yml
# <paste your copilot-setup-steps.yml here>
# Reminder: the job MUST be named `copilot-setup-steps`.
```

``` yaml
# .github/workflows/r-checks.yml
# <paste your r-checks.yml here>
# Runs on push to main and on every pull request; restores renv and runs testthat.
```

The template uses `renv`. Commit `renv.lock`, `.Rprofile`, and `renv/activate.R`, but not `renv/library/`. Participants run `renv::restore()` once after forking (see Materials), so their local environment matches the lockfile that both workflows restore.

The `.gitignore` should exclude `.Rhistory`, `.RData`, `.Rproj.user/`, `renv/library/`, and the standard temporary files.

The `README.md` should be brief: what the workshop is, how to fork, that the first step after forking is `renv::restore()`, and who to contact if access fails. Participants do not need a long README to do the exercises.
