# (PART\*) Reproducible Reports with Quarto {.unnumbered}

# Quarto: a 90-minute workshop {#sec-quarto-workshop}

```{r, child = '_setup.Rmd', warning = FALSE, message = FALSE}

```

```{r workshop-setup, include = FALSE}
# Code in this chapter is shown for reading, not executed, so the chapter
# renders with no data dependencies. The instructor runs the live
# demonstrations during the session.
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")
```

## 1. Background

Quarto is a widely-used tool for creating automated, reproducible, and share-worthy outputs, such as reports. It can generate static or interactive outputs, in Word, pdf, html, Powerpoint slides, and many other formats.

A Quarto file combines R code (or other programming code) and text such that the script actually becomes your output document. You can create an entire formatted document, including narrative text (can be dynamic to change based on your data), tables, figures, bullets/numbers, bibliographies, etc.

Documents produced with Quarto, allow analyses to be included easily - and make the link between raw data, analysis & and a published report *completely reproducible*.

With Quarto we can make reproducible html, word, pdf, powerpoints or websites and dashboards^[(https://quarto.org/docs/gallery/)]

This chapter assumes no prior Quarto experience. If you have used R Markdown the transition is straightforward, because the components are the same: a YAML header, Markdown narrative, and code chunks.

```{block, type = "info"}
**How to use this document.**

The numbered sections below are written to be *delivered live*. Each of the three exercises also contains a complete, copyable document and a worked solution in a collapsible box, so you can return afterwards, paste the exercise into a new `.qmd`, and repeat it on your own.

```


## 2. Quarto orientation for newcomers

Quarto is the successor to R Markdown. Its advantage is a single, consistent syntax across every output, so the same document conventions produce an HTML page, a Word file, a PDF, or a slide deck. It also works in several editors, including RStudio, VS Code, and JupyterLab, and supports R, Python, Julia, and Observable JS.

Create one with **File > New File > Quarto Document**. Choose a title and an output format, click OK, and you get a document with placeholder content. The **Render** button at the top converts it into the chosen format.

```{block, type = "tip"}
RStudio shows a Quarto document in **Visual** mode (partially rendered) or **Source** mode (plain text). The explanations below assume Source mode.
```

Every `.qmd` has three parts: a YAML header, Markdown text, and code chunks.

### YAML metadata

The YAML sits at the top between two lines of three dashes and holds the document's metadata as `key: value` pairs.

````md
---
title: "Penguins report"
author: "Your name"
format: html
---
````

```{block, type = "info"}
Quarto uses `format:`, not the R Markdown key `output:`. Change `html` to `docx`, `pptx`, or `pdf` to produce other formats. See the list of formats at <https://quarto.org/docs/output-formats/all-formats.html>.
```


```{block, type = "try"}
Edit the YAML to set yourself as author, then change `format:` to `docx` and render. Change it back to `html` afterwards.
```

### Code chunks

Everything after the YAML is Markdown text unless it sits inside a code chunk. A chunk opens with three back-ticks and `{r}`, and closes with three back-ticks.

````md
```{r}`r ''`
library(tidyverse)
library(palmerpenguins)
```
````

Quarto runs the chunk's R code on render and places the result in the document. Chunks can carry options, written one per line with the **hash-pipe** `#|` and YAML `key: value` syntax.

````md
```{r}`r ''`
#| echo: false
hist(penguins$bill_length_mm)
```
````

```{block, type = "info"}
Note the colon and the space after `#|`. The R Markdown forms `echo = FALSE` inside the chunk header, and `#| echo = TRUE`, are not valid Quarto chunk options.

The most useful options are `echo` (show the code), `eval` (run the code), `message` and `warning` (show messages and warnings), `include` (include any output), and the figure options `fig-width`, `fig-asp`, `out-width`, and `fig-align`.
```

Document-wide defaults go in an `execute:` block in the YAML, and any chunk can override them locally.

````md
---
title: "Penguins report"
format: html
execute:
  echo: false
  warning: false
  message: false
---
````

**Question.** To show the R code but **not** its output, which combination do you need? `r mcq(sample(c("echo: false, eval: false", "echo: false, eval: true", answer = "echo: true, eval: false", "echo: true, eval: true")))`

### Markdown text and inline code {-}

The narrative is written in Markdown: `*italic*`, `**bold**`, `` `code` ``, `#` for a level-one heading, `##` for level two, and `*` for bullets. The single most important habit at the prose level is that **every number in the text should be inline code, not a typed constant**, so that it updates on render.

````md
The data were collected on `r "\u0060r Sys.Date()\u0060"`.
The mean bill length is `r "\u0060r round(mean(penguins$bill_length_mm, na.rm = TRUE), 1)\u0060"` mm.
````

```{block, type = "info"}
**Instructor, live demo (~4 min).** Create a new Quarto document, paste a three-line YAML, a `library(tidyverse)` chunk, and a `ggplot` of the penguins, then render to HTML. Toggle Visual and Source. Add one inline number to the prose and re-render to show it update. Screenshots are no substitute here; do it live.
```

## 3. The durable kernel: what does not delegate (≈10 min)

On render, Quarto re-executes every non-frozen chunk, in order, in a fresh process. A clean render is therefore evidence that the analysis runs end to end today. It is **not**, on its own, evidence of reproducibility, because render does not:

* **pin package versions**, which is `renv::snapshot()` and `renv.lock`;

* **fix random results**, which is `set.seed()`;

* **fix file locations**, which is `here::here()` and an `.Rproj`.

Two disciplines outrank any single feature:

1. **Render from a clean session.** A fresh process holds no hidden objects, so it exposes missing `library()` calls, undefined variables, and wrong ordering that an interactive session quietly hides.



```{block, type = "warning"}
"It rendered on my machine" is the start of a reproducibility claim, not the claim itself. Keep heavy computation in functions under `R/` or in scripts you `source()`, and let the `.qmd` orchestrate: load, call, display. Thin chunks render quickly and stay testable.
```

## 4. The mechanics the exercises will exploit (≈10 min)

Learn only the options the exercises will later turn into defects. Four options can make a document render cleanly while showing output that was never computed from the current data, so learn to read them.

````md
```{r}`r ''`
#| eval: false     # displayed but NOT run
#| include: false  # run, but output hidden
#| cache: true     # reuse stored results; can serve STALE output
#| freeze: true    # skip re-execution; same staleness risk
```
````

Two more mechanics complete the set. First, **every chunk label must be unique**; a duplicate stops the render. Second, a figure **cross-reference** needs a `fig-` label and a caption, and is cited with `@`.

````md
```{r}`r ''`
#| label: fig-massflipper
#| fig-cap: "Body mass against flipper length."
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```

As @fig-massflipper shows, mass rises with flipper length.
````

```{block, type = "info"}
The label and the citation must match exactly. A figure labelled `fig-massflipper` cited as `@fig-mass` renders as a dangling `?@fig-mass`.
```

## 5. Exercise 1: loud failures, what to delegate (≈12 min) {-}

```{task}
Open the document below in a new `.qmd` and render it. It will not render: it has three defects that each raise an error. Fix them so it renders.
```

````md
---
title: "Exercise 1: loud failures"
format: html
---

```{r}`r ''`
#| label: clean
library(tidyverse)
library(palmerpenguins)
penguins <- penguins |> drop_na(body_mass_g, flipper_length_mm)
```

```{r}`r ''`
#| label: clean
summary(model)
```

```{r}`r ''`
model <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
```

```{r}`r ''`
extra <- read_csv("/Users/phil/Documents/penguins.csv")
```
````

`r hide("Solution")`

* **Duplicate chunk label.** Two chunks are labelled `clean`; this is a parse error that blocks the whole render. Rename one of them, for example `model-summary`.
* **Result before its cause.** `summary(model)` is called before `model` is fitted. A fresh render has no prior session state to rescue it. Move the fitting chunk above the summary.
* **User-specific absolute path.** `"/Users/phil/Documents/penguins.csv"` fails on every other machine. Here the cleanest fix is to drop the line and use `palmerpenguins::penguins` directly; in a real project, keep the data inside the project and read it with `here::here("data", "penguins.csv")`.

`r unhide()`

```{block, type = "info"}
**The lesson is the framing, not the bugs.** These are mechanical, fast, and an assistant catches them reliably. This is the category you delegate: ask the assistant to render from a clean session and fix what breaks. Spending your own attention here is the wrong allocation.
```

## 6. Exercise 2: silent failures, calibration (≈14 min) {-}

```{task}
Render the document below. It renders **without error**, and it is **wrong**. Find the three defects. Then do one more thing: ask your AI assistant to check the document, and judge whether it caught everything.
```

````md
---
title: "Exercise 2: silent failures"
format: html
---

```{r}`r ''`
#| label: setup
#| include: false
library(tidyverse)
library(palmerpenguins)
```

```{r}`r ''`
#| label: clean
#| eval: false
penguins <- penguins |> drop_na(sex)
```

```{r}`r ''`
#| label: mean-bill
mean_bill <- mean(penguins$bill_length_mm, na.rm = TRUE)
```

# Results

The mean bill length is 39.0 mm.

There are this many birds of each sex:

```{r}`r ''`
#| label: sex-counts
#| echo: false
penguins |> count(sex)
```

```{r}`r ''`
#| label: fig-massflipper
#| fig-cap: "Body mass against flipper length."
#| echo: false
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```

Body mass rises with flipper length (@fig-mass).
````

`r hide("Solution")`

* **A typed number.** The mean is computed into `mean_bill` and then ignored; the prose hard-codes "39.0 mm" when the true value is near 43.9 mm. Replace the typed number with inline code: `r "\u0060r round(mean_bill, 1)\u0060"` mm.
* **A chunk that does not run.** The cleaning chunk carries `eval: false`, so the unknown-sex birds are never removed and the sex counts include an `NA` category the narrative implies was dropped. The output looks normal. Remove `eval: false`.
* **A broken cross-reference.** The figure is labelled `fig-massflipper` but the prose cites `@fig-mass`, which renders as a dangling `?@fig-mass`. Make the citation match the label.

`r unhide()`

```{block, type = "info"}
**The lesson is calibration.** These are still mechanical, and a capable assistant usually catches them when asked the right question. So why see them yourself? Because you cannot supervise a checker you do not understand. You need enough competence to know when the assistant's check is trustworthy and when it is not, and to notice the defect it happened to miss. Delegation without calibration is over-reliance with extra steps.
```

## 7. Exercise 3: the plausible error, what only you can do (≈12 min) {-}

```{task}
Render the document below. It renders cleanly, every reported number is produced faithfully by inline code, every chunk runs, and the cross-reference resolves. It passes every check from Exercises 1 and 2. It is still wrong. Find out how.
```

````md
---
title: "Exercise 3: a plausible error"
format: html
---

```{r}`r ''`
#| label: setup
#| include: false
library(tidyverse)
library(palmerpenguins)

# An entry error has been injected into the data for this exercise.
penguins_data <- penguins |>
  drop_na(body_mass_g)
penguins_data$body_mass_g[1] <- 47000
```

# Results

```{r}`r ''`
#| label: heaviest
heaviest <- max(penguins_data$body_mass_g)
```

The heaviest individual in the sample weighed `r "\u0060r heaviest\u0060"` g.
````

`r hide("Solution")`

The data contains an injected entry error: one bird is recorded at 47,000 g, about 47 kg. The report duly states it, because the number in the prose genuinely matches the number in the data. Every automatable consistency check passes for exactly that reason. The only thing that flags it is knowing a plausible body mass: no penguin approaches 47 kg, and the heaviest bird in this dataset should be near 6.3 kg. That is domain knowledge, and it is yours to supply.

Note what this does to the assistant. A generic "verify this document is consistent" request returns nothing, because it *is* consistent. The assistant can catch the error, but only if you direct it to test biological plausibility against the right range, and that range is yours.

**The durable fix is to make the plausibility check run.** Encode your domain constraints as `assertr` checks at the point of import, so an impossible value stops the pipeline loudly instead of waiting for a reader to notice.

````md
```{r}`r ''`
#| label: import-checked
penguins_data <- penguins |>
  drop_na(body_mass_g, flipper_length_mm) |>
  assertr::verify(body_mass_g > 2000 & body_mass_g < 7000) |>
  assertr::verify(flipper_length_mm > 150 & flipper_length_mm < 250)
```
````

`r unhide()`

```{block, type = "warning"}
This is the connection to your wider workflow. The `assertr` block you develop during exploratory analysis is precisely the mechanism that converts a silent omission failure into a loud, render-stopping one. It is the only one of today's defects the assistant cannot resolve for you, because only you know which constraints matter.
```

## 8. Styling your report (≈8 min) {-}

Quarto's HTML appearance is controlled by Sass variables, which you set in a `.scss` file referenced from the YAML. A style file has two parts: `scss:defaults` for variables, and `scss:rules` for ordinary CSS.

```{block, type = "info"}
**Instructor.** For a room new to Quarto, run this as a demonstration with the files open rather than full hands-on; eight minutes is not enough for newcomers to edit Sass and re-render reliably. Offer it as take-home.
```

Reference the style file from the YAML, keeping the base theme.

````md
---
title: "Styled report"
format:
  html:
    theme: [default, styles.scss]
---
````

A minimal `styles.scss` that sets the body colour, adds a coloured left border to code blocks, and underlines the title:

````scss
/*-- scss:defaults --*/
$body-color: #000080;
$code-block-border-left: #FE019A;

/*-- scss:rules --*/
.title {
  text-decoration: underline;
  text-decoration-color: $code-block-border-left;
}
````

```{block, type = "try"}
**Add a Google Font.** Put the import at the top of `scss:defaults` (a CSS `@import` must precede other statements), then set it as the body font.

The font is fetched from Google's servers when the page is viewed, so this needs a network connection and sends the reader's IP to Google, which is a privacy consideration in an institutional setting. For an offline or self-contained document, download the `.woff2` files and declare them with a local `@font-face` instead.
```

````scss
/*-- scss:defaults --*/
@import url('https://fonts.googleapis.com/css2?family=Inter:wght@400;600;700&display=swap');
$font-family-sans-serif: "Inter", sans-serif !default;

$body-color: #000080;
$code-block-border-left: #FE019A;
````

## 9. The division of labour, and close (≈9 min) {-}

The right-hand column is the work that free AI access leaves to you. It is also the work that, under automation complacency, is most easily skipped. Where you can, move a right-hand item leftward by making it an explicit check (`assertr`, `testthat`, CI) rather than a thing someone must remember to do.

| Hand to the assistant | Keep for yourself |
|---|---|
| Render from a clean session and fix what breaks | Whether a result is biologically possible |
| Audit chunk options (`eval`, `include`, `cache`, `freeze`) | Whether the model specification is defensible |
| Check prose numbers are inline, not typed | Whether the claim's scope matches the data analysed |
| Confirm cross-references resolve | Which plausibility constraints to encode as `assertr` |
| Linting, style, boilerplate documentation | Interpretation and the scientific conclusion |

### Closing checklist {-}

- [ ] Renders cleanly **from a fresh session**, ideally in CI.
- [ ] Every reported number is inline code, not typed.
- [ ] No `eval: false`, `cache`, or `freeze` is hiding a step or serving stale output.
- [ ] Cross-references resolve.
- [ ] Paths are relative, data is local and archived, and the environment is recorded (`renv`, Quarto version).
- [ ] **Biological plausibility is encoded as `assertr` checks**, not left to the reader.
- [ ] You have asked the assistant to verify, *and* you know which of its checks you trust.

## Appendix A: Quarto refresher {-}

Three parts: YAML header, Markdown narrative, and code chunks. Render re-executes all non-frozen chunks. Useful chunk options beyond the four "silent" ones:

````md
```{r}`r ''`
#| label: unique-name   # duplicates stop the render
#| echo: true           # show or hide the code
#| message: false       # suppress messages
#| error: false         # false means a chunk error STOPS the render (good)
#| fig-cap: "..."       # required for a figure cross-reference
```
````

Inline code is written `r "\u0060r expr\u0060"` and updates on render.

For the full set of evaluation and display options see Appendix C, for output formats Appendix D, and for figure and `ggplot` sizing Appendix E.

## Appendix B: TADA mapped onto Quarto {-}

TADA [@ivimeycook2025] is a minimum standard for sharing functional analytical code.

* **Transferable.** Share the `.qmd`, not a PDF or Word file; use `here::here()`, never `setwd()`; commit any `_extensions/` directory and record the Quarto version, which sits outside `renv`.
* **Available.** Archive in an immutable repository with a DOI (Zenodo, Figshare); GitHub is for development, not archiving.
* **Documented.** A README with author, manuscript, run order, and licence; capture the environment with `sessioninfo::session_info()` and `renv::snapshot()`.
* **Annotated.** Label every chunk uniquely, signpost results, and comment the rationale rather than the action.

## Appendix C: Code chunk options and evaluation {-}

A chunk does two separable things: it *runs* code, and it *shows* things. Four options control these independently, and confusing them is the most common source of a document that looks right but is not. Think of them as four verbs:

* `eval` runs the code (default `true`).
* `echo` shows the source code (default `true`).
* `output` shows the results of running the code (default `true`).
* `include` is the master switch: `include: false` runs the code but shows nothing at all, neither code nor output.

Three more govern side output: `message` and `warning` show messages and warnings (default `true`), and `error` controls what happens when a chunk fails. `error: false`, the default, means a failing chunk **stops the render**, which is what you want; `error: true` lets the error appear in the document and carries on.

The practical question is usually "what do I want the reader to see?" This table maps intent to options.

| What you want | Options |
| --- | --- |
| Code and its output (the default) | none |
| Output only, hide the code | `#| echo: false` |
| Code only, do not run it | `#| eval: false` |
| Run it, but show nothing | `#| include: false` |
| Code, but hide its output | `#| output: false` |
| A clean report with no code, messages, or warnings | `echo: false`, `message: false`, `warning: false` |
| Let an error display instead of stopping the render | `#| error: true` |

```{block, type = "warning"}
`eval: false` is the dangerous one in a report. The chunk is shown but never runs, so a reader sees code that did not produce the numbers in the document. `cache: true` and `freeze: true` carry the related risk of serving stored output that no longer matches the current data. Treat all three as things to audit, not set and forget.
```

Set options for one chunk with the hash-pipe, one per line, using `key: value` (a colon, with a space after `#|`):

````md
```{r}`r ''`
#| label: model-summary
#| echo: false
#| warning: false
summary(model)
```
````

Set them for the whole document in the `execute:` block of the YAML; any chunk can then override a default locally.

````md
---
title: "My report"
format: html
execute:
  echo: false
  warning: false
  message: false
---
````

```{block, type = "info"}
Quarto's `#|` options use dashes and colons (`#| fig-width: 6`). The R Markdown forms, dotted names in the chunk header (`{r, fig.width = 6}`) and the equals sign (`#| echo = TRUE`), are not valid Quarto. This chapter is itself an `.Rmd`, so its own live chunks use the older knitr names; your `.qmd` files should use the dashed forms shown in the examples.
```

## Appendix D: Output formats {-}

The central promise of Quarto is one source document, many outputs, selected with `format:`. The same Markdown, chunks, and YAML produce a web page, a PDF, a Word file, or slides.

| Output | `format:` value | What it needs |
| --- | --- | --- |
| Web page | `html` | Nothing extra. Add `embed-resources: true` for a single self-contained file with no sidecar folder. |
| PDF | `pdf` | A LaTeX installation; the simplest is `tinytex::install_tinytex()`. Alternatively `format: typst` uses the bundled Typst engine and needs no LaTeX. |
| Word | `docx` | Optional: a reference `.docx` to carry house styles. |
| PowerPoint | `pptx` | Optional: a reference `.pptx` template. |
| HTML slides | `revealjs` | Nothing extra. |
| PDF slides | `beamer` | LaTeX. |

```{block, type = "warning"}
The PDF toolchain is the usual stumbling block in a workshop. If you do not need print-quality PDF, teach and demonstrate in `html`, which has no external dependency. Reserve PDF for a follow-up once `tinytex` is installed, or use `format: typst` to sidestep LaTeX entirely.
```

A single document can declare several formats at once; Quarto produces all of them on render, and you can nest options that apply only to one.

````md
---
title: "My report"
format:
  html:
    toc: true
  pdf:
    fig-width: 7
---
````

Quarto also produces multi-document projects, configured in a `_quarto.yml` file rather than a single header: full **websites**, multi-chapter **books** (this course book is one), and interactive **dashboards**. These are projects rather than formats; see <https://quarto.org/docs/output-formats/all-formats.html> and the project documentation for the configuration.

## Appendix E: Figure and `ggplot` sizes {-}

Sizing is the most common source of frustration because there are two different sizes, and they do different things.

* `fig-width` and `fig-height` (in inches) set the size of the device **R draws the figure on**. The default width is about 7 inches. This controls the *ratio* of text and point size to the plotting area: a larger device makes `ggplot` text and points look proportionally **smaller**, and a smaller device makes them look larger. This is why a plot can have text that is too big or too small even though it occupies the right amount of page.
* `out-width` and `out-height` set the size at which the **already-drawn figure is displayed** in the document, usually as a percentage such as `"70%"`. Scaling here changes how much page the image occupies without changing the text-to-plot ratio.
* `fig-asp` sets the height-to-width ratio. It is easier to reason about than a separate height: set `fig-width` and `fig-asp` and let the height follow. The golden ratio, `0.618`, is a reasonable default.

The recommended workflow, which keeps text size consistent across a document, is to set a width and aspect ratio as document defaults under the `format:` block, then adjust only `fig-asp` per chunk, and use `out-width` to fit the page.

````md
---
title: "My report"
format:
  html:
    fig-width: 6
    fig-asp: 0.618
---
````

To override for one figure, set the options on that chunk. The example below draws on a 7-inch device, displays at 70% of the text width, and is centred.

````md
```{r}`r ''`
#| label: fig-massflipper
#| fig-cap: "Body mass against flipper length."
#| fig-width: 7
#| fig-asp: 0.8
#| out-width: "70%"
#| fig-align: center
ggplot(penguins, aes(flipper_length_mm, body_mass_g, colour = species)) +
  geom_point() +
  theme_minimal(base_size = 11)
```
````

There is a second, independent lever for text. The `base_size` argument of a `ggplot` theme sets the font size inside the plot directly, regardless of the device size.

````md
```{r}`r ''`
#| fig-width: 7
ggplot(penguins, aes(flipper_length_mm, body_mass_g)) +
  geom_point() +
  theme_minimal(base_size = 14)   # larger text without changing the device
```
````

```{block, type = "try"}
Render one figure three times, changing only `fig-width` to 3, 7, and 10 with `fig-asp` fixed. Watch the points and axis text grow relatively smaller as the device widens, while the plot keeps its shape. Then change only `out-width` between `"40%"` and `"100%"` and note that the text-to-plot ratio does not change. Seeing both effects once makes the distinction stick.
```

```{block, type = "info"}
Use `fig-width` and `fig-asp` to compose the plot and control how large its elements look; use `out-width` to fit it to the page; use `theme(base_size = )` as a separate handle on text. In short: pick two of width, height, and aspect ratio, set the display size separately, and tune `base_size` for legibility.
```

## References {-}

The penguin examples use the `palmerpenguins` package [@R-palmerpenguins]; the data-entry error in Exercise 3 is fabricated for teaching.

* Trisovic, A., Lau, M. K., Pasquier, T., & Crosas, M. (2022). A large-scale study on research code quality and execution. *Scientific Data*, 9, 60.
* Kellner, K. F., Doser, J. W., & Belant, J. L. (2025). Functional R code is rare in species distribution and abundance papers. *Ecology*, 106, e4475.
* Ivimey-Cook, E. R., et al. (2025). *TADA! Simple guidelines to improve analytical code sharing for transparency and reproducibility.* (Manuscript.)
* Müller, K., & Bryan, J. (2020). *here: A Simpler Way to Find Your Files.*

```{block, type = "info"}
**Instructor.** Replace the framing claims in block 1 with precise, pinned references from the automation-bias literature before distribution.
```
